---
title: "01 Izar 2020"
author: "Jesslyn Goh"
date: "4/17/2020"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = here::here())
knitr::opts_chunk$set(echo = TRUE)
```

```{r init}
# Load packages
library(tidyverse) # ggplot, dplyr, read_r (read_xlsx)
library(Matrix) # interact with sparse data
library(Seurat)
library(patchwork) # package to combine plots
library(glue)
library(beepr)
library(readxl)
library(ggplot2)

# saving obj setup
proj = "Izar_2020"
date = Sys.Date()
fig.dir = "jesslyn_plots/cohort1"

# Read in Seurat objects
cohort1 = readRDS("data/Izar_2020/Izar_2020_10x.RDS")

# Read in gene lists
hkgenes = read_lines("gene_lists/tirosh_house_keeping.txt", skip = 2)
ccgenes = read_lines("gene_lists/regev_lab_cell_cycle_genes.txt")
s.genes <- ccgenes[1:43]
g2m.genes <- ccgenes[44:97]

#filter out clusters -1, 19-21
cohort1 <- subset(cohort1, subset = clst > -1 & clst < 19)
```

# Jesslyn Tasks for 4/24/20
*Run all tasks on datasets separately*

[Dplyr cheatsheet for data transformation](https://github.com/rstudio/cheatsheets/raw/master/data-transformation.pdf)

## Analysis & plots (for 4/29/20)
  - Read in expression data **done**
  - Read in gene lists **done**
  - Read in and wrangle supp tables
    - *hint1:* might have to make edited version of excel files 
    - *hint2:* use `read_xlsx()`
    - Supp table 1 for patient/PDX metadata (treatment info, etc)
    - Supp table 2 for cell type assignments (can enter this in manaully if you want) **done**
    - Add relevant metadata to metadata slots of Seurat objects 
    - Add cell type assignments **done**
  - Make distribution plots of gene expression by cells and by genes
  - Add cell cycle state to metadata slot **done**
  - Patient-specific metrics (*hint:* use `summarize()` and `count`)
    - Number of patients **done**
    - Number of cells per patient **done**
    - Number of cels per type per patient **done**
  - Run PCA, plot PCA using Seurat functions (dotplot + Heatmaps) **done**
  - Run UMAP **done**
  - Run clustering **done**
  - Run Find Marker genes **done**
  - Generate UMAP plots for all metadata clumns
  
hint: to save plot use `ggsave()`

## (for 4/29/2020)
COHORT1
- remove clusters from cohort1 (19-21, maybe -1)
- run clustering and umap separately for malignant and nonmalignant cells
   - create patient and treatment status specific overlays
- score cells using gene list (list describe certain biological functions)
- maybe run infercnv
- make heatmap (y = genes, x = cells, cells grouped by cell type) - top20
- average gene expression per gene per cell type (make a table) 
  - one for columns that are cell types 
  - one for columns that are clusters
- make table: number of cells per treatment status
- make table: number of cells per treatment status per cell type
- make table: subset malignant cells --> number of cells per patient per treatment status
- save tables to csv files
(important to have a big enough sample size)

#PROCESSING ALL CELLS 
```{r treatmentInfo}
supptab1 = read_excel("data/Izar_2020/Table_S1_FINAL_Rev3.xlsx")

#treatment info of cohort 1
pnt = supptab1 %>% mutate(Patient = parse_number(`Patient ID`)) %>%
                    select(`Patient`, `Treatment status of sample`) %>% filter(`Patient`<=6)

pnt = rbind(pnt[1:2,], c(2.1, "On-treatment"), pnt[-(1:2),])
pnt = rbind(pnt[1:6,], c(5.1, "After 1 cycle of chemotherapy"), pnt[-(1:6),])

pnt[6,2] = "Treatment-naïve"

pnt = mutate(pnt, sample.ID = c("1976", "3250", "3250.1", "3266", "3281", "3288", "3288.1", "3290"))

#add treatment info to metadata 
treatment <- rep("treatment", ncol(cohort1))
treatment[cohort1$sample.ID == "1976"] <- "On-treatment"
treatment[cohort1$sample.ID == "3250"] <- "On-treatment"
treatment[cohort1$sample.ID == "3250.1"] <- "On-treatment"
treatment[cohort1$sample.ID == "3266"] <- "Treatment-naive"
treatment[cohort1$sample.ID == "3281"] <- "After 1 cycle of chemotherapy"
treatment[cohort1$sample.ID == "3288"] <- "Treatment-naive"
treatment[cohort1$sample.ID == "3288.1"] <- "After 1 cycle of chemotherapy"
treatment[cohort1$sample.ID == "3290"] <- "Treatment-naive"

cohort1 <- AddMetaData(cohort1, treatment, col.name = "treatment.Status")
```

```{r sample Assignment}
sample <- rep("0", ncol(cohort1))
sample[cohort1$sample.ID == "1976"] <- "1"
sample[cohort1$sample.ID == "3250"] <- "2.0"
sample[cohort1$sample.ID == "3250.1"] <- "2.1"
sample[cohort1$sample.ID == "3266"] <- "3"
sample[cohort1$sample.ID == "3281"] <- "4"
sample[cohort1$sample.ID == "3288"] <- "5.0"
sample[cohort1$sample.ID == "3288.1"] <- "5.1"
sample[cohort1$sample.ID == "3290"] <- "6"

cohort1<- AddMetaData(cohort1, sample, col.name = "sample")
```

```{r cellTypeAssignment}
supptab2 = read_excel("data/Izar_2020/Table_S2.xlsx", skip=2) #make 2nd the header


celltypes = rep("NA", ncol(cohort1))
celltypes[cohort1$clst %in% c(1,2,3,4,5)] <- "Malignant" #gives a logical vector for every spot in the cluster vector

celltypes[cohort1$clst == "6"] <- "Fibroblast"
celltypes[cohort1$clst == "7"] <- "Fibroblast"
celltypes[cohort1$clst == "8"] <- "Fibroblast"
celltypes[cohort1$clst == "9"] <- "Fibroblast"
celltypes[cohort1$clst == "10"] <- "Macrophage"
celltypes[cohort1$clst == "11"] <- "Macrophage"
celltypes[cohort1$clst == "12"] <- "Macrophage"
celltypes[cohort1$clst == "13"] <- "Macrophage"
celltypes[cohort1$clst == "14"] <- "DC"
celltypes[cohort1$clst == "15"] <- "DC"
celltypes[cohort1$clst == "16"] <- "B cells"
celltypes[cohort1$clst == "17"] <- "T cells"
celltypes[cohort1$clst == "18"] <- "Erythrocytes"

cohort1 <- AddMetaData(cohort1, celltypes, col.name = "cell.type")
Idents(cohort1) <- "cell.type"

```

```{r patientMetrics}

#patient metrics for cohort1
numPatients = length(unique(cohort1@meta.data$patient)) #number of patients in the cohort 

numCellsPerPatient = table(cohort1@meta.data$patient) %>% as.data.frame() #number of cells per patient 
#write_csv(numCellsPerPatient, "jesslyn_plots/numCellsPerPatient.csv")

numCellsPerTypePerPatient = table(cohort1$patient, cohort1$cell.type)%>% as.data.frame() %>% rename(patient = Var1, celltype = Var2, numCells = Freq)
ggplot(numCellsPerTypePerPatient, aes(x=celltype, fill=patient, y=numCells)) + geom_bar(position="fill", stat="identity")

numCellsPerCellTypePerSample = table(cohort1$sample, cohort1$cell.type) %>% as.data.frame() %>% rename(sample = Var1, celltype = Var2, numCells = Freq)
ggplot(numCellsPerCellTypePerSample, aes(x=celltype, fill=sample, y=numCells)) + geom_bar(position="fill", stat="identity")


#write_csv(numCellsPerTypePerPatient, "jesslyn_plots/numCellsPerTypePerPatient.csv")
numPatientsPerTreatment = table(cohort1$patient, cohort1$treatment.Status)

numCellsPerTreatmentPerCellType = table(cohort1$cell.type, cohort1$treatment.Status) %>% as.data.frame() %>% rename(celltype = Var1, treatment = Var2, numCells = Freq)
ggplot(numCellsPerTreatmentPerCellType, aes(x=celltype, fill=treatment, y=numCells)) + geom_bar(position="fill", stat="identity")

numCellsPerTreatmentPerSample = table(cohort1$sample, cohort1$treatment.Status) %>% as.data.frame() %>% rename(sample = Var1, treatment=Var2, numCells = Freq)
ggplot(numCellsPerTreatmentPerSample, aes(x=treatment, fill=sample, y=numCells)) + geom_bar(position="fill", stat="identity")

```

```{r cellCycle}
cohort1 <- CellCycleScoring(cohort1, g2m.features = g2m.genes, s.features = s.genes)

#distribution of cells in each cell cycle phase for each cell type 
table(cohort1$cell.type, cohort1$Phase) %>% as.data.frame() %>% rename(Celltype = Var1, Phase = Var2, numCells = Freq) %>% ggplot(aes(x=Celltype, y=numCells, fill = Phase)) + geom_bar(position="fill", stat="identity")
```

```{r score cells}
#score house keeping genes
cohort1 <- AddModuleScore(cohort1, features = list(hkgenes), name = "hkgenes")
```

```{r subset malignant cells vs. nonmalignant cells}
cohort1Malignant <- subset(cohort1, subset = (cell.type == "Malignant"))
cohort1NonMalignant <- subset(cohort1, subset = (cell.type != "Malignant"))
```

```{r PCA for ALL}
#identify highly variable genes 
cohort1 <- FindVariableFeatures(cohort1, selection.method = "vst") #wasn't sure if this was already done

#scale data 
cohort1 <- ScaleData(cohort1, features = rownames(cohort1), do.scale = FALSE, do.center = TRUE) #data is already scaled, but have to "scale" it for pca 

#run PCA
cohort1 <- RunPCA(cohort1, features = VariableFeatures(cohort1))



#plot PCA - shows differential expresison between different cohorts
PCAPlot(cohort1)
ggsave("cohort1_pcaplot.jpeg", path="jesslyn_plots/cohort1/All")

#separate by treatment? 
PCAPlot(cohort1, group.by="treatment.Status")

#separate by cellcycle phase?
PCAPlot(cohort1, group.by="Phase")

#separate by patient?
PCAPlot(cohort1, group.by="patient")

#separate by cluster? 
PCAPlot(cohort1, group.by="clst")

#determine dimensionality of data (how many pcs to include)
ElbowPlot(cohort1)
#graph shows probably pc1 to pc15 are important
ggsave("cohort1_pcaDim.jpeg", path="jesslyn_plots")

DotPlot(cohort1, features = VariableFeatures(cohort1)[1:5])
ggsave("cohort1_pca_dotplot.jpeg", path = "jesslyn_plots")
DoHeatmap(cohort1, features = VariableFeatures(cohort1)[1:10], group.by = 'ident', size = 2)
ggsave("cohort1_pca_heatmap.jpeg", path = "jesslyn_plots", width = 10)
```

```{r run UMAP for ALL}
#run UMAP, specify dimensions previously determined
cohort1<- RunUMAP(cohort1, dims=1:15)

#plot UMAP- shows clustering of cohorts
DimPlot(cohort1, reduction = "umap")
ggsave("cohort1_umap_celltype_plot.jpeg", path = "jesslyn_plots/cohort1/All")

```

#FOCUS ON MALIGNANT CELLS vs. NONMALIGNANT CELLS

```{r patient metrics}
numMalCellsPerSample <- table(cohort1Malignant$sample, cohort1Malignant$cell.type) %>% as.data.frame() %>% rename(sample = Var1, numCells=Freq, celltype = Var2)
ggplot(numMalCellsPerSample, aes(x=celltype, fill=sample, y=numCells)) + geom_bar(stat="identity", position="fill")

numMalCellsPerPatientPerTreatment <- table(cohort1Malignant$patient, cohort1Malignant$treatment.Status)

numMalCellsPerSamplePerTreatment<-table(cohort1Malignant$sample, cohort1Malignant$treatment.Status) %>% as.data.frame() %>% rename(sample = Var1, treatment = Var2, numCells = Freq)
ggplot(numMalCellsPerSamplePerTreatment, aes(x=treatment, fill=sample, y=numCells)) + geom_bar(stat="identity", position="fill")

numMalCellsPerSamplePerPhase <- table(cohort1Malignant$sample, cohort1Malignant$Phase) %>% as.data.frame() %>% rename(sample= Var1, phase = Var2, numCells= Freq)
ggplot(numMalCellsPerSamplePerPhase, aes(x=sample, y=numCells, fill=phase)) + geom_bar(position ="fill", stat ="identity")

numNonMalCellsPerPatientPerTreatment <- table(cohort1NonMalignant$patient, cohort1NonMalignant$treatment.Status)

```

```{r identify variable genes and run PCA for MALIGNANT an NONMALIGNANT subsets}
#MALIGNANT

#identify highly variable genes 
cohort1Malignant <- FindVariableFeatures(cohort1Malignant, selection.method = "vst") 
#scale data 
cohort1Malignant <- ScaleData(cohort1Malignant, features = rownames(cohort1Malignant), do.scale = FALSE, do.center = TRUE) 

cohort1Malignant <- RunPCA(cohort1Malignant, features = VariableFeatures(cohort1Malignant))

PCAPlot(cohort1Malignant)
ggsave("cohort1Malignant_pcaplot.jpeg", path="jesslyn_plots/cohort1/Malignant")

#NONMALIGNANT 

cohort1NonMalignant <- FindVariableFeatures(cohort1NonMalignant, selection.method = "vst") 
cohort1NonMalignant <- ScaleData(cohort1NonMalignant, features = rownames(cohort1NonMalignant), do.scale = FALSE, do.center = TRUE) #data is already scaled, but have to "scale" it for pca 

cohort1NonMalignant <- RunPCA(cohort1NonMalignant, features = VariableFeatures(cohort1NonMalignant))
PCAPlot(cohort1NonMalignant)
ggsave("cohort1NonMalignant_pcaplot.jpeg", path="jesslyn_plots/cohort1/NonMalignant")
```

```{r run UMAP for Malignant and Nonmalignant subsets}

#MALIGNANT 
#run UMAP, specify dimensions previously determined

cohort1Malignant<- RunUMAP(cohort1Malignant, dims = 1:15)
#plot UMAP- shows clustering
DimPlot(cohort1Malignant, reduction = "umap", pt.size = 0.3)
ggsave("cohort1_umap_Malignant_plot.jpeg", path = "jesslyn_plots/cohort1/Malignant")

#NONMALIGNANT
#run UMAP, specify dimensions previously determined
cohort1NonMalignant<- RunUMAP(cohort1NonMalignant, dims = 1:15)

#plot UMAP- shows clustering of cohorts
DimPlot(cohort1NonMalignant, reduction = "umap")

ggsave("cohort1_umap_Nonmalignant_plot.jpeg", path = "jesslyn_plots/cohort1/NonMalignant")
```

```{r save SDS}
saveRDS(cohort1, file="data/Izar_2020/jesslyn_cohort1_processed.RDS")
saveRDS(cohort1Malignant, file="data/Izar_2020/jesslyn_cohort1Malignant_processed.RDS")
saveRDS(cohort1NonMalignant, file="data/Izar_2020/jesslyn_cohort1NonMalignant_processed.RDS")
```