---
title: "PDX Exploratory Analysis (test version)"
author: "Jesslyn Goh"
date: "6/18/2020"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = here::here())
knitr::opts_chunk$set(echo = TRUE)
```

# PART 1. PDX DATA EXPLORATORY ANALYSIS 
This is a modified version of our previous PDX Analysis. Previously, we subsetted each PDX model after normalizing and centering data across all cells. We noticed that this might have caused issues in our downstream DE analyses, as the differences between models are so drastic that it may have masked over the smaller differences between treatment conditions within each model. Instead of doing only one workflow for all of the cells in the PDX experiment, we therefore decided to do four identical workflows in parallel for each Seurat object (All models, DF20, DF101, DF68) to detect differences aross models (using the 'PDX_All" object) and also within models (using 'PDX_{model}' objects). 

In this R Markdown file, we will: 
  - Load in the normalized but uncentered PDX Seurat Object 
  - Subset the PDX Seurat Object for each model (DF20, DF101, DF68)
  - Process and Analyze **each** Seurat Object **separately** 
      - Scale and center data for each model (do.scale = do.center = T)
      - SetAssayData to replace data in "data" slot with "scale.data"
      - AddModuleScore for each hallmark and AddCellCycleScore 
      - Dimensionality reduction (PCA) and UMAPP 
      - DE Analysis 
          - VlnPlot and DotPlot to visualize expression score for each hallmark of interest 
          - Volcano Plot to look for individual DE genes 
          - GSEA to look for enrichment of entire genesets / hallmarks 

## STEP 1. LOAD PACKAGES, SEURAT OBJECTS, AND GENESETS + SUBSET SEURAT OBJECTS FOR EACH MODEL

```{r init, message = FALSE}
# Load packages
source(here::here('packages.R'))

#Read in PDX RDS object
PDX_All = readRDS("data/Izar_2020/Izar_2020_PDX.RDS")

#Read in genelists 
ccgenes = read_lines("gene_lists/regev_lab_cell_cycle_genes.txt")
s.genes <- ccgenes[1:43]
g2m.genes <- ccgenes[44:97]

#Read in hallmarks of interest
hallmark_names = read_lines("gene_lists/hallmarks.txt")
hallmark.list <- vector(mode = "list", length = length(hallmark_names))
names(hallmark.list) <- hallmark_names
for(hm in hallmark_names){
  file <- read_lines(glue("hallmarks/{hm}_updated.txt"), skip = 1)
  hallmark.list[[hm]] <- file
}

#Subset Seurat Object for each model 
PDX_DF20 <- subset(PDX_All, subset = (model_ID == "DF20"))
PDX_DF101 <- subset(PDX_All, subset = (model_ID == "DF101"))
PDX_DF68 <- subset(PDX_All, subset = (model_ID == "DF68"))
```

## STEP 2. NORMALIZE AND SCALE EACH SEURAT OBJECT 
Normalize and scale each Seurat Object separately so that differences within models is not masked by differences across models. 

Purpose of normalization and scaling:
    - **Normalization** attempts to make the zero-inflated distribution **normal** by bringing outliers closer to the mean 
    - **Scaling and Centering** standardizes the normal distribution with mean 0 and sd 1 
        - do.center centers the expression for each feature by subtracting the average expression for that feature (mean expression at 0)
        - do.scale divides the centered feature expression by their sd (sd of 1) if center is TRUE

```{r Normalization and scaling, message = FALSE}
#scale data
PDX_All <- ScaleData(PDX_All, features = rownames(PDX_All), do.scale = T, do.center = T)
PDX_DF20 <- ScaleData(PDX_DF20, features = rownames(PDX_DF20), do.scale = T, do.center = T)
PDX_DF101 <- ScaleData(PDX_DF101, features = rownames(PDX_DF101), do.scale = T, do.center = T)
PDX_DF68 <- ScaleData(PDX_DF68, features = rownames(PDX_DF68), do.scale = T, do.center = T)

#feature selection (prepare for dimensionality reduction)
PDX_All <- FindVariableFeatures(PDX_All, selection.method = "vst")
PDX_DF20 <- FindVariableFeatures(PDX_DF20, selection.method = "vst")
PDX_DF101 <- FindVariableFeatures(PDX_DF101, selection.method = "vst")
PDX_DF68 <- FindVariableFeatures(PDX_DF68, selection.method = "vst")

#SetAssayData 
scale.data <- GetAssayData(object = PDX_All, slot = "scale.data")
PDX_All <- SetAssayData(PDX_All, slot = "data", new.data =scale.data, assay = "RNA")

scale.data <- GetAssayData(object = PDX_DF20, slot = "scale.data")
PDX_DF20 <- SetAssayData(PDX_DF20, slot = "data", new.data =scale.data, assay = "RNA")

scale.data <- GetAssayData(object = PDX_DF101, slot = "scale.data")
PDX_DF101 <- SetAssayData(PDX_DF101, slot = "data", new.data =scale.data, assay = "RNA")

scale.data <- GetAssayData(object = PDX_DF68, slot = "scale.data")
PDX_DF68 <- SetAssayData(PDX_DF68, slot = "data", new.data =scale.data, assay = "RNA")

beep(3)
```

```{r set treatment levels}
treatment_levels <- c("vehicle", "MRD", "relapse")
PDX_All$treatment.status = factor(PDX_All$treatment.status, levels = treatment_levels)
PDX_DF20$treatment.status = factor(PDX_DF20$treatment.status, levels = treatment_levels)
PDX_DF101$treatment.status = factor(PDX_DF101$treatment.status, levels = treatment_levels)
PDX_DF68$treatment.status = factor(PDX_DF68$treatment.status, levels = treatment_levels)
```

# STEP 3. DIMENSIONALITY REDUCTION (PCA)
This step helps reduce all of our features to ones that contribute most to the variability between cells. 
  1) We would first need to find the Variable Features by running FindVariableFeatures 
  2) run PCA 
```{r PCA, message = FALSE}
#PCA
PDX_All <- RunPCA(PDX_All, features = VariableFeatures(PDX_All))
PDX_DF20 <- RunPCA(PDX_DF20, features = VariableFeatures(PDX_DF20))
PDX_DF101 <- RunPCA(PDX_DF101, features = VariableFeatures(PDX_DF101))
PDX_DF68 <- RunPCA(PDX_DF68, features = VariableFeatures(PDX_DF68))
```

## STEP 4. CELLCYCLE AND HALLMARK SCORING
Scoring each cell by their level expression of all the genes in a specified geneset. Prepares for future DE analysis of GO hallmarks. 
```{r cell scoring, message = FALSE}
# cell cycle scoring 
PDX_All <- CellCycleScoring(PDX_All, g2m.features = g2m.genes, s.features = s.genes)
PDX_DF20 <- CellCycleScoring(PDX_DF20, g2m.features = g2m.genes, s.features = s.genes)
PDX_DF101 <- CellCycleScoring(PDX_DF101, g2m.features = g2m.genes, s.features = s.genes)
PDX_DF68 <- CellCycleScoring(PDX_DF68, g2m.features = g2m.genes, s.features = s.genes)

#hallmark scoring
PDX_All <- AddModuleScore(PDX_All, features = hallmark.list, name = names(hallmark.list), nbin = 25, search = T)
PDX_DF20 <- AddModuleScore(PDX_DF20, features = hallmark.list, name = names(hallmark.list), nbin = 25, search = T)
PDX_DF101 <- AddModuleScore(PDX_DF101, features = hallmark.list, name = names(hallmark.list), nbin = 25, search = T)
PDX_DF68 <- AddModuleScore(PDX_DF68, features = hallmark.list, name = names(hallmark.list), nbin = 25, search = T)

beep(3)
```

## STEP 5. EXAMINE HOW CELLS CLUSTER IN UMAP SPACE 
Exploring how cells clusters - whether treatment conditions result in DE that makes cells cluster in a certain way. 

```{r UMAPP, message = FALSE}
#cluster by model? 
PDX_All<- RunUMAP(PDX_All, dims=1:15)
UMAPPlot(PDX_All, group.by = "model_ID")
ggsave("PDX_All_UMAP.png", path = "jesslyn_plots/PDX_test")

#cluster by treatment within each model? 
PDX_DF20<- RunUMAP(PDX_DF20, dims=1:15)
UMAPPlot(PDX_DF20, group.by = "treatment.status")
ggsave("DF20_UMAP.png", path = "jesslyn_plots/PDX_test")

PDX_DF101<- RunUMAP(PDX_DF101, dims=1:15)
UMAPPlot(PDX_DF101, group.by = "treatment.status")
ggsave("DF101_UMAP.png", path = "jesslyn_plots/PDX_test")


PDX_DF68<- RunUMAP(PDX_DF68, dims=1:15)
UMAPPlot(PDX_DF68, group.by = "treatment.status")
ggsave("DF68_UMAP.png", path = "jesslyn_plots/PDX_test")
```
Results from UMAPs show that they do cluster pretty clearly by model, but not by treatment. 

```{r save RDS}
saveRDS(PDX_All, file="data/Izar_2020/test/jesslyn_PDX_All_processed.RDS")
saveRDS(PDX_DF20, file="data/Izar_2020/test/jesslyn_PDX_DF20_processed.RDS")
saveRDS(PDX_DF101, file="data/Izar_2020/test/jesslyn_PDX_DF101_processed.RDS")
saveRDS(PDX_DF68, file="data/Izar_2020/test/jesslyn_PDX_DF68_processed.RDS")
```
