---
title: "Izar 2020 PDX and SS2 data specific questions"
author: "Mike Cuoco"
output:
  html_document:
    theme: united
    df_print: kable
date: 'Compiled: `r format(Sys.Date(), "%B %d, %Y")`'
editor_options: 
  chunk_output_type: console
---
***

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  tidy = TRUE,
  tidy.opts = list(width.cutoff = 120),
  message = FALSE,
  warning = FALSE
)
knitr::opts_knit$set(root.dir = here::here())
set.seed(10)
source('packages.R')
```

```{r init}
# saving obj setup
proj = "Izar_2020"
date = Sys.Date()

# setup directories
if (!dir.exists(here("mike_plots",proj))){
  dir.create(here("mike_plots",proj))
  dir.create(here("mike_plots",proj,date))
} else if (!dir.exists(here("mike_plots",proj,date))){
  dir.create(here("mike_plots",proj,date))
}

fig.dir = here("mike_plots",proj,date)

# Signatures
ovca_sigs = read_csv("gene_lists/Ovarian_Signatures_combined.csv")[,1:26] %>% 
  as.list() %>%
  lapply(function(x) x[!is.na(x)])
hallmarks = read_lines("gene_lists/hallmarks.txt")
hall_sigs = msigdbr(species = "Homo sapiens", category = "H") %>% 
  filter(gs_name %in% hallmarks) %>% 
  select(gs_name, gene_symbol) %>% 
  dlply("gs_name", function(x) pull(x,gene_symbol))
hall_sigs = msigdbr(species = "Homo sapiens") %>%
  filter(gs_name %in% c("GO_RESPONSE_TO_ENDOPLASMIC_RETICULUM_STRESS",
                  "GO_REGULATION_OF_RESPONSE_TO_ENDOPLASMIC_RETICULUM_STRESS")) %>%
  select(gs_name, gene_symbol) %>%
  dlply("gs_name", function(x) pull(x,gene_symbol)) %>%
  c(hall_sigs)
sigs = names(hall_sigs)

# Read in processed SS2 and PDX Seurat objects
ovca_SS2 = list.files(here("data","Izar_2020"), full.names = T) %>% 
  grep("preproc",.,value=T) %>%
  grep("SS2",.,value=T) %>%
  readRDS()
ovca_PDX = list.files(here("data","Izar_2020"), full.names = T) %>% 
  grep("preproc",.,value=T) %>%
  grep("PDX",.,value=T) %>%
  readRDS()
```


# 1. PDX

by treatment status and model

## Vln plots
```{r PDX vln plots}
# setup
feats = c("HALLMARK_UNFOLDED_PROTEIN_RESPONSE","HALLMARK_OXIDATIVE_PHOSPHORYLATION")
Idents(ovca_PDX) <- "model_ID"
ovca_PDX$treatment.status = factor(x = ovca_PDX$treatment.status , levels = c("vehicle","MRD","relapse"))

# Vln Plots
plots = VlnPlot(ovca_PDX, features = feats, split.by = "treatment.status", combine = F, slot = "scale.data") 
names(plots) = feats
plots = lapply(feats, function(x) {
      plots[[x]] = plots[[x]] + 
        guides(fill=guide_legend("Treatment Status")) + 
        labs(x = "model", y = "normalized expression") + 
        theme(plot.title = element_text(size = 10))
      })
names(plots) = feats
lapply(feats, function(x) {
  p = plots[[x]] + theme(plot.title = element_text(size = 14))
  ggsave(glue('{fig.dir}/PDX_{x}_vln.png'), p, width = 8, height = 6)
  })
p = wrap_plots(plots) + plot_layout(guides = "collect") 
ggsave(glue('{fig.dir}/PDX_vlns.png'), width = 12, height = 6, plot = p)
```

## Dot Plots
```{r PDX DotPlots}
# Setup
feats = c("HALLMARK_UNFOLDED_PROTEIN_RESPONSE","HALLMARK_OXIDATIVE_PHOSPHORYLATION")
Idents(ovca_PDX) <- "treatment.status"
ovca_PDX$treatment.status = factor(x = ovca_PDX$treatment.status , levels = c("vehicle","MRD","relapse"))
# colors = c("#FA0400","green","red")
# DotPlot3(ovca_PDX, features = feats, group.by = "treatment.status", split.by = "model_ID")

# will all hallmark signatures
plots = map(unique(ovca_PDX$model_ID), function(.x){
  p = subset(ovca_PDX, model_ID == .x) %>% 
    DotPlot(features = names(hall_sigs), group.by = "treatment.status") + 
    coord_flip() + RotatedAxis() + 
    labs(title = .x, x = "Gene Signature", y = "Treatment Status")  + 
    theme(axis.text = element_text(size = 8), 
          plot.title = element_text(size = 12),
          legend.position = "none",
          axis.title = element_blank())
  ggsave(glue('{fig.dir}/PDX_{.x}_hallsigs_ts_dot.png'),plot = p, width = 6, height = 12)
  return(p)
})

# with selected signatures
plots = map(unique(ovca_PDX$model_ID), function(.x){
  p = subset(ovca_PDX, model_ID == .x) %>% 
    DotPlot(features = feats, group.by = "treatment.status") + 
    coord_flip() + RotatedAxis() + 
    labs(title = .x, x = "Gene Signature", y = "Treatment Status")  + 
    theme(axis.text = element_text(size = 8), 
          plot.title = element_text(size = 12),
          legend.position = "none",
          axis.title = element_blank())
  ggsave(glue('{fig.dir}/PDX_{.x}_feats_ts_dot.png'),plot = p, width = 7, height = length(feats))
  return(p)
})
plots = map(plots, function(p){
  p = p + theme(axis.title = element_blank(), legend.position = "none") + guides(radius = F)
  return(p)
})
wrap_plots(plots, ncol=1, guides = "collect") %>% 
  ggsave(filename=glue('{fig.dir}/PDX_all_models_feats_ts_dot.png'),plot = .,width = 7, height = 3*length(feats))

# Dot Plots
## Use custom function
plots = DotPlot2(ovca_PDX, features = feats, group1 = "model_ID", group2 = "treatment.status")
plots = lapply(plots, function(x) {
                x = x + coord_flip() + theme(plot.title = element_text(size = 12))
                })
wrap_plots(plots, guides = "collect", ncol = 1) %>% 
  ggsave(glue('{fig.dir}/PDX_dotplots.png'), width = 7, height = 14, plot = .)
```

## 3-way GSEA
```{r PDX 3-way GSEA}
this.fig.dir = file.path(fig.dir,"PDX_3way_GSEA")
if(!dir.exists(this.fig.dir)){dir.create(this.fig.dir)}
# GSEA
gsea_sigs = hall_sigs
group1 = unique(ovca_PDX$model_ID)
group2 = levels(ovca_PDX$treatment.status)
data.gsea = lapply(group1, function(x) { # iterate through models
  # DE between treatment statuses within each model
  data.use = subset(ovca_PDX, subset = model_ID == x) %>% 
    wilcoxauc("treatment.status", assay = 'scale.data')
  data.list = lapply(group2, function(y){ # iterates through treatment statuses
      data.use = dplyr::filter(data.use, group == y) %>%
        dplyr::arrange(desc(logFC)) %>%
        dplyr::select(feature, logFC)
      ranks = deframe(data.use)
      gseaRes = fgsea(gsea_sigs, ranks, minSize = 15, maxSize = 500, nperm=10000) %>%
        as_tibble() %>%
        dplyr::arrange(desc(NES))
      return(list("ranks" = ranks, "res" = gseaRes))
    })
})
data.gsea = setNames(lapply(data.gsea, setNames, group2), group1)
lapply(names(gsea_sigs), function(z){
  lapply(group1, function(x) {
    data.use = data.gsea[[x]]
    plots = lapply(group2, function(y){
      ranks = data.use[[y]]$ranks
      p = plotEnrichment(gsea_sigs[[z]], ranks) + 
          labs(title = z,subtitle = glue("model_ID: {x} | treatment.status: {y}"))
      return(p)
      })
    wrap_plots(plots, ncol = 1, nrow = 3) %>% 
      ggsave(filename = glue('{this.fig.dir}/PDX_{x}_{z}_GSEA-LE.png'), width = 5, height = 10, plot = .)
    })
})

## Summarize GSEA with tables for each pathway
data.plot = lapply(data.gsea, function(x) ldply(x, function(y) y$res, .id = "treatment.status")) %>% 
    ldply(., .id = "model_ID") %>%
    dplyr::select(-leadingEdge, -nMoreExtreme,  -pval) %>%
    group_by(model_ID)
data.plot$treatment.status = factor(data.plot$treatment.status, levels = c("vehicle","MRD","relapse"))
lapply(names(gsea_sigs), function(y) {
    data.use = filter(data.plot, pathway == y) %>% 
      dplyr::select(-pathway)
    size = unique(data.use$size)
    data.use %>% 
      dplyr::select(-size) %>%
      dplyr::arrange(treatment.status) %>%
      gt(rowname_col = "treatment.status") %>%
      tab_header(title = glue("Gene Set Enrichment Analysis"),  subtitle = glue("PDX data: {y} ({size} genes)")) %>%
      tab_footnote("GSEA was performed on wilcoxauc DE log-fold-change values comparing treatments within each model", cells_stub()) %>%
      tab_footnote("Padj = FDR-adjusted p-value (p = ES(sig) vs null distribution of ES permutations", cells_column_labels(c("padj"))) %>%
      tab_footnote("Enrichment Score", cells_column_labels(c("ES"))) %>%
      tab_footnote("Normalized Enrichment Score = ES/(mean(ESs of all dataset permutations))", cells_column_labels(c("NES"))) %>%
      data_color(vars(NES), colors = RColorBrewer::brewer.pal(9, "YlOrRd")) %>%
      fmt_number(vars(padj, ES,NES),decimals = 4) %>%
      tab_style(style = list(cell_fill(color = "lightcyan"),cell_text(weight = "bold")),
                locations = cells_body(columns = vars(padj),rows = padj < 0.05)) %>%
      tab_options(
        row_group.background.color = "lightgray",
        heading.title.font.size = 20 ,
        heading.title.font.weight = "bold",
        column_labels.background.color = "#B4DCF7",
        column_labels.font.size = 20,
        column_labels.font.weight = "bold"
      ) %>%
      gtsave(glue("PDX_GSEA_{y}_tab.png"), expand = 10, path = this.fig.dir)
})

```

# 2. SS2 
## Vln plots
```{r SS2 vln}
feats = c("HALLMARK_UNFOLDED_PROTEIN_RESPONSE","HALLMARK_OXIDATIVE_PHOSPHORYLATION","ATF3","ATF4","JUN")
pats = c(8,9)
ovca_SS2 = subset(ovca_SS2, subset = cell.type == "Malignant") %>%  subset(subset = Patient %in% pats) 
Idents(ovca_SS2) = "Patient"
ovca_SS2@active.ident = factor(x = ovca_SS2@active.ident , levels = c(sort(unique(ovca_SS2$Patient))))

# Vln Plots
plots = VlnPlot(ovca_SS2,features = feats, split.by = "Time", combine = F)
names(plots) = feats
plots = lapply(feats, function(x) {
      plots[[x]] = plots[[x]] + 
        guides(fill=guide_legend("Time")) + 
        labs(x = "Patient") + theme(plot.title = element_text(size = 10))
      })
names(plots) = feats
lapply(feats, function(x) {
  p = plots[[x]] + theme(plot.title = element_text(size = 14))
  ggsave(glue('{fig.dir}/SS2_{x}_vln.png'), p, width = 8, height = 6)
  })
p = wrap_plots(plots) + plot_layout(guides = "collect") 
ggsave(glue('{fig.dir}/SS2_vlns.png'), width = 12, height = 12, plot = p)
```
## DotPlots
```{r SS2 DotPlots}
# Setup
feats = c("HALLMARK_UNFOLDED_PROTEIN_RESPONSE","HALLMARK_OXIDATIVE_PHOSPHORYLATION","ATF3","ATF4","JUN")
pats = c(8,9)
ovca_SS2 = subset(ovca_SS2, subset = cell.type == "Malignant") %>%  subset(subset = Patient %in% pats) 
Idents(ovca_SS2) = "Patient"
ovca_SS2@active.ident = factor(x = ovca_SS2@active.ident , levels = c(sort(unique(ovca_SS2$Patient))))

# Dot Plots
## Use custom function
ovca_SS2$Time = factor(ovca_SS2$Time, levels = c("0","1","2"))
plots = DotPlot2(ovca_SS2, features = feats, group1 = "Patient", group2 = "Time")
plots = lapply(plots, function(x) {
                x = x + coord_flip() + theme(plot.title = element_text(size = 12))
                })
wrap_plots(plots) %>% ggsave(glue('{fig.dir}/SS2_dotplots.png'), width = 7, height = 14, plot = .)
```

## 3-way GSEA
```{r SS2 3-way GSEA}
# Setup
this.fig.dir = file.path(fig.dir,"SS2_3way_GSEA")
if(!dir.exists(this.fig.dir)){dir.create(this.fig.dir)}
gsea_sigs = hall_sigs[c("HALLMARK_UNFOLDED_PROTEIN_RESPONSE","HALLMARK_OXIDATIVE_PHOSPHORYLATION")]
pats = c(8,9)
ovca_SS2 = subset(ovca_SS2, subset = cell.type == "Malignant") %>%  subset(subset = Patient %in% pats) 
Idents(ovca_SS2) = "Patient"
ovca_SS2@active.ident = factor(x = ovca_SS2@active.ident , levels = c(sort(unique(ovca_SS2$Patient))))

group1 = unique(ovca_SS2$Patient) %>% as.character()
group2 = unique(ovca_SS2$Time) %>% as.character()
data.gsea = lapply(group1, function(x) {
  # DE between treatment statuses within each model
  data.use = subset(ovca_SS2, subset = Patient == x) %>% 
    wilcoxauc("Time", assay = 'scale.data')
  data.list = lapply(group2, function(y){
      data.use = dplyr::filter(data.use, group == y) %>%
        dplyr::arrange(desc(logFC)) %>%
        dplyr::select(feature, logFC)
      ranks = deframe(data.use)
      gseaRes = fgsea(gsea_sigs, ranks, minSize = 15, maxSize = 500, nperm=1000) %>%
        as.tibble() %>%
        dplyr::arrange(desc(NES))
      return(list("ranks" = ranks, "res" = gseaRes))
    })
})
data.gsea = setNames(lapply(data.gsea, setNames, group2), group1)
lapply(names(gsea_sigs), function(z){
  lapply(group1, function(x) {
    data.use = data.gsea[[x]]
    plots = lapply(group2, function(y){
      ranks = data.use[[y]]$ranks
      p = plotEnrichment(gsea_sigs[[z]], ranks) + 
          labs(title = z,subtitle = glue("Patient: {x} | Time: {y}"))
      return(p)
    })
    wrap_plots(plots, ncol = 1, nrow = 3) %>% 
      ggsave(glue('{this.fig.dir}/SS2_{x}_{z}_GSEA-LE.png'), width = 5, height = 10, plot = .)
  })
})

## Summarize GSEA with tables for each pathway
data.plot = lapply(data.gsea, function(x) ldply(x, function(y) y$res, .id = "Time")) %>% 
    ldply(., .id = "Patient") %>%
    dplyr::select(-leadingEdge, -nMoreExtreme,  -pval) %>%
    group_by(Patient)
data.plot$Time = factor(data.plot$Time, levels = c("0","1","2"))
lapply(names(gsea_sigs), function(y) {
    data.use = filter(data.plot, pathway == y) %>% 
      dplyr::select(-pathway)
    size = unique(data.use$size)
    data.use %>% 
      dplyr::select(-size) %>%
      dplyr::arrange(Time) %>%
      gt(rowname_col = "Time") %>%
      tab_header(title = glue("Gene Set Enrichment Analysis"),  subtitle = glue("SS2 data: {y} ({size} genes)")) %>%
      tab_footnote("GSEA was performed on wilcoxauc DE log-fold-change values comparing Time within each Patient", cells_stub()) %>%
      tab_footnote("Padj = FDR-adjusted p-value (p = ES(sig) vs null distribution of ES permutations", cells_column_labels(c("padj"))) %>% 
      tab_footnote("Enrichment Score", cells_column_labels(c("ES"))) %>%
      tab_footnote("Normalized Enrichment Score = ES/(mean(ESs of all dataset permutations))", cells_column_labels(c("NES"))) %>%
      data_color(vars(NES), colors = RColorBrewer::brewer.pal(9, "YlOrRd")) %>%
      fmt_number(vars(padj, ES,NES),decimals = 3) %>%
      tab_style(style = list(cell_fill(color = "lightcyan"),cell_text(weight = "bold")),
                locations = cells_body(columns = vars(padj),rows = padj < 0.05)) %>%
      tab_options(
        row_group.background.color = "lightgray",
        heading.title.font.size = 20 ,
        heading.title.font.weight = "bold",
        column_labels.background.color = "#B4DCF7",
        column_labels.font.size = 20,
        column_labels.font.weight = "bold"
      ) %>%
      gtsave(glue("SS2_GSEA_{y}_tab.png"), expand = 10, path = this.fig.dir)
})
```

## 2-way GSEA 

```{r SS2 2-way GSEA}
# Setup
this.fig.dir = file.path(fig.dir,"SS2_2way_GSEA")
if(!dir.exists(this.fig.dir)){dir.create(this.fig.dir)}
gsea_sigs = hall_sigs[c("HALLMARK_UNFOLDED_PROTEIN_RESPONSE","HALLMARK_OXIDATIVE_PHOSPHORYLATION")]
pats = c(8,9)
ovca_SS2 = subset(ovca_SS2, subset = cell.type == "Malignant") %>%  subset(subset = Patient %in% pats) 
Idents(ovca_SS2) = "Time"
ovca_SS2@active.ident = factor(x = ovca_SS2@active.ident , levels = c(sort(unique(ovca_SS2$Patient))))

group1 = unique(ovca_SS2$Patient) %>% as.character()
ident2 = as.factor(c(1,2))
data.gsea = lapply(group1, function(x) {
  data.list = lapply(ident2, function(y) {
    data.use = subset(ovca_SS2, subset = Patient == x) %>% 
      FindMarkers(ident.1 = as.factor("0"),ident.2 = y, slot = 'data', logfc.threshold = 0) %>%
      dplyr::arrange(desc(avg_logFC)) %>%
      dplyr::select(avg_logFC) %>%
      rownames_to_column("features")
    ranks = deframe(data.use)
    gseaRes = fgsea(gsea_sigs, ranks, minSize = 15, maxSize = 500, nperm=1000) %>%
      as.tibble() %>%
      dplyr::arrange(desc(NES))
    return(list("ranks" = ranks, "res" = gseaRes))
    })
})
data.gsea = setNames(lapply(data.gsea, setNames, group2), group1)
lapply(names(gsea_sigs), function(z){
  lapply(group1, function(x) {
    data.use = data.gsea[[x]]
    plots = lapply(group2, function(y){
      ranks = data.use[[y]]$ranks
      p = plotEnrichment(gsea_sigs[[z]], ranks) + 
          labs(title = z,subtitle = glue("Patient: {x} | Time: {y}"))
      return(p)
    })
    wrap_plots(plots, ncol = 1, nrow = 3) %>% 
      ggsave(glue('{this.fig.dir}/SS2_{x}_{z}_GSEA-LE.png'), width = 5, height = 10, plot = .)
  })
})

## Summarize GSEA with tables for each pathway
data.plot = lapply(data.gsea, function(x) ldply(x, function(y) y$res, .id = "Time")) %>% 
    ldply(., .id = "Patient") %>%
    dplyr::select(-leadingEdge, -nMoreExtreme,  -pval) %>%
    group_by(Patient)
data.plot$Time = factor(data.plot$Time, levels = c("0","1","2"))
lapply(names(gsea_sigs), function(y) {
    data.use = filter(data.plot, pathway == y) %>% 
      dplyr::select(-pathway)
    size = unique(data.use$size)
    data.use %>% 
      dplyr::select(-size) %>%
      dplyr::arrange(Time) %>%
      gt(rowname_col = "Time") %>%
      tab_header(title = glue("Gene Set Enrichment Analysis"),  subtitle = glue("SS2 data: {y} ({size} genes)")) %>%
      tab_footnote("GSEA was performed on wilcoxauc DE log-fold-change values comparing Time within each Patient", cells_stub()) %>%
      tab_footnote("Padj = FDR-adjusted p-value (p = ES(sig) vs null distribution of ES permutations", cells_column_labels(c("padj"))) %>% 
      tab_footnote("Enrichment Score", cells_column_labels(c("ES"))) %>%
      tab_footnote("Normalized Enrichment Score = ES/(mean(ESs of all dataset permutations))", cells_column_labels(c("NES"))) %>%
      data_color(vars(NES), colors = RColorBrewer::brewer.pal(9, "YlOrRd")) %>%
      fmt_number(vars(padj, ES,NES),decimals = 3) %>%
      tab_style(style = list(cell_fill(color = "lightcyan"),cell_text(weight = "bold")),
                locations = cells_body(columns = vars(padj),rows = padj < 0.05)) %>%
      tab_options(
        row_group.background.color = "lightgray",
        heading.title.font.size = 20 ,
        heading.title.font.weight = "bold",
        column_labels.background.color = "#B4DCF7",
        column_labels.font.size = 20,
        column_labels.font.weight = "bold"
      ) %>%
      gtsave(glue("SS2_GSEA_{y}_tab.png"), expand = 10, path = this.fig.dir)
})
