---
title: "Izar 2020 10X (Cohort 1) DE Analysis"
author: "Jesslyn Goh and Mike Cuoco"
date: "7/19/2020"
editor_options: 
  chunk_output_type: console
output:
 html_document:
  code_folding: hide
  toc_float: TRUE 
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = here::here())
knitr::opts_chunk$set(echo = TRUE)
options(width=120)
```

# IZAR 2020 10X (COHORT 1) DATA ANALYSIS - PART 2

### OVERVIEW
* This is the second part of our 3-part analysis of the Izar 2020 10X (Cohort 1) data. 
* The 10X data consists of **CD45+ depleted 8 ascites samples from 6 patients** 
    - Includes both **malignant** and **nonmalignant** populations: Provides a broad view of the diverse cell types in the ascites ecosystem
    - We are only interested in the **malignant** ascites population

* We split our 10X analysis into three parts: 
    1. Loading and Processing Data
    2. Visualizing Clusters (PCA and UMAP)
    3. **DE Analysis (Focus on Malignant Cells)**
        a. Load Processed Seurat Objects
        b. DE Analysis Type #1. Find Individual DE Genes
            i. Call FindMarkers on **cohort1Malignant** by **treatment.status**
            ii. Visualize results using: 
                  - Heatmap
                  - Volcano Plot
        c. DE Analysis Type #2. Evaluate DE Hallmark Genesets 
            i. Call FindMarkers and rank genes for GSEA 

### STEP 3A. LOAD AND SUBSET SEURAT OBJECT

```{r init, message = FALSE, warning = FALSE}
# Load packages
source(here::here('packages.R'))

#Read in PDX RDS object
cohort1Malignant = readRDS("data/Izar_2020/jesslyn_cohort1Malignant_processed.RDS")

# Read in gene lists
ccgenes = read_lines("data/gene_lists/regev_lab_cell_cycle_genes.txt")
s.genes <- ccgenes[1:43]
g2m.genes <- ccgenes[44:97]

#Read in hallmarks of interest
hallmark_names = read_lines("data/gene_lists/hallmarks.txt")
hallmark.list <- vector(mode = "list", length = length(hallmark_names))
names(hallmark.list) <- hallmark_names
for(hm in hallmark_names){
  file <- read_lines(glue("data/gene_lists/hallmarks/{hm}_updated.txt"), skip = 1)
  hallmark.list[[hm]] <- file
}

if(!dir.exists("jesslyn_plots/cohort1/Malignant")){dir.create("jesslyn_plots/cohort1/Malignant")}
```

### STEP 3B.I CALL FINDMARKERS 
**not sure which statistical test to use yet**

```{r FindMarkers, message=FALSE, warning=FALSE}
tn.oc.markers <- FindMarkers_code(cohort1Malignant, group.by = "treatment.status", group.1 = "Treatment-naïve", group.2 = "After 1 cycle of chemotherapy")
rownames(tn.oc.markers) <- tn.oc.markers$rowname

tn.ot.markers <- FindMarkers_code(cohort1Malignant, group.by = "treatment.status", group.1 = "Treatment-naïve", group.2 = "On-treatment")
rownames(tn.ot.markers) <- tn.ot.markers$rowname

oc.ot.markers <- FindMarkers_code(cohort1Malignant, group.by = "treatment.status", group.1 = "After 1 cycle of chemotherapy", group.2 = "On-treatment")
rownames(oc.ot.markers) <- oc.ot.markers$rowname
```

### STEP 3B.II VISUALIZE RESULTS 
```{r visualize FindMarkers results, message=FALSE, warning = FALSE, fig.align='center', fig.width = 20, fig.height=10}
#treatment naive vs. one cycle of chemotherapy 
tn.oc.volcano <- DEAnalysis_code(cohort1Malignant, markers = tn.oc.markers, 
    group.by = "treatment.status", group.1 = "Treatment-naïve", group.2 = "After 1 cycle of chemotherapy", graph = TRUE) + 
    theme(plot.title = element_text(size = 8), plot.subtitle = element_text(size = 8))
tn.oc.volcano.oxphos <- DEAnalysis_code(cohort1Malignant, markers = tn.oc.markers, 
    group.by = "treatment.status", group.1 = "Treatment-naïve", 
    group.2 = "After 1 cycle of chemotherapy", geneset = hallmark.list[["HALLMARK_OXIDATIVE_PHOSPHORYLATION"]]) +
    theme(plot.title = element_text(size = 8), plot.subtitle = element_text(size = 8))
tn.oc.volcano.upr <- DEAnalysis_code(cohort1Malignant, markers = tn.oc.markers, 
    group.by = "treatment.status", group.1 = "Treatment-naïve", 
    group.2 = "After 1 cycle of chemotherapy", geneset = hallmark.list[["HALLMARK_UNFOLDED_PROTEIN_RESPONSE"]]) + 
    theme(plot.title = element_text(size = 8), plot.subtitle = element_text(size = 8))

tn.oc.volcano.all <- tn.oc.volcano + tn.oc.volcano.oxphos + tn.oc.volcano.upr
tn.oc.volcano.all
ggsave(plot = tn.oc.volcano.all, filename = "10X_Mal_TNvsOC_Volcano.png", path = "jesslyn_plots/cohort1/Malignant", width = 20, height = 10)

#treatment naive vs. on-treatment 
tn.ot.volcano <- DEAnalysis_code(cohort1Malignant, markers = tn.ot.markers, 
    group.by = "treatment.status", group.1 = "Treatment-naïve", group.2 = "On-treatment", 
    graph = TRUE) + 
    theme(plot.title = element_text(size = 8), plot.subtitle = element_text(size = 8))
tn.ot.volcano.oxphos <- DEAnalysis_code(cohort1Malignant, markers = tn.ot.markers, 
    group.by = "treatment.status", group.1 = "Treatment-naïve", 
    group.2 = "On-treatment", geneset = hallmark.list[["HALLMARK_OXIDATIVE_PHOSPHORYLATION"]]) + 
    theme(plot.title = element_text(size = 8), plot.subtitle = element_text(size = 8))
tn.ot.volcano.upr <- DEAnalysis_code(cohort1Malignant, markers = tn.ot.markers, 
    group.by = "treatment.status", group.1 = "Treatment-naïve", 
    group.2 = "On-treatment", geneset = hallmark.list[["HALLMARK_UNFOLDED_PROTEIN_RESPONSE"]]) + 
    theme(plot.title = element_text(size = 8), plot.subtitle = element_text(size = 8))

tn.ot.volcano.all <- tn.ot.volcano + tn.ot.volcano.oxphos + tn.ot.volcano.upr
tn.ot.volcano.all
ggsave(plot = tn.ot.volcano.all, filename = "10X_Mal_TNvsOT_Volcano.png", path = "jesslyn_plots/cohort1/Malignant", width =20, height = 10)

#one cycle of chemotherapy vs. on-treatment
oc.ot.volcano <- DEAnalysis_code(cohort1Malignant, markers = oc.ot.markers, 
    group.by = "treatment.status", group.1 = "After 1 cycle of chemotherapy", group.2 = "On-treatment", graph = TRUE) +
    theme(plot.title = element_text(size = 8), plot.subtitle = element_text(size = 8))
oc.ot.volcano.oxphos <- DEAnalysis_code(cohort1Malignant, markers = oc.ot.markers, 
    group.by = "treatment.status", group.1 = "After 1 cycle of chemotherapy", 
    group.2 = "On-treatment", geneset = hallmark.list[["HALLMARK_OXIDATIVE_PHOSPHORYLATION"]]) + 
    theme(plot.title = element_text(size = 8), plot.subtitle = element_text(size = 8))
oc.ot.volcano.upr <- DEAnalysis_code(cohort1Malignant, markers = oc.ot.markers, 
    group.by = "treatment.status", group.1 = "After 1 cycle of chemotherapy", 
    group.2 = "On-treatment", geneset = hallmark.list[["HALLMARK_UNFOLDED_PROTEIN_RESPONSE"]]) + 
    theme(plot.title = element_text(size = 8), plot.subtitle = element_text(size = 8))

oc.ot.volcano.all <- oc.ot.volcano + oc.ot.volcano.oxphos + oc.ot.volcano.upr
oc.ot.volcano.all
ggsave(plot = oc.ot.volcano.all, filename = "10X_Mal_OCvsOT_Volcano.png", path = "jesslyn_plots/cohort1/Malignant", width = 20, height = 10)
```

### STEP 4A.I GSEA
**still not sure if we should rank by AUC or logFC, will edit this part after we decide**
