---
title: "Izar 2020 SS2 (Cohort 2) Load + Plots"
author: "Jesslyn Goh and Mike Cuoco"
date: "7/20/2020"
editor_options: 
  chunk_output_type: console
output:
 html_document:
  code_folding: hide
  toc_float: TRUE 
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = here::here())
knitr::opts_chunk$set(echo = TRUE)
options(width=120)
```

# IZAR 2020 SS2 (COHORT 2) DATA ANALYSIS - PART 1

### OVERVIEW
* The SS2 data from Izar2020 consists of cells from 14 ascites samples from 6 individuals: 
    - selected for malignant EPCAM+CD24+ cells by fluorophore-activated cell sorting into 96-well plates
    - Includes both **malignant** and **nonmalignant** populations but differs from 10X data - examines malignant ascites more closely. 
    - We are only interested in the **malignant** ascites population, specifically samples from **patient 8 and 9** because they are the only patients that have samples from different treatment statuses. 
    
* In our 3-part analysis of the Izar 2020 SS2 data, we are interested in identifying differentially expressed genes and hallmark genesets between treatment statuses **within** each model. 
* We split our SS2 analysis into three parts: 
    1. **Load Data and Create SS2 Seurat Object**
        a. The code to this part of our analysis is in the **read_Izar_2020.R** file in the **code** folder. During this part of our analysis we: 
            i. Load in SS2 count matrix and Create Seurat Object 
            ii. Assign Metadata identities including: 
                - Patient ID
                - Time
                - Sample ID
                - Treatment Status 
            iii. Score cells for **cell cycle** and **hallmark genesets** 
                - **Note**: It does not matter whether we call AddModuleScore before or after subsetting and scaling each model because AddModuleScore uses the **data** slot. 
            iv. Save Seurat Object
          
    2. **Process Data and Exploratory Data Analysis**
        a. The code to this part of our analysis can be found in the **03_Izar2020_SS2_Load_Plots.Rmd** file in the **old/edited** folder. During this part of our analysis we: 
            i. Load in SS2 Seurat Object from Part 1 and **subset for Malignant cells then for patient 8 and 9**. Continue analysis separately for each patient. 
            ii. Scale and FindVariableFeatures (prepares data for dimensionality reduction)
            iii. Dimensionality Reduction (PCA + UMAP)
            iv. Visualize how cells separate based on metadata identities via UMAP 
                  - Intermodel heterogeneity: How do cells separate by model? 
                  - Intramodel heterogeneity: 
                      * How do cells separate by treament status? 
                      * How do cells separate by cell cycle phase? 
            v. Compute summary metrics for SS2 data such as:
                - Number of cells per patient per treatment 
                - Number of cells per treatment per cell cycle phase 
            vi. Save Seurat Objects 
            
    3. **DE Analysis**
        a. **TYPE #1 DE ANALYSIS**: Visualizing and Quantifying Differentially Expression on **Predefined** GO Genesets
            i. Violin Plots and UMAP 
            ii. Gene Set Enrichment Analysis (GSEA)
        b. **TYPE #2 DE ANALYSIS**: Finding DE Genes from scratch
            i. Volcano Plots 
        c. **CELL CYCLE ANALYSIS**
            i. Evaluate the idea that cell cycle might influence expression of signatures 

This is the second part of our 3-part analysis of the Izar 2020 SS2 (Cohort 2) data. 

### STEP 1 LOAD SEURAT OBJECT

```{r init, message = FALSE, warning = FALSE}
# Load packages
source(here::here('packages.R'))

#Read in SS2 RDS object
SS2 = readRDS(file = "data/Izar_2020/Izar_2020_SS2.RDS")

# Read in gene lists
ccgenes = read_lines("data/gene_lists/regev_lab_cell_cycle_genes.txt")
s.genes <- ccgenes[1:43]
g2m.genes <- ccgenes[44:97]

#Read in hallmarks of interest
hallmark_names = read_lines("data/gene_lists/hallmarks.txt")
hallmark.list <- vector(mode = "list", length = length(hallmark_names) + 2)
names(hallmark.list) <- c(hallmark_names, "GO.OXPHOS", "KEGG.OXPHOS")
for(hm in hallmark_names){
  file <- read_lines(glue("data/gene_lists/hallmarks/{hm}_updated.txt"), skip = 1)
  hallmark.list[[hm]] <- file
}

hallmark.list[["GO.OXPHOS"]] <- read_lines("data/gene_lists/extra/GO.OXPHOS.txt", skip = 1)
hallmark.list[["KEGG.OXPHOS"]] <- read_lines("data/gene_lists/extra/KEGG.OXPHOS.txt", skip = 2)

```

### STEP 2 SUBSET AND SCALE SEURAT OBJECT FOR MALIGNANT CELLS AND FOR PATIENTS 8 & 9 ONLY 
```{r subset and scale Malignant, message=FALSE, warning = FALSE}
#subset
SS2Malignant <- subset(SS2, subset = (cell.type == "Malignant"))
SS2Malignant.8 <- subset(SS2Malignant, subset = (Patient == 8))
SS2Malignant.9 <- subset(SS2Malignant, subset = (Patient == 9))

#scale
SS2Malignant <- ScaleData(SS2Malignant, do.scale = TRUE, do.center = TRUE)
SS2Malignant.8 <- ScaleData(SS2Malignant.8, do.scale = TRUE, do.center = TRUE)
SS2Malignant.9 <- ScaleData(SS2Malignant.9, do.scale = TRUE, do.center = TRUE)
```

### STEP 3 SS2 MALIGNANT CELLS METRICS 
```{r malignant cells metrics, message=FALSE, warning=FALSE}
#set levels
SS2Malignant$sample = factor(SS2Malignant$sample, levels = trimws(sort(formatC(unique(SS2$sample)))))
SS2Malignant$treatment.status = factor(SS2Malignant$treatment.status, levels = c("Treatment-naïve","On-treatment"))
SS2Malignant.8$treatment.status = factor(SS2Malignant.8$treatment.status, levels = c("Treatment-naïve","On-treatment"))
SS2Malignant.9$treatment.status = factor(SS2Malignant.9$treatment.status, levels = c("Treatment-naïve","On-treatment"))

#numMalignantCellsPerSample
numMalignantCellsPerSample <- table(SS2Malignant$sample, SS2Malignant$cell.type, SS2Malignant$treatment.status) %>%
  as.data.frame() %>% 
  dplyr::rename(sample = Var1, celltype = Var2, treatment = Var3, numCells = Freq) %>% 
  filter(`celltype` == "Malignant")
ggplot(numMalignantCellsPerSample, aes(x = sample, y=numCells, fill = treatment)) + 
  geom_bar(stat = "identity") + 
  labs(title = "Number of Malignant Cells Per Sample in SS2 Data") 
ggsave("numMalignantCellsPerSample.png", path = "jesslyn_plots/SS2", width = 8, height = 8)

seurat_tab(SS2Malignant, col_var = "sample", row_var = "treatment.status", title = "Number of Malignant cells per Sample per treatment status")

seurat_tab(SS2Malignant, col_var = "Patient", row_var = "Time", title = "Number of Malignant cells per Sample per Patient")

#metrics for patient 8 and 9
numMalignantCellsPerSample.8 <- table(SS2Malignant.8$Patient, SS2Malignant.8$sample) %>%
  as.data.frame() %>% 
  dplyr::rename(Patient = Var1, sample = Var2, numCells = Freq)
p8 <- ggplot(numMalignantCellsPerSample.8, aes(x = Patient, y = numCells, fill = sample)) + 
    geom_bar(position = "fill", stat = "identity") + 
    labs(title = "Number of Malignant Cells Per Sample in Patient 8") + 
    geom_text(aes(label = c("Treatment-naïve", "On-treatment", "On-treatment")), position = "fill", hjust = 0.5, vjust = 2, color = "white") + 
    geom_text(aes(label = paste(numCells, "cells")), size =3, position = "fill", hjust = 0.5, vjust = 4, color = "white")

numMalignantCellsPerSample.9 <- table(SS2Malignant.9$Patient, SS2Malignant.9$sample) %>%
  as.data.frame() %>% 
  dplyr::rename(Patient = Var1, sample = Var2, numCells = Freq)
p9 <- ggplot(numMalignantCellsPerSample.9, aes(x = Patient, y = numCells, fill = sample)) + 
    geom_bar(position = "fill", stat = "identity") + 
    labs(title = "Number of Malignant Cells Per Sample in Patient 9") + 
    geom_text(aes(label = c("Treatment-naïve", "On-treatment", "On-treatment")), position = "fill", hjust = 0.5, vjust = 2, color = "white") + 
    geom_text(aes(label = paste(numCells, "cells")), size =3, position = "fill", hjust = 0.5, vjust = 4, color = "white")

p <- p8 + p9
ggsave(plot = p, filename = "SS2Malignant_P8P9_metrics.png", path = "jesslyn_plots/SS2", width = 10)
```

### STEP 4. DIMENSIONALITY REDUCTION 
```{r Dimensionality Reduction, message=FALSE, warning=FALSE}
#PCA data has already been scaled and centered above
SS2Malignant <- FindVariableFeatures(SS2Malignant, selection.method = "vst")
SS2Malignant.8 <- FindVariableFeatures(SS2Malignant.8, selection.method = "vst")
SS2Malignant.9 <- FindVariableFeatures(SS2Malignant.9, selection.method = "vst")

SS2Malignant <- RunPCA(SS2Malignant, features = VariableFeatures(SS2Malignant))
SS2Malignant.8 <- RunPCA(SS2Malignant.8, features = VariableFeatures(SS2Malignant.8))
SS2Malignant.9 <- RunPCA(SS2Malignant.9, features = VariableFeatures(SS2Malignant.9))

ElbowPlot(SS2Malignant)
ElbowPlot(SS2Malignant.8)
ElbowPlot(SS2Malignant.9)

#UMAP
SS2Malignant<- RunUMAP(SS2Malignant, dims=1:15)
SS2Malignant.8<- RunUMAP(SS2Malignant.8, dims=1:15)
SS2Malignant.9<- RunUMAP(SS2Malignant.9, dims=1:15)
```

### STEP 5. VISUALIZE METADATA WITH UMAP 
```{r PCA and UMAP, message=FALSE, warning=FALSE, fig.align='center', fig.width=20}
#cluster by patient/ sample? 
p1 <- UMAPPlot(SS2Malignant, group.by = "Patient") + labs(title = "SS2 Malignant Cells UMAP by Patient")
p2 <- UMAPPlot(SS2Malignant, group.by = "sample") + labs(title = "SS2 Malignant Cells UMAP by Sample")
p3 <- p1 + p2 
p3
ggsave(plot = p3, "SS2Malignant_All_UMAP.png", path = "jesslyn_plots/SS2", width = 10)

#cluster by treatment/sample within patient 8 and 9? 
p4 <- UMAPPlot(SS2Malignant.8, group.by = "treatment.status") + labs(title = "SS2 Patient 8 Malignant Cells by Treatment")
p5 <- UMAPPlot(SS2Malignant.8, group.by = "sample") + labs(title = "SS2 Patient 8 Malignant Cells by Sample")
p6 <- p4 + p5
p6
ggsave(plot = p6, "SS2Malignant.8_UMAP.png", path = "jesslyn_plots/SS2", width = 10)

p7 <- UMAPPlot(SS2Malignant.9, group.by = "treatment.status") + labs(title = "SS2 Patient 9 Malignant Cells by Treatment")
p8 <- UMAPPlot(SS2Malignant.9, group.by = "sample") + labs(title = "SS2 Patient 8 Malignant Cells by Sample")
p9 <- p7 + p8
p9
ggsave(plot = p9, "SS2Malignant.9_UMAP.png", path = "jesslyn_plots/SS2", width = 10)

```

### STEP 6. SAVE SEURAT OBJECT 
```{r save SDS}
saveRDS(SS2Malignant, file="data/Izar_2020/jesslyn_SS2Malignant_processed.RDS")
saveRDS(SS2Malignant.8, file="data/Izar_2020/jesslyn_SS2Malignant8_processed.RDS")
saveRDS(SS2Malignant.9, file="data/Izar_2020/jesslyn_SS2Malignant9_processed.RDS")
```
