---
title: "Izar 2020 10X (Cohort 1) Metadata Plots"
author: "Jesslyn Goh and Mike Cuoco"
date: "7/19/2020"
editor_options: 
  chunk_output_type: console
output:
 html_document:
  code_folding: hide
  toc_float: TRUE 
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = here::here())
knitr::opts_chunk$set(echo = TRUE)
options(width=120)
```

# IZAR 2020 10X (COHORT 1) DATA ANALYSIS - PART 2

### OVERVIEW
* This is the second part of our 3-part analysis of the Izar 2020 10X (Cohort 1) data. 
* The 10X data consists of **CD45+ depleted 8 ascites samples from 6 patients** 
    - Includes both **malignant** and **nonmalignant** populations: Provides a broad view of the diverse cell types in the ascites ecosystem
    - We are only interested in the **malignant** ascites population

* We split our 10X analysis into three parts: 
    1. Loading and Processing Data
    2. **Visualizing Clusters (PCA and UMAP)**
        a. Load in processed 10X and 10X Malignant Seurat Objects 
        b. For each Seurat Object, examine how cells separate by their metadata identities by visualizing with PCA / UMAP: 
            i. Patient / Sample ID 
            ii. Treatment Status 
            iii. Cell Cycle Phase
    3. DE Analysis 

### STEP 2A LOAD AND SUBSET SEURAT OBJECT

```{r init, message = FALSE, warning = FALSE}
# Load packages
source(here::here('packages.R'))

#Read in PDX RDS object
cohort1 = readRDS("data/Izar_2020/jesslyn_cohort1_processed.RDS")
cohort1Malignant = readRDS("data/Izar_2020/jesslyn_cohort1Malignant_processed.RDS")

# Read in gene lists
ccgenes = read_lines("data/gene_lists/regev_lab_cell_cycle_genes.txt")
s.genes <- ccgenes[1:43]
g2m.genes <- ccgenes[44:97]

#Read in hallmarks of interest
hallmark_names = read_lines("data/gene_lists/hallmarks.txt")
hallmark.list <- vector(mode = "list", length = length(hallmark_names))
names(hallmark.list) <- hallmark_names
for(hm in hallmark_names){
  file <- read_lines(glue("data/gene_lists/hallmarks/{hm}_updated.txt"), skip = 1)
  hallmark.list[[hm]] <- file
}

if(!dir.exists("jesslyn_plots/cohort1/All")){dir.create("jesslyn_plots/cohort1/All")}
if(!dir.exists("jesslyn_plots/cohort1/Malignant")){dir.create("jesslyn_plots/cohort1/Malignant")}
```

### STEP 2B VISUALIZE METADATA IDENTITIES FOR ALL CELLS 

```{r metadata pca - all cells, message=FALSE, warning=FALSE, fig.align='center', fig.width=20}
#plot PCA - shows differential expresison between different cohorts
pca1 <- PCAPlot(cohort1) + labs(title = "10X All Cells PCA")

#separate by treatment? 
pca2 <- PCAPlot(cohort1, group.by="treatment.status") + labs(title = "10X All Cells PCA by Treatment Status")

#separate by cellcycle phase?
pca3 <- PCAPlot(cohort1, group.by="Phase") + labs(title = "10X All Cells PCA by Cell Cycle Phase")

#separate by patient?
pca4 <- PCAPlot(cohort1, group.by="patient") + labs(title = "10X All Cells PCA by Patient")

pca <- pca1 + pca2 + pca3 + pca4
ggsave(plot = pca, filename = "10X_All_PCA", path = "jesslyn_plots/cohort1/All")
```

```{r metadata umap - all cells, message=FALSE, warning = FALSE, fig.align='center', fig.width=20}

umap1 <- UMAPPlot(cohort1, group.by="clst") + labs(title = "10X All Cells UMAP By Cluster")

umap2 <- UMAPPlot(cohort1, group.by='sample.ID') + labs(title = "10X All Cells UMAP By Sample")

umap3 <- UMAPPlot(cohort1, group.by = 'treatment.status') + labs(title = "10X All Cells UMAP By Treatment Status")

umap4 <- UMAPPlot(cohort1, group.by='Phase') + labs(title = "10X All Cells UMAP By Cell Cycle Phase")

umap <- umap1 + umap2 + umap3 + umap4
ggsave(plot = umap, filename = "10X_All_UMAP", path = "jesslyn_plots/cohort1/All")
```

### STEP 2C VISUALIZE METADATA IDENTITIES FOR MALIGNANT CELLS 

```{r metadata umap - malignant cells, message = FALSE, warning = FALSE, fig.align = 'center', fig.width=20}

#umap by sample
malumap1 <- UMAPPlot(cohort1Malignant, group.by="sample.ID") + labs(title = "10X Malignant Cells UMAP by Sample")

#umap by treatment
malumap2 <- UMAPPlot(cohort1Malignant, group.by="treatment.status") + labs(title = "10X Malignant Cells UMAP by Treatment Status")

#umap by cell cycle phase
malumap3 <- UMAPPlot(cohort1Malignant, group.by="Phase") + labs(title = "10X Malignant Cells UMAP by Cell Cycle Phase")

malumap <- malumap1 + malumap2 + malumap3
ggsave(plot = malumap, filename = "10X_Mal_UMAP", path = "jesslyn_plots/cohort1_Malignant")
```

