---
title: "Izar 2020 10X (Cohort 1) Load Data"
author: "Jesslyn Goh and Mike Cuoco"
date: "7/15/2020"
editor_options: 
  chunk_output_type: console
output:
 html_document:
  code_folding: hide
  toc_float: TRUE 
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = here::here())
knitr::opts_chunk$set(echo = TRUE)
options(width=120)
```

# IZAR 2020 10X (COHORT 1) DATA ANALYSIS - PART 1

### OVERVIEW
* This is the first part of our 3-part analysis of the Izar 2020 10X (Cohort 1) data. 
* The 10X data consists of **CD45+ depleted 8 ascites samples from 6 patients** 
    - Includes both **malignant** and **nonmalignant** populations: Provides a broad view of the diverse cell types in the ascites ecosystem
    - We are only interested in the **malignant** ascites population

* We split our 10X analysis into three parts: 
    1. **Loading and Processing Data**
        a. Load in Izar2020 10X data and create Seurat Objects
        b. Assign metadata (Patient, Sample IDs, treatment status)
        c. Subset the Seurat Object for Malignant Cells 
        d. Score Cells (Cell cycle and Each hallmark using **AddModuleScore**)
        e. Evaluate Patient and Treatment Metrics
        f. Scale and Center Cells 
        g. Dimensional Reduction (PCA)
        h. Save Seurat Obejcts
    2. Visualizing Clusters (UMAP)
    3. DE Analysis 

### STEP 1A-C LOAD AND SUBSET SEURAT OBJECT

```{r init, message = FALSE, warning = FALSE}
# Load packages
source(here::here('packages.R'))

#Read in PDX RDS object
cohort1 = readRDS("data/Izar_2020/Izar_2020_10x.RDS")

#filter out clusters -1, 19-21
cohort1 <- subset(cohort1, subset = clst > -1 & clst < 19)

#set levels 
treatment_levels <- c("Treatment-naïve", "After 1 cycle of chemotherapy", "On-treatment")
cohort1$treatment.status = factor(cohort1$treatment.status, levels = treatment_levels)

# Read in gene lists
ccgenes = read_lines("data/gene_lists/regev_lab_cell_cycle_genes.txt")
s.genes <- ccgenes[1:43]
g2m.genes <- ccgenes[44:97]

#Read in hallmarks of interest
hallmark_names = read_lines("data/gene_lists/hallmarks.txt")
hallmark.list <- vector(mode = "list", length = length(hallmark_names))
names(hallmark.list) <- hallmark_names
for(hm in hallmark_names){
  file <- read_lines(glue("data/gene_lists/hallmarks/{hm}_updated.txt"), skip = 1)
  hallmark.list[[hm]] <- file
}

#Subset Malignant Cells 
cohort1Malignant <- subset(cohort1, subset = (cell.type == "Malignant"))

if(!dir.exists("jesslyn_plots/cohort1")){dir.create("jesslyn_plots/cohort1")}
```

### STEP 1D. SCORE CELLS

```{r score cells, message = FALSE, warning = FALSE}
#cell cycle 
cohort1 <- CellCycleScoring(cohort1, g2m.features = g2m.genes, s.features = s.genes)
cohort1Malignant <- CellCycleScoring(cohort1Malignant, g2m.features = g2m.genes, s.features = s.genes)

#hallmark score
cohort1 <- AddModuleScore(cohort1, features = hallmark.list, name = names(hallmark.list), nbin = 25, search = T) 
cohort1Malignant <- AddModuleScore(cohort1Malignant, features = hallmark.list, name = names(hallmark.list), nbin = 25, search = T) 

```

### STEP 1E. EVALUATE PATIENT AND TREATMENT METRICS

```{r All cells Metrics, message = FALSE, warning = FALSE}

#PATIENT METRICS
numCellsPerPatient = table(cohort1@meta.data$patient) %>% as.data.frame()

numCellsPerTypePerPatient = table(cohort1$patient, cohort1$cell.type) %>% 
  as.data.frame() %>% 
  rename(patient = Var1, celltype = Var2, numCells = Freq)
ggplot(numCellsPerTypePerPatient, aes(x=celltype, fill=patient, y=numCells)) + geom_bar(position="fill", stat="identity")

numCellsPerCellTypePerSample = table(cohort1$sample.ID, cohort1$cell.type) %>% 
  as.data.frame() %>% 
  rename(sample = Var1, celltype = Var2, numCells = Freq)
ggplot(numCellsPerCellTypePerSample, aes(x=celltype, fill=sample, y=numCells)) + geom_bar(position="fill", stat="identity")

#create table using Mike's function
seurat_tab(cohort1, col_var = "sample.ID", row_var = "cell.type", group_var = "treatment.status", title = "Number of cells per celltype per sample per treatment status") %>% 
    gtsave("10X_numCellsPerCelltypePerSamplePerTreatment.png", path = "jesslyn_plots/cohort1")

#TREATMENT METRICS 
numPatientsPerTreatment = table(cohort1$patient, cohort1$treatment.status)

numCellsPerTreatmentPerCellType = table(cohort1$cell.type, cohort1$treatment.status) %>% 
  as.data.frame() %>% 
  rename(celltype = Var1, treatment = Var2, numCells = Freq)
ggplot(numCellsPerTreatmentPerCellType, aes(x=celltype, fill=treatment, y=numCells)) + geom_bar(position="fill", stat="identity") + labs(fill = "treatment")

numCellsPerTreatmentPerSample = table(cohort1$sample.ID, cohort1$treatment.status) %>% 
  as.data.frame() %>% 
  rename(sample = Var1, treatment=Var2, numCells = Freq) 
ggplot(numCellsPerTreatmentPerSample, aes(x=treatment, fill=sample, y=numCells)) + geom_bar(position="fill", stat="identity") + xlab("treatment")
```

```{r Malignant cells Metrics, message = FALSE, warning = FALSE}

numMalCellsPerSample <- table(cohort1Malignant$sample.ID, cohort1Malignant$cell.type) %>% 
  as.data.frame() %>% 
  rename(sample = Var1, numCells=Freq, celltype = Var2)
ggplot(numMalCellsPerSample, aes(x=celltype, fill=sample, y=numCells)) + geom_bar(stat="identity", position="fill") + labs(title = "Number of Malignant Cells Per Sample")

numMalCellsPerSamplePerTreatment<-table(cohort1Malignant$sample.ID, cohort1Malignant$treatment.status) %>% 
  as.data.frame() %>% 
  rename(sample = Var1, treatment = Var2, numCells = Freq)
ggplot(numMalCellsPerSamplePerTreatment, aes(x=treatment, fill=sample, y=numCells)) + geom_bar(stat="identity", position="fill") + labs(title = "Number of Malignant Cells Per Sample Per Treatment", x = "treatment")

numMalCellsperTreatment <- table(cohort1Malignant$treatment.status, cohort1Malignant$cell.type) %>% as.data.frame() %>% rename(celltype = Var2, treatment = Var1, numCells = Freq)
ggplot(numMalCellsperTreatment, aes(x=celltype, y=numCells, fill=treatment)) + geom_bar(stat = "identity", position = "fill")

#CELL CYCLE METRICS
table(cohort1$cell.type, cohort1$Phase) %>% 
  as.data.frame() %>% 
  rename(Celltype = Var1, Phase = Var2, numCells = Freq) %>% 
  ggplot(aes(x=Celltype, y=numCells, fill = Phase)) + geom_bar(position="fill", stat="identity")

numMalCellsPerTreatmentPerPhase = table(cohort1Malignant$Phase, cohort1Malignant$treatment.status) %>% 
  as.data.frame() %>% 
  rename(Phase = Var1, Treatment = Var2, numCells = Freq)
ggplot(numMalCellsPerTreatmentPerPhase, aes(x=Treatment, y=numCells, fill=Phase)) + geom_bar(position="fill", stat="identity")

```

### STEP 1F. SCALING AND CENTERING FOR PCA 
Prepare data for dimensionality reduction: 
    * FindVariableFeatures
    * ScaleData
    
```{r ScaleData and FindVariableFeatures, message = FALSE, warning = FALSE}
#identify highly variable genes 
cohort1 <- FindVariableFeatures(cohort1, selection.method = "vst")
cohort1Malignant <- FindVariableFeatures(cohort1Malignant, selection.method = "vst")

#scale data 
cohort1 <- ScaleData(cohort1, do.scale = TRUE, do.center = TRUE) 
cohort1Malignant <- ScaleData(cohort1Malignant, do.scale = TRUE, do.center = TRUE)
```

### STEP 1G. DIMENSIONALITY REDUCTION (PCA AND UMAP)

```{r PCA and UMAP}

#run PCA
cohort1 <- RunPCA(cohort1, features = VariableFeatures(cohort1))
cohort1Malignant <- RunPCA(cohort1Malignant, features = VariableFeatures(cohort1Malignant))

#determine dimensionality of data (how many pcs to include)
ElbowPlot(cohort1)
ElbowPlot(cohort1Malignant)
#graph shows probably pc1 to pc15 are important

#UMAP 
cohort1 <- RunUMAP(cohort1, dims=1:15)
cohort1Malignant <- RunUMAP(cohort1Malignant, dims=1:15)

```

### STEP 1H. CLUSTERING 

```{r run clustering}
#find neighbors first
cohort1 <- FindNeighbors(cohort1)
cohort1Malignant <- FindNeighbors(cohort1Malignant)

#find clusters
cohort1 <- FindClusters(cohort1)
cohort1Malignant <- FindClusters(cohort1Malignant)

```

```{r save SDS}
saveRDS(cohort1, file="data/Izar_2020/jesslyn_cohort1_processed.RDS")
saveRDS(cohort1Malignant, file="data/Izar_2020/jesslyn_cohort1Malignant_processed.RDS")
```
