---
title: "Izar 2020 SS2 (Cohort 2) DE Analysis"
author: "Jesslyn Goh and Mike Cuoco"
date: "7/20/2020"
editor_options: 
  chunk_output_type: console
output:
 html_document:
  code_folding: hide
  toc_float: TRUE 
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = here::here())
knitr::opts_chunk$set(echo = TRUE)
options(width=120)
```

# IZAR 2020 SS2 (COHORT 2) DATA DIFFERANTIAL EXPRESISON ANALYSIS 

### OVERVIEW
* The SS2 data from Izar2020 consists of cells from 14 ascites samples from 6 individuals: 
    - selected for malignant EPCAM+CD24+ cells by fluorophore-activated cell sorting into 96-well plates
    - Includes both **malignant** and **nonmalignant** populations but differs from 10X data - examines malignant ascites more closely. 
    - We are only interested in the **malignant** ascites population, specifically samples from **patient 8 and 9** because they are the only patients that have samples from different treatment statuses. 
    
* In our 5-part analysis of the Izar 2020 SS2 data, we are interested in identifying differentially expressed genes and hallmark genesets between treatment statuses **within** patients 8 and 9 
* We split our SS2 analysis into three parts: 
    1. **Load Data and Create SS2 Seurat Object**
        a. The code to this part of our analysis is in the **read_Izar_2020.R** file in the **code** folder. During this part of our analysis we: 
            i. Load in SS2 count matrix and Create Seurat Object 
            ii. Assign Metadata identities including: 
                - Patient ID
                - Time
                - Sample ID
                - Treatment Status 
            iii. Score cells for **cell cycle** and **hallmark genesets** 
                - **Note**: It does not matter whether we call AddModuleScore before or after subsetting and scaling each model because AddModuleScore uses the **data** slot. 
            iv. Save Seurat Object
          
    2. **Process Data and Exploratory Data Analysis**
        a. The code to this part of our analysis can be found in the **02_Izar2020_SS2_Load_Plots** file in the **old/edited** folder. During this part of our analysis we: 
            i. Load in SS2 Seurat Object from Part 1 and **subset by patients 8 and 9**. Continue analysis separately for each model. 
            ii. Scale and FindVariableFeatures (prepares data for dimensionality reduction)
            iii. Dimensionality Reduction (PCA + UMAP)
            vi. Save Seurat Objects 
      
    3. **Exploratory Data Analysis**
        a. The code to this part of our analysis can be found in the **02.0_Izar2020_SS2_Exploratory Analysis** file in the **analysis** folder. During this part of our analysis we: 
            i. Load in SS2 Seurat Object from Part 2. Analyze separately for patients 8 and 9
            ii. Compute summary metrics for SS2 data such as: 
                - Number of cells per patient per treatment 
                - Number of cells per treatment per cell cycle phase 
            iii. Visualize how cells separate based on metadata identities via UMAP and PCA
                - Intermodel heterogeneity: How do cells separate by sample? 
                - Intramodel heterogeneity: 
                    * How do cells separate by treament status / sample? 
                    * How do cells separate by cell cycle phase?
            
    4. **DE Analysis**
        a. **TYPE #1 DE ANALYSIS**: Visualizing and Quantifying Differentially Expression on **Predefined** GO Genesets
            i. Violin Plots and UMAP 
            ii. Gene Set Enrichment Analysis (GSEA)
        b. **TYPE #2 DE ANALYSIS**: Finding DE Genes from scratch
            i. Volcano Plots 
          
    5. **CELL CYCLE ANALYSIS**
        a. **TYPE #1 CELL CYCLE ANALYSIS**: Examine correlation between treatment condition and cell cycle phase
        b. Evaluate the idea that cell cycle might influence expression of signatures 

This is the third part of our 5-part analysis of the Izar 2020 SS2 (Cohort 2) data. 

### STEP 1 LOAD SEURAT OBJECT

```{r init, message = FALSE, warning = FALSE}
# Load packages
source(here::here('packages.R'))

#Read in SS2 RDS object
SS2Malignant = readRDS(file = "data/Izar_2020/jesslyn_SS2Malignant_processed.RDS")
SS2Malignant.8 = readRDS(file = "data/Izar_2020/jesslyn_SS2Malignant8_processed.RDS")
SS2Malignant.9 = readRDS(file = "data/Izar_2020/jesslyn_SS2Malignant9_processed.RDS")

#Read in hallmarks of interest
g0g1.genes <- read_lines("data/gene_lists/cellcycle/G0G1.txt", skip = 2)
g1S.DNAdamage.genes <- read_lines("data/gene_lists/cellcycle/G1SDNAdamage.txt", skip = 2)
g1S.transcription.genes <- read_lines("data/gene_lists/cellcycle/G1Stranscription.txt", skip = 2)
g2m.checkpoint.genes <- read_lines("data/gene_lists/cellcycle/G2Mcheckpoint.txt", skip = 2)
g2m.DNAdamage.genes <- read_lines("data/gene_lists/cellcycle/G2MDNAdamage.txt", skip = 2)
g2m.replicationCP.genes <- read_lines("data/gene_lists/cellcycle/G2MreplicationCheckpoint.txt", skip = 2)

hallmark_names = read_lines("data/gene_lists/hallmarks.txt")
hallmark.list <- vector(mode = "list", length = length(hallmark_names) + 9)
names(hallmark.list) <- c(hallmark_names, "GO.OXPHOS", "KEGG.OXPHOS", "UNUPDATED.OXPHOS", 
                          "UNUPDATED.UPR", "G0G1", "G1S.DNAdamage", "G1S.transcription", "G2M.checkpoint", 
                          "G2M.DNAdamage")
for(hm in hallmark_names){
  file <- read_lines(glue("data/gene_lists/hallmarks/{hm}_updated.txt"), skip = 1)
  hallmark.list[[hm]] <- file
}

hallmark.list[["GO.OXPHOS"]] <- read_lines("data/gene_lists/extra/GO.OXPHOS.txt", skip = 1)
hallmark.list[["KEGG.OXPHOS"]] <- read_lines("data/gene_lists/extra/KEGG.OXPHOS.txt", skip = 2)
hallmark.list[["UNUPDATED.OXPHOS"]] <- read_lines("data/gene_lists/oxphos.txt", skip =1)
hallmark.list[["UNUPDATED.UPR"]] <- read_lines("data/gene_lists/upr.txt", skip =1)
hallmark.list[["G0G1"]] <- g0g1.genes
hallmark.list[["G1S.DNAdamage"]] <- g1S.DNAdamage.genes
hallmark.list[["G1S.transcription"]] <- g1S.transcription.genes
hallmark.list[["G2M.checkpoint"]] <- g2m.checkpoint.genes
hallmark.list[["G2M.DNAdamage"]] <- g2m.DNAdamage.genes

#center module and cell cycle scores and reassign to the metadata of each Seurat object
hm.names <- names(SS2Malignant@meta.data)[14:58]

for(i in hm.names){
    SS2Malignant.hm.centered <- scale(SS2Malignant[[i]], center = TRUE, scale = FALSE)
    SS2Malignant <- AddMetaData(SS2Malignant, SS2Malignant.hm.centered, col.name = glue("{i}.centered"))
    
    SS2Malignant8.hm.centered <- scale(SS2Malignant.8[[i]], center = TRUE, scale = FALSE)
    SS2Malignant.8 <- AddMetaData(SS2Malignant.8, SS2Malignant8.hm.centered, col.name = glue("{i}.centered"))
    
    SS2Malignant9.hm.centered <- scale(SS2Malignant.9[[i]], center = TRUE, scale = FALSE)
    SS2Malignant.9 <- AddMetaData(SS2Malignant.9, SS2Malignant9.hm.centered, col.name = glue("{i}.centered"))
}
```

### STEP 2 DETERMINING OXPHOS GENESET 
* Before doing any DE Analysis on the genesets, we investigate which of the three OXPHOS genesets we found is best for our data. Although we've investigated this for our PDX data, the geneset that is more suitable for SS2 may be different. The **OXPHOS** genesets that we test are listed as follows:
    - Unupdated Version of HALLMARK_OXIDATIVE_PHOSPHORYLATION (200 genes) https://www.gsea-msigdb.org/gsea/msigdb/cards/HALLMARK_OXIDATIVE_PHOSPHORYLATION
    - Updated Version of HALLMARK_OXIDATIVE_PHOSPHORYLATION (200 genes) https://www.gsea-msigdb.org/gsea/msigdb/cards/HALLMARK_OXIDATIVE_PHOSPHORYLATION
    - GO_OXIDATIVE_PHOSPHORYLATION (144 genes) https://www.gsea-msigdb.org/gsea/msigdb/cards/GO_OXIDATIVE_PHOSPHORYLATION 
    - KEGG_OXIDATIVE_PHOSPHORYLATION (131 genes) https://www.gsea-msigdb.org/gsea/msigdb/cards/KEGG_OXIDATIVE_PHOSPHORYLATION

* We determine which geneset to use by asking these questions: 
    1. How many and which genes are not found in the SS2 Seurat Object for each geneset?
    2. What makes the genesets different from each other? Which genes in each geneset are the most DE or have the higest logFC? Are they the same?
        a. Volcano Plot: Call FindMarkers on each model, plot onto Volcano Plot, label cells by each geneset 
    3. Which geneset gives us the most statistically significant results? 
        a. VlnPlot: Plot hallmark scores for each geneset group by treatment status for each model 
        b. Label the VlnPlot with the number of genes found in SS2 Seurat Object for each OXPHOS geneset 
  
**ANSWERING QUESTION #1: How many and which genes are not found in the SS2 Seurat Object for each geneset?**

```{r oxphosMetrics, message=FALSE, warning=FALSE, fig.align = 'center'}

hm.length.df <- data.frame(
  "UNUPDATED.HM.OXPHOS" = length(hallmark.list[["UNUPDATED.OXPHOS"]]),
  "HALLMARK.OXPHOS" = length(hallmark.list[["HALLMARK_OXIDATIVE_PHOSPHORYLATION"]]), 
  "GO.OXPHOS" = length(hallmark.list[["GO.OXPHOS"]]), 
  "KEGG.OXPHOS" = length(hallmark.list[["KEGG.OXPHOS"]])
)

Found.df <- data.frame(
  "UNUPDATED.HM.OXPHOS" = sum((hallmark.list[["UNUPDATED.OXPHOS"]] %in% rownames(SS2Malignant))),
  "HALLMARK.OXPHOS" = sum((hallmark.list[["HALLMARK_OXIDATIVE_PHOSPHORYLATION"]] %in% rownames(SS2Malignant))), 
  "GO.OXPHOS" = sum((hallmark.list[["GO.OXPHOS"]] %in% rownames(SS2Malignant))), 
  "KEGG.OXPHOS" = sum((hallmark.list[["KEGG.OXPHOS"]] %in% rownames(SS2Malignant)))
)

PFound.df <- data.frame(
  "UNUPDATED.HM.OXPHOS" = (sum((hallmark.list[["UNUPDATED.OXPHOS"]] %in% rownames(SS2Malignant)))/length(hallmark.list[["UNUPDATED.OXPHOS"]]))*100,
  "HALLMARK.OXPHOS" = (sum((hallmark.list[["HALLMARK_OXIDATIVE_PHOSPHORYLATION"]] %in% rownames(SS2Malignant)))/length(hallmark.list[["HALLMARK_OXIDATIVE_PHOSPHORYLATION"]])) * 100, 
  "GO.OXPHOS" = (sum((hallmark.list[["GO.OXPHOS"]] %in% rownames(SS2Malignant)))/length(hallmark.list[["GO.OXPHOS"]]))*100, 
  "KEGG.OXPHOS" = (sum((hallmark.list[["KEGG.OXPHOS"]] %in% rownames(SS2Malignant)))/length(hallmark.list[["KEGG.OXPHOS"]]))*100
)

NA.df <- data.frame(
  "UNUPDATED.HM.OXPHOS" = sum(!(hallmark.list[["UNUPDATED.OXPHOS"]] %in% rownames(SS2Malignant))),
  "HALLMARK.OXPHOS" = sum(!(hallmark.list[["HALLMARK_OXIDATIVE_PHOSPHORYLATION"]] %in% rownames(SS2Malignant))), 
  "GO.OXPHOS" = sum(!(hallmark.list[["GO.OXPHOS"]] %in% rownames(SS2Malignant))), 
  "KEGG.OXPHOS" = sum(!(hallmark.list[["KEGG.OXPHOS"]] %in% rownames(SS2Malignant)))
)

all.df <- rbind(hm.length.df, Found.df, PFound.df, NA.df)
rownames(all.df) <- c("NumGenes", "Found", "%Found", "Not Found")
all.df[,"GO.OXPHOS"] <- round(all.df[,"GO.OXPHOS"])
all.df[,"KEGG.OXPHOS"] <- round(all.df[,"KEGG.OXPHOS"])
all.df

# IDENTIFY GENES THAT ARE NOT FOUND -----------
NA.genes.df <- vector("list", length = 4)
names <- c("UNUPDATED.OXPHOS", "HALLMARK_OXIDATIVE_PHOSPHORYLATION", "GO.OXPHOS", "KEGG.OXPHOS")
names(NA.genes.df) <- names
for(i in names){
  NA.genes.df[[i]] <- (hallmark.list[[i]])[which(!(hallmark.list[[i]] %in% rownames(SS2Malignant)))]
}
```
* **OBSERVATIONS**
    - Similar to what we observed in the PDX data, the **UNUPDATED.OXPHOS** genelist actually has the most genes that are found in the SS2 object. 
    - Since the **UNUPDATED.OXPHOS** genelist works better than the updated version (HALLMARK_OXIDATIVE_PHOSPHORYLATION), we will continue our comparison of genesets using the UNUPDATED version. 
* We now wonder: 
    - Which OXPHOS genes are the most DE (or have the highest logFC) if we call FindMarkers? 
    - Which geneset(s) do the most DE OXPHOS genes belong to? 

**ANSWERING QUESTION 2: Which genes in each geneset are the most DE? Are they the same?**

* Used the **wilcoxon rank sum test** for FindMarkers 

```{r oxphosVolcano, message=FALSE, warning=FALSE, fig.align='center', fig.width=19, fig.height=8}
SS2s <- c(SS2Malignant.8, SS2Malignant.9)
SS2.names <- c("SS2 Patient 8", "SS2 Patient 9")
oxphos.hm <- c("UNUPDATED.OXPHOS", "GO.OXPHOS", "KEGG.OXPHOS")
SS2.hm.plots <- vector("list", length = 3)
names(SS2.hm.plots) <- SS2.names

SS2.markers <- vector("list", length = 3)
names(SS2.markers) <- SS2.names

SS2.markers[["SS2 Patient 8"]] <- FindMarkers(SS2Malignant.8, group.by = "sample", ident.1 = "8.1", ident.2 = "8.0", test.use = "wilcox", logfc.threshold = 0) 
SS2.markers[["SS2 Patient 9"]]  <- FindMarkers(SS2Malignant.9, group.by = "sample", ident.1 = "9.1", ident.2 = "9.0", test.use = "wilcox", logfc.threshold = 0)

for(i in 1:length(SS2s)){
    SS2 <- SS2.names[[i]]
    DF.hm.plot <- vector("list", length = 3)
    names(DF.hm.plot) <- oxphos.hm
    marker <- SS2.markers[[SS2]]
    
  for(oxphos in oxphos.hm){
    
    avgLFC <- rownames(marker[which(abs(marker$avg_logFC) > 0.5),])
    keyvals <- ifelse(!(rownames(marker) %in% hallmark.list[[oxphos]]), 'black', 
                      ifelse(abs(marker$avg_logFC) > 0.5, 'red', 'grey'))
    names(keyvals)[keyvals == 'red'] <- 'OXPHOS & logFC'
    names(keyvals)[keyvals == 'grey'] <- 'OXPHOS x logFC'
    names(keyvals)[keyvals == 'black'] <- 'Not OXPHOS'
    found = length(avgLFC[which(avgLFC %in% hallmark.list[[oxphos]])])
    
    p <- EnhancedVolcano(marker, 
           lab = rownames(marker),
           selectLab = avgLFC[which(avgLFC %in% hallmark.list[[oxphos]])],
           labCol = "red",
           x='avg_logFC', y='p_val_adj', pCutoff = 0.05, 
           FCcutoff = 0.5, 
           colCustom = keyvals,
           pointSize = c(ifelse(rownames(marker) %in% hallmark.list[[oxphos]], 2.5,1)), 
           drawConnectors = TRUE,
           boxedLabels = TRUE,
           labvjust = 1,
           title= glue("{SS2} p.1 vs. p.0"), subtitle= "LogFC cutoff: 0.5, p cutoff: 0.05",
           caption = glue("{oxphos}: {found} high LFC oxphos genes found")
           )
    DF.hm.plot[[oxphos]] <- p
  }
  
  SS2.hm.plots[[SS2]] <- DF.hm.plot[["UNUPDATED.OXPHOS"]] + DF.hm.plot[["GO.OXPHOS"]] + DF.hm.plot[["KEGG.OXPHOS"]]
}

SS2.hm.plots[["SS2 Patient 8"]]
SS2.hm.plots[["SS2 Patient 9"]]

# number of DE oxphos genes found in each geneset within each model -----------------
P8.de.df <- data.frame(
  "UNUPDATED.HM.OXPHOS" = sum(rownames(SS2.markers[["SS2 Patient 8"]][which(abs(SS2.markers[["SS2 Patient 8"]]$avg_logFC) > 0.5),]) %in% hallmark.list[["UNUPDATED.OXPHOS"]]),
  "GO.OXPHOS" = sum(rownames(SS2.markers[["SS2 Patient 8"]][which(abs(SS2.markers[["SS2 Patient 8"]]$avg_logFC) > 0.5),]) %in% hallmark.list[["GO.OXPHOS"]]),
  "KEGG.OXPHOS" = sum(rownames(SS2.markers[["SS2 Patient 8"]][which(abs(SS2.markers[["SS2 Patient 8"]]$avg_logFC) > 0.5),]) %in% hallmark.list[["KEGG.OXPHOS"]])
)

P9.de.df <- data.frame(
  "UNUPDATED.HM.OXPHOS" = sum(rownames(SS2.markers[["SS2 Patient 9"]][which(abs(SS2.markers[["SS2 Patient 9"]]$avg_logFC) > 0.5),]) %in% hallmark.list[["UNUPDATED.OXPHOS"]]),
  "GO.OXPHOS" = sum(rownames(SS2.markers[["SS2 Patient 9"]][which(abs(SS2.markers[["SS2 Patient 9"]]$avg_logFC) > 0.5),]) %in% hallmark.list[["GO.OXPHOS"]]),
  "KEGG.OXPHOS" = sum(rownames(SS2.markers[["SS2 Patient 9"]][which(abs(SS2.markers[["SS2 Patient 9"]]$avg_logFC) > 0.5),]) %in% hallmark.list[["KEGG.OXPHOS"]])
)

all.de.df <- rbind(P8.de.df, P9.de.df)
rownames(all.de.df) <- c("P8.1vs8.0", "P9.1vs9.0")
all.de.df

percent.logFC <- data.frame(
  "UNUPDATED.HM.OXPHOS" = round(all.de.df[, "UNUPDATED.HM.OXPHOS"]/all.df["Found", "UNUPDATED.HM.OXPHOS"]*100, 2), 
  "GO.OXPHOS" = round(all.de.df[, "GO.OXPHOS"]/all.df["Found", "GO.OXPHOS"]*100, 2), 
  "KEGG.OXPHOS" = round(all.de.df[, "KEGG.OXPHOS"]/all.df["Found", "KEGG.OXPHOS"]*100, 2)
)
rownames(percent.logFC) <- c("P8 %Found highLFC", "P9 %Found highLFC")
rbind(all.df[,names(all.df) != "HALLMARK.OXPHOS"], percent.logFC)
```

* **OBSERVATIONS** 
    - **UNUPDATED.OXPHOS** geneset consistently has the most number OXPHOS genes with **high logFC** (logFC > 0.5, regardless of padj) relative to the other two genesets for each model comparison between MRD and vehicle. However, **GO.OXPHOS** actually has the highest percentage of OXPHOS genes with high logFC out of the OXPHOS genes Found. 
    - Although it seems that there are a few overlaps of OXPHOS genes found, it is hard to visualize which ones overlap, and whether the ones that overlap are among the top DE genes (regardless of padj). We therefore create tables for each model, extract 5 genes with the highest logFC from each geneset and compare if they're the same. 
    
```{r oxphosGSDE, message=FALSE, warning=FALSE}

p8.top5 <- SS2.markers[["SS2 Patient 8"]] %>% arrange(-abs(avg_logFC))
p9.top5 <- SS2.markers[["SS2 Patient 9"]] %>% arrange(-abs(avg_logFC))

p8.gs.df <- data.frame(
  "UNUPDATED.OXPHOS" = head(rownames(p8.top5)[which(rownames(p8.top5) %in% hallmark.list[["UNUPDATED.OXPHOS"]])], 5), 
  "UNUPDATED.OXPHOS" = select(p8.top5[head(rownames(p8.top5)[which(rownames(p8.top5) %in% hallmark.list[["UNUPDATED.OXPHOS"]])], 5),], avg_logFC),
  "GO.OXPHOS" = head(rownames(p8.top5)[which(rownames(p8.top5) %in% hallmark.list[["GO.OXPHOS"]])], 5),
  "GO.OXPHOS" = select(p8.top5[head(rownames(p8.top5)[which(rownames(p8.top5) %in% hallmark.list[["GO.OXPHOS"]])], 5),], avg_logFC),
  "KEGG.OXPHOS" = head(rownames(p8.top5)[which(rownames(p8.top5) %in% hallmark.list[["KEGG.OXPHOS"]])], 5), 
  "KEGG.OXPHOS" = select(p8.top5[head(rownames(p8.top5)[which(rownames(p8.top5) %in% hallmark.list[["KEGG.OXPHOS"]])], 5),], avg_logFC)
)
rownames(p8.gs.df) <- seq(from = 1, length = nrow(p8.gs.df))

p9.gs.df <- data.frame(
  "UNUPDATED.OXPHOS" = head(rownames(p9.top5)[which(rownames(p9.top5) %in% hallmark.list[["UNUPDATED.OXPHOS"]])], 5), 
  "UNUPDATED.OXPHOS" = select(p9.top5[head(rownames(p9.top5)[which(rownames(p9.top5) %in% hallmark.list[["UNUPDATED.OXPHOS"]])], 5),], avg_logFC),
  "GO.OXPHOS" = head(rownames(p9.top5)[which(rownames(p9.top5) %in% hallmark.list[["GO.OXPHOS"]])], 5),
  "GO.OXPHOS" = select(p9.top5[head(rownames(p9.top5)[which(rownames(p9.top5) %in% hallmark.list[["GO.OXPHOS"]])], 5),], avg_logFC),
  "KEGG.OXPHOS" = head(rownames(p9.top5)[which(rownames(p9.top5) %in% hallmark.list[["KEGG.OXPHOS"]])], 5), 
  "KEGG.OXPHOS" = select(p9.top5[head(rownames(p9.top5)[which(rownames(p9.top5) %in% hallmark.list[["KEGG.OXPHOS"]])], 5),], avg_logFC)
)
rownames(p9.gs.df) <- seq(from = 1, length = nrow(p9.gs.df))

p8.gs.df
p9.gs.df
```

* **OBSERVATIONS** 
    - Note: All of these top 5 OXPHOS genes (logFC > 0.5) have a padj of 1.00
    - There are quite a few overlaps of OXPHOS genes with high LFC across all three genesets for both patient 8 and 9
    - The **UNUPDATED.OXPHOS** and the **GO.OXPHOS** genesets both seem to have top OXPHOS genes with higher logFC values relative to **KEGG.OXPHOS**

* Our answers to **QUESTION #1 and #2** seem to suggest that **UNUPDATED.OXPHOS** hallmark geneset is the best our of all the genesets tested because: 
    1. It has the least number of genes that are not found within the SS2 Seurat Object, and the highest percentage of genes that are found. 
    2. It has OXPHOS genes with the highest LFC (logFC > 0.5, regardless of padj) relative to the other two genesets for each model comparison between MRD and vehicle
    3. The high LFC OXPHOS genes present within this geneset have relatively higher logFC values (regardless of padj) in comparison to the high LFC OXPHOS genes present in the other two genesets, especially KEGG. 
* We answer our 3rd question to confirm whether the **UNUPDATED.OXPHOS** geneset is indeed the most appropriate one for our data relative to the other two genesets. 

**ANSWERING QUESTION #3: which geneset gives us the most statistically significant results**

```{r oxphosGS, message=FALSE, warning=FALSE, fig.align='center', fig.width=20, fig.height=10}
oxphos.centered <- c("UNUPDATED.OXPHOS37.centered", "GO.OXPHOS35.centered", "KEGG.OXPHOS36.centered")
SS2.Oxphos.Vln.plots <- vector("list", length(SS2s))
names(SS2.Oxphos.Vln.plots) <- SS2.names
patient <- c("8", "9")

for (i in 1:length(SS2s)){
  obj <- SS2s[[i]]
  name <- SS2.names[[i]]
  numCells <- nrow(SS2s[[i]]@meta.data)
  
  my_comparisons <- list(
    c(glue("{patient[[i]]}.0"), glue("{patient[[i]]}.1")), 
    c(glue("{patient[[i]]}.1"), glue("{patient[[i]]}.2")), 
    c(glue("{patient[[i]]}.0"), glue("{patient[[i]]}.2"))
  )
  
  if(patient[[i]] == "8"){
    p <- VlnPlot(obj, features = oxphos.centered, group.by = "sample", pt.size = 0, combine = F, cols = c("#00AFBB", "#E7B800", "#FC4E07"), y.max = 0.7)
  }
  else{
    p <- VlnPlot(obj, features = oxphos.centered, group.by = "sample", pt.size = 0, combine = F, cols = c("#00AFBB", "#E7B800", "#FC4E07"), y.max = 1.0)
  }
  unupdated.found <- sum(hallmark.list[["UNUPDATED.OXPHOS"]] %in% rownames(obj))
  unupdated.length <- length(hallmark.list[["UNUPDATED.OXPHOS"]])
  unupdated.pFound <- round((unupdated.found / unupdated.length)*100, 2)
  p[[1]] <- p[[1]] + labs(title = glue("{name} UNUPDATED.OXPHOS scores across sample"), x = name, subtitle = glue("{numCells} Malignant Cells, {unupdated.found} out of {unupdated.length} OXPHOS genes found ({unupdated.pFound}%)")) + 
    theme(plot.title = element_text(size = 12), plot.caption = element_text(size = 10)) + 
    geom_boxplot(width = 0.15, position = position_dodge(0.9), alpha = 0.3, show.legend = F) + 
    geom_text(label = paste(sum(str_detect(obj$sample, "[:digit:].0")), "cells"), x = glue("{patient[[i]]}.0"), y = min(obj$UNUPDATED.OXPHOS37.centered) -0.03) + 
    geom_text(label = paste(sum(str_detect(obj$sample, "[:digit:].1")), "cells"), x = glue("{patient[[i]]}.1"), y = min(obj$UNUPDATED.OXPHOS37.centered) - 0.03) + 
    geom_text(label = paste(sum(str_detect(obj$sample, "[:digit:].2")), "cells"), x = glue("{patient[[i]]}.2"), y = min(obj$UNUPDATED.OXPHOS37.centered) - 0.03) + 
    stat_compare_means(comparisons = my_comparisons, method = "wilcox.test", label = "p.format", step.increase = 0.05) + 
    stat_compare_means(comparisons = my_comparisons, method = "wilcox.test", label = "p.signif", step.increase = 0.05, bracket.size = 0, vjust = 1.8)
  
  GO.found <- sum(hallmark.list[["GO.OXPHOS"]] %in% rownames(obj))
  GO.length <- length(hallmark.list[["GO.OXPHOS"]])
  GO.pFound <- round((GO.found / GO.length)*100,2)
  p[[2]] <- p[[2]] + labs(title = glue("{name} GO.OXPHOS scores across sample"), x = name, subtitle = glue("{numCells} Malignant Cells, {GO.found} out of {GO.length} OXPHOS genes found ({GO.pFound}%)")) +
    theme(plot.title = element_text(size = 12), plot.caption = element_text(size = 10)) + 
    geom_boxplot(width = 0.15, position = position_dodge(0.9), alpha = 0.3, show.legend = F) + 
    geom_text(label = paste(sum(str_detect(obj$sample, "[:digit:].0")), "cells"), x = glue("{patient[[i]]}.0"), y = min(obj$GO.OXPHOS35.centered) -0.03) + 
    geom_text(label = paste(sum(str_detect(obj$sample, "[:digit:].1")), "cells"), x = glue("{patient[[i]]}.1"), y = min(obj$GO.OXPHOS35.centered) - 0.03) + 
    geom_text(label = paste(sum(str_detect(obj$sample, "[:digit:].2")), "cells"), x = glue("{patient[[i]]}.2"), y = min(obj$GO.OXPHOS35.centered) - 0.03) +
    stat_compare_means(comparisons = my_comparisons, method = "wilcox.test", label = "p.format", step.increase = 0.05) + 
    stat_compare_means(comparisons = my_comparisons, method = "wilcox.test", label = "p.signif", step.increase = 0.05, bracket.size = 0, vjust = 1.8)
  
  KEGG.found <- sum(hallmark.list[["KEGG.OXPHOS"]] %in% rownames(obj))
  KEGG.length <- length(hallmark.list[["KEGG.OXPHOS"]])
  KEGG.pFound <- round((KEGG.found / KEGG.length)*100,2)
  p[[3]] <- p[[3]] + labs(title = glue("{name} KEGG.OXPHOS scores across sample"), x = name, subtitle = glue("{numCells} Malignant Cells, {KEGG.found} out of {KEGG.length} OXPHOS genes found ({KEGG.pFound}%)")) +
    theme(plot.title = element_text(size = 12), plot.caption = element_text(size = 10)) + 
    geom_boxplot(width = 0.15, position = position_dodge(0.9), alpha = 0.3, show.legend = F) + 
    geom_text(label = paste(sum(str_detect(obj$sample, "[:digit:].0")), "cells"), x = glue("{patient[[i]]}.0"), y = min(obj$KEGG.OXPHOS36.centered) -0.03) + 
    geom_text(label = paste(sum(str_detect(obj$sample, "[:digit:].1")), "cells"), x = glue("{patient[[i]]}.1"), y = min(obj$KEGG.OXPHOS36.centered) - 0.03) + 
    geom_text(label = paste(sum(str_detect(obj$sample, "[:digit:].2")), "cells"), x = glue("{patient[[i]]}.2"), y = min(obj$KEGG.OXPHOS36.centered) - 0.03) +
    stat_compare_means(comparisons = my_comparisons, method = "wilcox.test", label = "p.format", step.increase = 0.05) + 
    stat_compare_means(comparisons = my_comparisons, method = "wilcox.test", label = "p.signif", step.increase = 0.05, bracket.size = 0, vjust = 1.8)
  
  p <- p[[1]] + p[[2]] + p[[3]] + plot_layout(guides= 'collect')
  
  SS2.Oxphos.Vln.plots[[name]] <- p
}

SS2.Oxphos.Vln.plots[["SS2 Patient 8"]]
SS2.Oxphos.Vln.plots[["SS2 Patient 9"]]
```

* **OBSERVATIONS**
    - The overall trend we see between treatment groups within each model is the same across all three oxphos genesets. However, it seems like the difference in scores is most obvious/drastic when using the **KEGG** geneset. 
    - Statistical tests show us that while **KEGG** gives us the most statistically significant results in Patient 8 and **UNUPDATED** gives us the most statistically significant results in Patient 9, the **GO** geneset is actually the most consistent, giving us statistically signifcant results in both patients. 

* **CONCLUSIONS** 
    - Although the **UNUPDATED** seems like the most suitable one for criteria 1 and 2, the **GO** geneset gives us the best result for the 3rd criterion: 
        1. Question 1: **GO** has the most number of unfound genes in the SS2 object. However: 
        2. Question 2: It has a pretty close tie in comparison to the **UNUPDATED** geneset, where they had similar numbers of high LFC OXPHOS genes present from FindMarkers, and the top high LFC OXPHOS genes in both genesets have comparable logFC. 
        3. Question 3: **GO** gives us statistically significant results in both patients. 
* Therefore, considering these three criteria, we decide that the **GO** geneset is best for our SS2 data. 

* **SUMMARY STATISTICS**
```{r oxphosStat, message=FALSE, warning=FALSE}
#Patient 8 ---------------------------------
SS2Malignant8.0 <- subset(SS2Malignant.8, subset = (sample == "8.0"))
SS2Malignant8.1 <- subset(SS2Malignant.8, subset = (sample == "8.1"))
SS2Malignant8.2 <- subset(SS2Malignant.8, subset = (sample == "8.2"))

SS2M8.uhm.oxphos.df <- data.frame(
  "p.0vsp.1" = wilcox.test(SS2Malignant8.0$UNUPDATED.OXPHOS37.centered, SS2Malignant8.1$UNUPDATED.OXPHOS37.centered)$p.value, 
  "p.1vsp.2" = wilcox.test(SS2Malignant8.1$UNUPDATED.OXPHOS37.centered, SS2Malignant8.2$UNUPDATED.OXPHOS37.centered)$p.value, 
  "p.0vsp.2" = wilcox.test(SS2Malignant8.0$UNUPDATED.OXPHOS37.centered, SS2Malignant8.2$UNUPDATED.OXPHOS37.centered)$p.value
)

SS2M8.go.oxphos.df <- data.frame(
  "p.0vsp.1" = wilcox.test(SS2Malignant8.0$GO.OXPHOS35.centered, SS2Malignant8.1$GO.OXPHOS35.centered)$p.value, 
  "p.1vsp.2" = wilcox.test(SS2Malignant8.1$GO.OXPHOS35.centered, SS2Malignant8.2$GO.OXPHOS35.centered)$p.value, 
  "p.0vsp.2" = wilcox.test(SS2Malignant8.0$GO.OXPHOS35.centered, SS2Malignant8.2$GO.OXPHOS35.centered)$p.value
)

SS2M8.kegg.oxphos.df <- data.frame(
  "p.0vsp.1" = wilcox.test(SS2Malignant8.0$KEGG.OXPHOS36.centered, SS2Malignant8.1$KEGG.OXPHOS36.centered)$p.value, 
  "p.1vsp.2" = wilcox.test(SS2Malignant8.1$KEGG.OXPHOS36.centered, SS2Malignant8.2$KEGG.OXPHOS36.centered)$p.value, 
  "p.0vsp.2" = wilcox.test(SS2Malignant8.0$KEGG.OXPHOS36.centered, SS2Malignant8.2$KEGG.OXPHOS36.centered)$p.value
)

#Patient 9 ---------------------------------
SS2Malignant9.0 <- subset(SS2Malignant.9, subset = (sample == "9.0"))
SS2Malignant9.1 <- subset(SS2Malignant.9, subset = (sample == "9.1"))
SS2Malignant9.2 <- subset(SS2Malignant.9, subset = (sample == "9.2"))

SS2M9.uhm.oxphos.df <- data.frame(
  "p.0vsp.1" = wilcox.test(SS2Malignant9.0$UNUPDATED.OXPHOS37.centered, SS2Malignant9.1$UNUPDATED.OXPHOS37.centered)$p.value, 
  "p.1vsp.2" = wilcox.test(SS2Malignant9.1$UNUPDATED.OXPHOS37.centered, SS2Malignant9.2$UNUPDATED.OXPHOS37.centered)$p.value, 
  "p.0vsp.2" = wilcox.test(SS2Malignant9.0$UNUPDATED.OXPHOS37.centered, SS2Malignant9.2$UNUPDATED.OXPHOS37.centered)$p.value
)

SS2M9.go.oxphos.df <- data.frame(
  "p.0vsp.1" = wilcox.test(SS2Malignant9.0$GO.OXPHOS35.centered, SS2Malignant9.1$GO.OXPHOS35.centered)$p.value, 
  "p.1vsp.2" = wilcox.test(SS2Malignant9.1$GO.OXPHOS35.centered, SS2Malignant9.2$GO.OXPHOS35.centered)$p.value, 
  "p.0vsp.2" = wilcox.test(SS2Malignant9.0$GO.OXPHOS35.centered, SS2Malignant9.2$GO.OXPHOS35.centered)$p.value
)

SS2M9.kegg.oxphos.df <- data.frame(
  "p.0vsp.1" = wilcox.test(SS2Malignant9.0$KEGG.OXPHOS36.centered, SS2Malignant9.1$KEGG.OXPHOS36.centered)$p.value, 
  "p.1vsp.2" = wilcox.test(SS2Malignant9.1$KEGG.OXPHOS36.centered, SS2Malignant9.2$KEGG.OXPHOS36.centered)$p.value, 
  "p.0vsp.2" = wilcox.test(SS2Malignant9.0$KEGG.OXPHOS36.centered, SS2Malignant9.2$KEGG.OXPHOS36.centered)$p.value
)

#combine ------------------------------
hm.oxphos.DF <- rbind(SS2M8.uhm.oxphos.df, SS2M9.uhm.oxphos.df)
rownames(hm.oxphos.DF) <- c("HM.OXPHOS.P8", "HM.OXPHOS.P9")

go.oxphos.DF <- rbind(SS2M8.go.oxphos.df, SS2M9.go.oxphos.df)
rownames(go.oxphos.DF) <- c("GO.OXPHOS.P8", "GO.OXPHOS.P9")

kegg.oxphos.DF <- rbind(SS2M8.kegg.oxphos.df, SS2M9.kegg.oxphos.df)
rownames(kegg.oxphos.DF) <- c("KEGG.OXPHOS.P8", "KEGG.OXPHOS.P9")

all.oxphos.DF <- rbind(hm.oxphos.DF, go.oxphos.DF, kegg.oxphos.DF)
DT::datatable(all.oxphos.DF) %>% 
  DT::formatSignif(names(all.oxphos.DF), digits = 2) %>% 
  DT::formatStyle(names(all.oxphos.DF), color = DT::styleInterval(0.05, c('red', 'black')))
```

We also tested how the **UNUPDATED** version of **HALLMARK_UNFOLDED_PROTEIN_RESPONSE** differs from the updated version.
```{r uprGS, message = FALSE, warning = FALSE}
upr.length.df <- data.frame(
  "UNUPDATED.HM.UPR" = length(hallmark.list[["UNUPDATED.UPR"]]),
  "HALLMARK.UPR" = length(hallmark.list[["HALLMARK_UNFOLDED_PROTEIN_RESPONSE"]])
)

Found.upr.df <- data.frame(
  "UNUPDATED.HM.UPR" = sum((hallmark.list[["UNUPDATED.UPR"]] %in% rownames(SS2Malignant))),
  "HALLMARK.UPR" = sum((hallmark.list[["HALLMARK_UNFOLDED_PROTEIN_RESPONSE"]] %in% rownames(SS2Malignant)))
)

PFound.upr.df <- data.frame(
  "UNUPDATED.HM.UPR" = (sum((hallmark.list[["UNUPDATED.UPR"]] %in% rownames(SS2Malignant)))/length(hallmark.list[["UNUPDATED.UPR"]]))*100,
  "HALLMARK.UPR" = (sum((hallmark.list[["HALLMARK_UNFOLDED_PROTEIN_RESPONSE"]] %in% rownames(SS2Malignant)))/length(hallmark.list[["HALLMARK_UNFOLDED_PROTEIN_RESPONSE"]])) * 100
)

upr.df <- rbind(upr.length.df, Found.upr.df, PFound.upr.df)
upr.df[,"UNUPDATED.HM.UPR"] <- round(upr.df[,"UNUPDATED.HM.UPR"])
upr.df[,"HALLMARK.UPR"] <- round(upr.df[,"HALLMARK.UPR"])
rownames(upr.df) <- c("Num UPR Genes", "Num Found", "% Found")
upr.df
```

* Since the **UNUPDATED** version identifies two more genes than the updated version, we decide to proceed with the **UNUPDATED.UPR** geneset. 

After deciding the geneset to use for our data, we can now analyze the differential expression of OXPHOS and UPR genes across treatment condition within each model: 
    - **OXPHOS**: GO_OXIDATIVE_PHOSPHORYLATION geneset 
    - **UPR**: unupdated HALLMARK_UNFOLDED_PROTEIN_RESPONSE geneset

### STEP 3 DE ANALYSIS #1. VISUALIZING AND QUANTIFYING DE HALLMARK GENESETS
* **QUESTION** Are modules (**OXPHOS and UPR**) differentially expressed across treatment conditions within each patient? 
* **HYPOTHESIS** We hypothesize that the on-treatment time-points (**p.1 and p.2**) will express enriched levels of OXPHOS and UPR hallmarks relative to cells treatment-naive time-point (**p.0**). We are not so sure what to expect between p.1 and p.2 on-treatment samples because we are not told how the two samples are different in terms of their treatment status. 
    - **APPROACH #1** Violin Plots and UMAP
        * Visualize differences in hallmark score across treatment conditions with:
            - Violin Plot (center module scores at 0)
        * Statistical test (**wilcoxon rank sum test**): is the difference in hallmark score across treatment conditions statistically significant? 

```{r DEAnalysis1VlnPlot, message = FALSE, warning = FALSE, fig.align = 'center', fig.width = 17, fig.height = 10}
hms.centered = c("GO.OXPHOS35.centered", "UNUPDATED.UPR38.centered")
#SWARM PLOTS 
#OXPHOS ----------
oxphos.swarm.plots <- vector("list", length = 2)
names(oxphos.swarm.plots) <- SS2.names
patient <- c("8", "9")

for(i in 1:2){
  SS2 <- SS2s[[i]]
  SS2.name <- SS2.names[[i]]
  
  my_comparisons <- list(
    c(glue("{patient[[i]]}.0"), glue("{patient[[i]]}.1")), 
    c(glue("{patient[[i]]}.1"), glue("{patient[[i]]}.2")), 
    c(glue("{patient[[i]]}.0"), glue("{patient[[i]]}.2"))
  )
  
  feature.found <- sum(hallmark.list[["UNUPDATED.OXPHOS"]] %in% rownames(SS2))
  feature.length <- length(hallmark.list[["UNUPDATED.OXPHOS"]])
  feature.pFound <- round((feature.found / feature.length)*100, 2)
  numCells <- nrow(SS2@meta.data)
  
  p <- ggplot(SS2@meta.data, aes(x= sample , y = UNUPDATED.OXPHOS37.centered, color = sample)) +
  geom_quasirandom(groupOnX =TRUE) + 
  theme_bw() + 
  scale_color_manual(values = c("#00AFBB", "#E7B800", "#FC4E07")) + 
    labs(title = glue("{SS2.name} Malignant Cells OXPHOS expression across sample"), x = SS2.name, subtitle = glue("{numCells} Malignant Cells, {feature.found} out of {feature.length} OXPHOS genes found ({feature.pFound}%)")) + 
  geom_boxplot(width = 0.10, position = position_dodge(0.9), alpha = 1, show.legend = F, color = "black", outlier.alpha = 0) + 
    geom_text(label = paste(sum(str_detect(SS2$sample, "[:digit:].0")), "cells"), x = glue("{patient[[i]]}.0"), y = min(SS2$UNUPDATED.OXPHOS37.centered) -0.03, color = "black", show.legend = FALSE) + 
    geom_text(label = paste(sum(str_detect(SS2$sample, "[:digit:].1")), "cells"), x = glue("{patient[[i]]}.1"), y = min(SS2$UNUPDATED.OXPHOS37.centered) - 0.03, color = "black", show.legend = FALSE) + 
    geom_text(label = paste(sum(str_detect(SS2$sample, "[:digit:].2")), "cells"), x = glue("{patient[[i]]}.2"), y = min(SS2$UNUPDATED.OXPHOS37.centered) - 0.03, color = "black", show.legend = FALSE) + 
    stat_compare_means(comparisons = my_comparisons, method = "wilcox.test", label = "p.format", step.increase = 0.06) + 
    stat_compare_means(comparisons = my_comparisons, method = "wilcox.test", label = "p.signif", step.increase = 0.06, bracket.size = 0, vjust = 1.8)
  
  oxphos.swarm.plots[[SS2.name]] <- p 
}

oxphos.swarm.plots[["SS2 Patient 8"]] + oxphos.swarm.plots[["SS2 Patient 9"]]


#UPR ----------
upr.swarm.plots <- vector("list", length = 2)
names(upr.swarm.plots) <- SS2.names

for(i in 1:2){
  SS2 <- SS2s[[i]]
  SS2.name <- SS2.names[[i]]
  
  my_comparisons <- list(
    c(glue("{patient[[i]]}.0"), glue("{patient[[i]]}.1")), 
    c(glue("{patient[[i]]}.1"), glue("{patient[[i]]}.2")), 
    c(glue("{patient[[i]]}.0"), glue("{patient[[i]]}.2"))
  )
  
  feature.found <- sum(hallmark.list[["UNUPDATED.UPR"]] %in% rownames(SS2))
  feature.length <- length(hallmark.list[["UNUPDATED.UPR"]])
  feature.pFound <- round((feature.found / feature.length)*100, 2)
  numCells <- nrow(SS2@meta.data)
  
  p <- ggplot(SS2@meta.data, aes(x= sample , y = UNUPDATED.UPR38.centered, color = sample)) +
  geom_quasirandom(groupOnX =TRUE) + 
  theme_bw() + 
  scale_color_manual(values = c("#00AFBB", "#E7B800", "#FC4E07")) + 
    labs(title = glue("{SS2.name} Malignant Cells UPR expression across sample"), x = SS2.name, subtitle = glue("{numCells} Malignant Cells, {feature.found} out of {feature.length} UPR genes found ({feature.pFound}%)")) + 
  geom_boxplot(width = 0.10, position = position_dodge(0.9), alpha = 1, show.legend = F, color = "black", outlier.alpha = 0) + 
    geom_text(label = paste(sum(str_detect(SS2$sample, "[:digit:].0")), "cells"), x = glue("{patient[[i]]}.0"), y = min(SS2$UNUPDATED.UPR38.centered) -0.03, color = "black", show.legend = FALSE) + 
    geom_text(label = paste(sum(str_detect(SS2$sample, "[:digit:].1")), "cells"), x = glue("{patient[[i]]}.1"), y = min(SS2$UNUPDATED.UPR38.centered) - 0.03, color = "black", show.legend = FALSE) + 
    geom_text(label = paste(sum(str_detect(SS2$sample, "[:digit:].2")), "cells"), x = glue("{patient[[i]]}.2"), y = min(SS2$UNUPDATED.UPR38.centered) - 0.03, color = "black", show.legend = FALSE) + 
    stat_compare_means(comparisons = my_comparisons, method = "wilcox.test", label = "p.format", step.increase = 0.06) + 
    stat_compare_means(comparisons = my_comparisons, method = "wilcox.test", label = "p.signif", step.increase = 0.06, bracket.size = 0, vjust = 1.8)
  
  upr.swarm.plots[[SS2.name]] <- p 
}

upr.swarm.plots[["SS2 Patient 8"]] + upr.swarm.plots[["SS2 Patient 9"]]

```

* **CONCLUSIONS**
    - The statistically significant results are (p < 0.05): 
        * **Patient 8**
            - OXPHOS: Sample 8.0 > 8.1 ; 8.0 > 8.2 (does not support our hypothesis)
            - **UPR: Sample 8.2 > 8.0 ; 8.1 > 8.0 (supports our hypothesis)**
        * **Patient 9** 
            - **OXPHOS: Sample 9.1 > 9.2 ; 9.1 > 9.0 (supports our hypothesis)** ; 9.0 > 9.2 (does not support our hypothesis)
    - The trends that support our hypothesis for both OXPHOS and UPR are not consistent across patients 8 and 9. Since our sample size is so small, we're also not sure how representative this data is of the true population. 
    - We are, however, not sure how to compare between p.1 vs. p.2 because they are both categorized as "on-treatment" and the clinical status of samples obtained from patients 8 and 9 is "Recurrent", instead of denoting two types of treatment status "Diagnosis/Initial treatment" that corresponds to the two treatment timepoints for the 10X data. 

* **STATISTICAL SUMMARY** 
```{r DEAnalysis1Statistical test, message = FALSE, warning = FALSE}
#Patient 8 ---------------------------------
SS2M8.oxphos.df <- data.frame(
  "p.0vsp.1" = wilcox.test(SS2Malignant8.0$GO.OXPHOS35.centered, SS2Malignant8.1$GO.OXPHOS35.centered)$p.value, 
  "p.1vsp.2" = wilcox.test(SS2Malignant8.1$GO.OXPHOS35.centered, SS2Malignant8.2$GO.OXPHOS35.centered)$p.value, 
  "p.0vsp.2" = wilcox.test(SS2Malignant8.0$GO.OXPHOS35.centered, SS2Malignant8.2$GO.OXPHOS35.centered)$p.value
)

SS2M8.UPR.df <- 
  data.frame(
  "p.0vsp.1" = wilcox.test(SS2Malignant8.0$UNUPDATED.UPR38.centered, SS2Malignant8.1$UNUPDATED.UPR38.centered)$p.value, 
  "p.1vsp.2" = wilcox.test(SS2Malignant8.1$UNUPDATED.UPR38.centered, SS2Malignant8.2$UNUPDATED.UPR38.centered)$p.value, 
  "p.0vsp.2" = wilcox.test(SS2Malignant8.0$UNUPDATED.UPR38.centered, SS2Malignant8.2$UNUPDATED.UPR38.centered)$p.value
)

#Patient 9 ---------------------------------
SS2M9.oxphos.df <- data.frame(
  "p.0vsp.1" = wilcox.test(SS2Malignant9.0$GO.OXPHOS35.centered, SS2Malignant9.1$GO.OXPHOS35.centered)$p.value, 
  "p.1vsp.2" = wilcox.test(SS2Malignant9.1$GO.OXPHOS35.centered, SS2Malignant9.2$GO.OXPHOS35.centered)$p.value, 
  "p.0vsp.2" = wilcox.test(SS2Malignant9.0$GO.OXPHOS35.centered, SS2Malignant9.2$GO.OXPHOS35.centered)$p.value
)

SS2M9.UPR.df <- 
  data.frame(
  "p.0vsp.1" = wilcox.test(SS2Malignant9.0$UNUPDATED.UPR38.centered, SS2Malignant9.1$UNUPDATED.UPR38.centered)$p.value, 
  "p.1vsp.2" = wilcox.test(SS2Malignant9.1$UNUPDATED.UPR38.centered, SS2Malignant9.2$UNUPDATED.UPR38.centered)$p.value, 
  "p.0vsp.2" = wilcox.test(SS2Malignant9.0$UNUPDATED.UPR38.centered, SS2Malignant9.2$UNUPDATED.UPR38.centered)$p.value
)

#combine ------------------------------
SS2.oxphos.DF <- rbind(SS2M8.oxphos.df, SS2M9.oxphos.df)
rownames(SS2.oxphos.DF) <- c("OXPHOS.PATIENT8", "OXPHOS.PATIENT9")

SS2.UPR.DF <- rbind(SS2M8.UPR.df, SS2M9.UPR.df)
rownames(SS2.UPR.DF) <- c("UPR.PATIENT8", "UPR.PATIENT9")

SS2.all.DF <- rbind(SS2.oxphos.DF, SS2.UPR.DF)
DT::datatable(SS2.all.DF) %>% 
  DT::formatSignif(names(SS2.all.DF), digits = 2) %>% 
  DT::formatStyle(names(SS2.all.DF), color = DT::styleInterval(0.05, c('red', 'black')))
```


We try to confirm our results again with our second appraoch: **GSEA**

* **APPROACH #2** Geneset Enrichment Analysis (GSEA)
      - GSEA enrichment plots for hallmarks of interest between condition.1 vs. condition.2 
      - Rank genes and compute GSEA enrichment scores 
      - Statistical test: how significant are the enrichment scores? 

**have not finalized the statistical test and ranking method to use for GSEA**

```{r GSEA, message=FALSE, warning=FALSE, fig.align='center', fig.width=20, fig.height=10,}

```

### STEP 3 DE ANALYSIS #2. IDENTIFYING INDIVIDUAL DE GENES
We are interesting in detecting individual OXPHOS and UPR genes that might be differentially expressed. 

* We will focus our analysis on Patient 8 UPR and Patient 9 OXPHOS genes because those are the s**tatistically significant** gene-set scores that **support our hypothesis** based on the results we got from our VlnPlots.

```{r VolcanoPlotSig, message = FALSE, warning = FALSE, fig.align='center', fig.width = 25, fig.height = 15}

# PATIENT 8 -----------------------
p8.1p8.0 <- FindMarkers(SS2Malignant.8, group.by = "sample", ident.1 = "8.1", ident.2 = "8.0", test.use = "wilcox", logfc.threshold = 0)
p8.2p8.0 <- FindMarkers(SS2Malignant.8, group.by = "sample", ident.1 = "8.2", ident.2 = "8.0", test.use = "wilcox", logfc.threshold = 0)
p8.1p8.2 <- FindMarkers(SS2Malignant.8, group.by = "sample", ident.1 = "8.1", ident.2 = "8.2", test.use = "wilcox", logfc.threshold = 0)

p8.all.plots <- vector("list", length = 3)
samples <- c("8.1 8.0", "8.2 8.0", "8.1 8.2")
names(p8.all.plots) <- samples

p8.UPR.plots <- vector("list", length = 3)
names(p8.UPR.plots) <- samples
markers <- list(p8.1p8.0, p8.2p8.0, p8.1p8.2)

for(i in 1:3){
  marker <- markers[[i]]
  name = samples[[i]]
  name.split <- stringr::str_split(samples[[i]], pattern = " ")
  group.1 <- name.split[[1]][1]
  group.2 <- name.split[[1]][2]
  
  p8.all.plots[[name]] <- DEAnalysis_code(SS2Malignant.8, markers = marker, group.by = "sample", group.1 = group.1, group.2 = group.2, graph = TRUE)
  
  p8.UPR.plots[[name]] <- DEAnalysis_code(SS2Malignant.8, markers = marker, group.by = "sample", group.1 = group.1, group.2 = group.2, geneset = hallmark.list[["UNUPDATED.UPR"]]) + labs(title = paste(group.1, "vs", group.2, "DE UPR Genes"))
}

p <- p8.all.plots[["8.1 8.0"]] + p8.all.plots[["8.2 8.0"]] + p8.all.plots[["8.1 8.2"]] + p8.UPR.plots[["8.1 8.0"]] + p8.UPR.plots[["8.2 8.0"]] + p8.UPR.plots[["8.1 8.2"]]
p
ggsave(plot = p, filename = "SS2Malignant8.Volcano.png", path = "jesslyn_plots/SS2/DE", width = 25, height = 15)

# PATIENT 9 -----------------------
p9.1p9.0 <- FindMarkers(SS2Malignant.9, group.by = "sample", ident.1 = "9.1", ident.2 = "9.0", test.use = "wilcox", logfc.threshold = 0)
p9.2p9.0 <- FindMarkers(SS2Malignant.9, group.by = "sample", ident.1 = "9.2", ident.2 = "9.0", test.use = "wilcox", logfc.threshold = 0)
p9.1p9.2 <- FindMarkers(SS2Malignant.9, group.by = "sample", ident.1 = "9.1", ident.2 = "9.2", test.use = "wilcox", logfc.threshold = 0)

p9.all.plots <- vector("list", length = 3)
samples <- c("9.1 9.0", "9.2 9.0", "9.1 9.2")
names(p9.all.plots) <- samples

p9.OXPHOS.plots <- vector("list", length = 3)
names(p9.OXPHOS.plots) <- samples
markers <- list(p9.1p9.0, p9.2p9.0, p9.1p9.2)

for(i in 1:3){
  marker <- markers[[i]]
  name = samples[[i]]
  name.split <- stringr::str_split(samples[[i]], pattern = " ")
  group.1 <- name.split[[1]][1]
  group.2 <- name.split[[1]][2]
  
  p9.all.plots[[name]] <- DEAnalysis_code(SS2Malignant.9, markers = marker, group.by = "sample", group.1 = group.1, group.2 = group.2, graph = TRUE)
  
  p9.OXPHOS.plots[[name]] <- DEAnalysis_code(SS2Malignant.9, markers = marker, group.by = "sample", group.1 = group.1, group.2 = group.2, geneset = hallmark.list[["GO.OXPHOS"]]) + labs(title = paste(group.1, "vs", group.2, "DE OXPHOS Genes"))
}

p2 <- p9.all.plots[["9.1 9.0"]] + p9.all.plots[["9.2 9.0"]] + p9.all.plots[["9.1 9.2"]] + p9.OXPHOS.plots[["9.1 9.0"]] + p9.OXPHOS.plots[["9.2 9.0"]] + p9.OXPHOS.plots[["9.1 9.2"]]
p2
ggsave(plot = p2, filename = "SS2Malignant9.Volcano.png", path = "jesslyn_plots/SS2/DE", width = 25, height = 15)

```

* **OBSERVATIONS** 
    - No statistically significant individual DE UPR genes in Patient 8 across all three conditions
    - 9.2 vs. 9.0: **9.2 depletes the expression of NDUFB11 relative to 9.0**
    - 9.1 vs. 9.2: **9.1 enriches NDUFB1 relative to 9.2** 

* We will also look at Patient 8 OXPHOS and Patient 9 UPR genes just in case a few genes are differentially expressed but not collectively, which may justify why the gene-sets are not differentially expressed as a whole, but specific OXPHOS or UPR genes may be differentially expressed in drug resistant persisters.


```{r VolcanoPlot2, message = FALSE, warning = FALSE, fig.align='center', fig.width = 25, fig.height = 10}

# PATIENT 8 -----------------------
p8.OXPHOS.plots <- vector("list", length = 3)
samples <- c("8.1 8.0", "8.2 8.0", "8.1 8.2")
names(p8.OXPHOS.plots) <- samples
markers <- list(p8.1p8.0, p8.2p8.0, p8.1p8.2)

for(i in 1:3){
  marker <- markers[[i]]
  name = samples[[i]]
  name.split <- stringr::str_split(samples[[i]], pattern = " ")
  group.1 <- name.split[[1]][1]
  group.2 <- name.split[[1]][2]
  
  p8.OXPHOS.plots[[name]] <- DEAnalysis_code(SS2Malignant.8, markers = marker, group.by = "sample", group.1 = group.1, group.2 = group.2, geneset = hallmark.list[["GO.OXPHOS"]]) + labs(title = paste(group.1, "vs", group.2, "DE OXPHOS Genes"))
}

p3 <- p8.OXPHOS.plots[["8.1 8.0"]] + p8.OXPHOS.plots[["8.2 8.0"]] + p8.OXPHOS.plots[["8.1 8.2"]]
p3
ggsave(plot = p3, filename = "SS2Malignant8.VolcanoOXPHOS.png", path = "jesslyn_plots/SS2/DE", width = 25, height = 10)

# PATIENT 9 -----------------------
p9.UPR.plots <- vector("list", length = 3)
samples <- c("9.1 9.0", "9.2 9.0", "9.1 9.2")
names(p9.UPR.plots) <- samples
markers <- list(p9.1p9.0, p9.2p9.0, p9.1p9.2)

for(i in 1:3){
  marker <- markers[[i]]
  name = samples[[i]]
  name.split <- stringr::str_split(samples[[i]], pattern = " ")
  group.1 <- name.split[[1]][1]
  group.2 <- name.split[[1]][2]
  
  p9.UPR.plots[[name]] <- DEAnalysis_code(SS2Malignant.9, markers = marker, group.by = "sample", group.1 = group.1, group.2 = group.2, geneset = hallmark.list[["UNUPDATED.UPR"]]) + labs(title = paste(group.1, "vs", group.2, "DE UPR Genes"))
}

p4 <- p9.UPR.plots[["9.1 9.0"]] + p9.UPR.plots[["9.2 9.0"]] + p9.UPR.plots[["9.1 9.2"]]
p4
ggsave(plot = p4, filename = "SS2Malignant9.VolcanoUPR.png", path = "jesslyn_plots/SS2/DE", width = 25, height = 10)

```

* Interestingly, there turned out to be more DE OXPHOS genes in Patient 8 comparisons than in Patient 9. However, the DE OXPHOS genes detected are not the same between patients. 
* Similar to Patient 8, there are no DE UPR genes in Patient 9 as well. 

Since DE genes detected are not consistent across patients, we cannot conclude that there is an enrichment of OXPHOS or UPR genes in p.1. 


