---
title: "Izar 2020 SS2 (Cohort 2) Load + Plots"
author: "Jesslyn Goh and Mike Cuoco"
date: "7/20/2020"
editor_options: 
  chunk_output_type: console
output:
 html_document:
  code_folding: hide
  toc_float: TRUE 
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = here::here())
knitr::opts_chunk$set(echo = TRUE)
options(width=120)
```

# IZAR 2020 SS2 (COHORT 2) DATA ANALYSIS - PART 1

### OVERVIEW
* This is the first and second part of our 3-part analysis of the Izar 2020 SS2 (Cohort 2) data. 
* The SS2 data consists of cells from 14 ascites samples from 6 individuals. 
    - selected for malignant EPCAM+CD24+ cells by fluorophore-activated cell sorting into 96-well plates
    - Includes both **malignant** and **nonmalignant** populations but differs from 10X data - examines malignant ascites more closely. 
    - We are only interested in the **malignant** ascites population

* We split our SS2 analysis into three parts: 
    1. **Loading and Processing Data**
        a. Load in Izar2020 10X data and create Seurat Objects
        b. Assign metadata (Patient, Sample IDs, treatment status)
        c. Subset the Seurat Object for Malignant Cells 
        d. Score Cells (Cell cycle and Each hallmark using **AddModuleScore**)
        e. Scale and Center Cells 
        f. Evaluate Patient and Treatment Metrics
        g. Dimensional Reduction (PCA and UMAP)
        
    2. **Visualizing Clusters (UMAP)**
        a. Visualize metadata identities with PCA and UMAP 
            i. Sample ID 
            ii. Treatment Status 
        b. Save Seurat Obejcts
        
    3. DE Analysis 

### STEP 1A LOAD SEURAT OBJECT

```{r init, message = FALSE, warning = FALSE}
# Load packages
source(here::here('packages.R'))

#Read in SS2 RDS object
SS2 = readRDS(file = "data/Izar_2020/Izar_2020_SS2.RDS")

# Read in gene lists
ccgenes = read_lines("data/gene_lists/regev_lab_cell_cycle_genes.txt")
s.genes <- ccgenes[1:43]
g2m.genes <- ccgenes[44:97]

#Read in hallmarks of interest
hallmark_names = read_lines("data/gene_lists/hallmarks.txt")
hallmark.list <- vector(mode = "list", length = length(hallmark_names))
names(hallmark.list) <- hallmark_names
for(hm in hallmark_names){
  file <- read_lines(glue("data/gene_lists/hallmarks/{hm}_updated.txt"), skip = 1)
  hallmark.list[[hm]] <- file
}
```

### STEP 1B ASSIGN SAMPLE ID TO METADATA 
```{r sampleID, message=FALSE, warning = FALSE}
#add sample
sample_ID <- rep("NA", ncol(SS2))
for(patient in unique(SS2$Patient)){
  for (time in unique(SS2$Time)){
    sample_ID[SS2$Patient == patient & SS2$Time == time] <- glue("{patient}.{time}")
  }
}
SS2 <- AddMetaData(SS2, sample_ID, col.name = "sample")

#add BRAC status
BRAC.status <- rep("NA", ncol(SS2))
BRAC.status[SS2$Patient %in% c(5,8,9,10,122)] <- "WT"
BRAC.status[SS2$Patient == 7] <- "BRCA2 MUT"
SS2 <- AddMetaData(SS2, BRAC.status, col.name = "BRAC.status")

#set levels 
SS2$sample = factor(SS2$sample, levels = trimws(sort(formatC(unique(SS2$sample)))))
SS2$treatment.status = factor(SS2$treatment.status, levels = c("Treatment-naïve","On-treatment"))
```

### STEP 1C-E SUBSET, SCORE, SCALE SEURAT OBJECT FOR MALIGNANT CELLS ONLY 
```{r subset and scale Malignant, message=FALSE, warning = FALSE}
#subset
SS2Malignant <- subset(SS2, subset = (cell.type == "Malignant"))

#score
SS2Malignant <- AddModuleScore(SS2Malignant, features = hallmark.list, name = names(hallmark.list), nbin = 25, search = T) 
SS2Malignant <- CellCycleScoring(SS2Malignant, g2m.features = g2m.genes, s.features = s.genes)

#scale
SS2Malignant <- ScaleData(SS2Malignant, do.scale = TRUE, do.center = TRUE)
```

### STEP 1F SS2 MALIGNANT CELLS METRICS 
```{r malignant cells metrics, message=FALSE, warning=FALSE}
#numMalignantCellsPerSample
numMalignantCellsPerSample <- table(SS2Malignant$sample, SS2Malignant$cell.type, SS2Malignant$treatment.status) %>%
  as.data.frame() %>% 
  dplyr::rename(sample = Var1, celltype = Var2, treatment = Var3, numCells = Freq) %>% 
  filter(`celltype` == "Malignant")
ggplot(numMalignantCellsPerSample, aes(x = sample, y=numCells, fill = treatment)) + 
  geom_bar(stat = "identity") + 
  labs(title = "Number of Malignant Cells Per Sample in SS2 Data")
ggsave("numMalignantCellsPerSample.png", path = "jesslyn_plots/SS2", width = 8, height = 8)

seurat_tab(SS2Malignant, col_var = "sample", row_var = "treatment.status", title = "Number of Malignant cells per Sample per treatment status")

seurat_tab(SS2Malignant, col_var = "Patient", row_var = "treatment.status", title = "Number of Malignant cells per Sample per Patient")

seurat_tab(SS2Malignant, col_var = "Patient", row_var = "Time", title = "Number of Malignant cells per Sample per Patient")
```

### STEP 1G. DIMENSIONALITY REDUCTION 
```{r Dimensionality Reduction, message=FALSE, warning=FALSE}
#PCA data has already been scaled and centered above
SS2Malignant <- FindVariableFeatures(SS2Malignant, selection.method = "vst")

SS2Malignant <- RunPCA(SS2Malignant, features = VariableFeatures(SS2Malignant))
ElbowPlot(SS2Malignant)

#UMAP
SS2Malignant<- RunUMAP(SS2Malignant, dims=1:15)
```

### STEP 2A. VISUALIZE METADATA WITH PCA AND UMAP 
```{r PCA and UMAP, message=FALSE, warning=FALSE, fig.align='center', fig.width=20}
PCAPlot(SS2Malignant)

umap1 <- UMAPPlot(SS2Malignant, group.by = "sample") + labs(title = "SS2 Malignant UMAP by Sample")
umap2 <- UMAPPlot(SS2Malignant, group.by = "treatment.status") + labs(title = "SS2 Malignant UMAP by Treatment Status")
umap <- umap1 + umap2 
umap
ggsave(plot = umap, filename = "SS2Malignant_UMAP.png", path = "jesslyn_plots/SS2")
```

### STEP 2B. SAVE SEURAT OBJECT 
```{r save SDS}
saveRDS(SS2Malignant, file="data/Izar_2020/jesslyn_SS2Malignant_processed.RDS")
```
