---
title: "Izar 2020 PDX (Cohort 3) DE Analysis"
author: "Jesslyn Goh and Mike Cuoco"
date: "7/20/2020"
editor_options: 
  chunk_output_type: console
output:
 html_document:
  code_folding: hide
  toc_float: TRUE 
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = here::here())
knitr::opts_chunk$set(echo = TRUE)
options(width=120)
```

# IZAR 2020 PDX (COHORT 3) DATA DIFFERANTIAL EXPRESISON ANALYSIS 

### OVERVIEW
* The PDX data from Izar2020 consists of **only Malignant cells** from three HGSOC PDX models derived from patients with different treatment histories were selected for implantation: 
    - DF20 (BRCA WT treatment-naive, clinically platinum sensitive) 
    - DF101 (BRCA1 mutant, 2 lines of prior therapy, clinically platinum resistant) 
    - DF68 (BRCA1 mutant, 6 lines of prior therapy, clinically platinum resistant) 
* After tumors were established, animals were divided into two groups per model: 
    - Vehicle (treated with DMSO) 
    - Carboplatin (treated with IP carboplatin) 
        * Carboplatin-treated mice for minimal residual disease (**MRD**) group were harvested for scRNA-seq 
        * The remaining carboplatin-treated mice were harvested at endpoint (**vehicle**)
* In our 3-part analysis of the Izar 2020 PDX data, we are interested in identifying differentially expressed genes and hallmark genesets between treatment statuses **within** each model. 
* We split our PDX analysis into three parts: 
    1. **Load Data and Create PDX Seurat Object**
        a. The code to this part of our analysis is in the **read_Izar_2020.R** file in the **code** folder. During this part of our analysis we: 
            i. Load in PDX count matrix and Create Seurat Object 
            ii. Assign Metadata identities including: 
                - Mouse ID 
                - Model ID 
                - Treatment Status 
            iii. Score cells for **cell cycle** and **hallmark genesets** 
                - **Note**: It does not matter whether we call AddModuleScore before or after subsetting and scaling each model because AddModuleScore uses the **data** slot. 
            iv. Save Seurat Object
          
    2. **Process Data and Exploratory Data Analysis**
        a. The code to this part of our analysis can be found in the **03_Izar2020_PDX_Load** file in the **old/edited** folder. During this part of our analysis we: 
            i. Load in PDX Seurat Object from Part 1 and **subset by model**. Continue analysis separately for each model. 
            ii. Scale and FindVariableFeatures (prepares data for dimensionality reduction)
            iii. Dimensionality Reduction (PCA + UMAP)
            iv. Visualize how cells separate based on metadata identities via UMAP 
                  - Intermodel heterogeneity: How do cells separate by model? 
                  - Intramodel heterogeneity: 
                      * How do cells separate by treament status? 
                      * How do cells separate by cell cycle phase? 
            v. Compute summary metrics for PDX data such as:
                - Number of cells per model per treatment 
                - Number of cells per treatment per cell cycle phase 
            vi. Save Seurat Objects 
            
    3. **DE Analysis**
        a. **TYPE #1 DE ANALYSIS**: Visualizing and Quantifying Differentially Expression on **Predefined** GO Genesets
            i. Violin Plots and UMAP 
            ii. Gene Set Enrichment Analysis (GSEA)
        b. **TYPE #2 DE ANALYSIS**: Finding DE Genes from scratch
            i. Volcano Plots 
        c. **CELL CYCLE ANALYSIS**
            i. Evaluate the idea that cell cycle might influence expression of signatures 

This is the third part of our 3-part analysis of the Izar 2020 PDX (Cohort 3) data. 

### DIFFERENTIAL EXPRESSION ANALYSIS IN-DEPTH EXPLANATION 
We are interested in answering a few questions for our DE Analysis:

**DE ANALYSIS #1. Visualizing and Quantifying DE Hallmark Genesets** 

* **QUESTION** Are modules (**OXPHOS and UPR**) differentially expressed across treatment conditions within each model? 
    - **APPROACH #1** Violin Plots and UMAP 
        * Visualize differences in hallmark score across treatment conditions with: 
            - Violin Plot 
            - UMAP by treatment vs. UMAP by hallmark score 
        * Statistical test: is the difference in hallmark score across treatment conditions statistically significant? 
    - **APPROACH #2** GSEA
        * GSEA enrichment plots for hallmarks of interest between condition.1 vs. condition.2 
        * Rank genes and compute GSEA enrichment scores 
        * Statistical test: how significant are the enrichment scores? 

**DE ANALYSIS #2. Identifying Individual DE Genes** 

**CELL CYCLE ANALYSIS** 

* **QUESTION #1** Are more cells at different cell cycle phases between treatments? 
    - **APPROACH** Violin Plots and UMAP 
        * Visualize differences in cell cycle score across treatment conditions with: 
            - Stacked Bar Plot
            - Violin Plot 
            - UMAP by treatment vs. UMAP by cell cycle phase vs. UMAP by S and G2M Score 
        * Statistical test: is the difference in cell cycle scores across treatment conditions statistically
* **QUESTION #2** Is there a correlation between cell cycle phase and the expression of signatures? 
    - **APPROACH** Violin Plots
        * Create Violin Plots for each hallmark of interest like **DE Analysis #1/APPROACH #1. 
        * Stratify the Violin Plots by cell cycle phase

### STEP 1 LOAD IN SEURAT OBJECTS AND GENESETS 

```{r init, message = FALSE, warning = FALSE}
# Load packages
source(here::here('packages.R'))

#Read in PDX RDS object
PDX_All = readRDS("data/Izar_2020/test/jesslyn_PDX_All_processed.RDS")
PDX_DF20 = readRDS("data/Izar_2020/test/jesslyn_PDX_DF20_processed.RDS")
PDX_DF101 = readRDS("data/Izar_2020/test/jesslyn_PDX_DF101_processed.RDS")
PDX_DF68 = readRDS("data/Izar_2020/test/jesslyn_PDX_DF68_processed.RDS")

#Read in hallmarks of interest
hallmark_names = read_lines("data/gene_lists/hallmarks.txt")
hallmark.list <- vector(mode = "list", length = length(hallmark_names))
names(hallmark.list) <- hallmark_names
for(hm in hallmark_names){
  file <- read_lines(glue("data/gene_lists/hallmarks/{hm}_updated.txt"), skip = 1)
  hallmark.list[[hm]] <- file
}

#center module and cell cycle scores and reassign to the metadata of each Seurat object
hm.names <- names(PDX_All@meta.data)[9:44]

for(i in hm.names){
    DF20.hm.centered <- scale(PDX_DF20[[i]], center = TRUE, scale = FALSE)
    PDX_DF20 <- AddMetaData(PDX_DF20, DF20.hm.centered, col.name = glue("{i}.centered"))
    
    DF101.hm.centered <- scale(PDX_DF101[[i]], center = TRUE, scale = FALSE)
    PDX_DF101 <- AddMetaData(PDX_DF101, DF101.hm.centered, col.name = glue("{i}.centered"))
    
    DF68.hm.centered <- scale(PDX_DF68[[i]], center = TRUE, scale = FALSE)
    PDX_DF68 <- AddMetaData(PDX_DF68, DF68.hm.centered, col.name = glue("{i}.centered"))
}
```

### STEP 2 DE ANALYSIS #1. VISUALIZING AND QUANTIFYING DE HALLMARK GENESETS
* **QUESTION** Are modules (**OXPHOS and UPR**) differentially expressed across treatment conditions within each model? 
* **HYPOTHESIS** We hypothesize that the **MRD** treatment condition will express enriched levels of OXPHOS and UPR hallmarks relative to cells in the **vehicle** and **relapse** treatment conditions. 
    - **APPROACH #1** Violin Plots and UMAP
        * Center **OXPHOS and UPR** module scores at 0 and reassign to metadata
        * Visualize differences in hallmark score across treatment conditions with:
            - UMAP by treatment vs. UMAP by hallmark score 
            - Violin Plot (center module scores at 0)
        * Statistical test (**wilcoxon rank sum test**): is the difference in hallmark score across treatment conditions statistically significant? 

```{r DEanalysis1 UMAP, message = FALSE, warning = FALSE, fig.align = 'center', fig.width = 20, fig.height = 10}
hms.centered <- c("HALLMARK_OXIDATIVE_PHOSPHORYLATION25.centered", "HALLMARK_UNFOLDED_PROTEIN_RESPONSE33.centered")

#UMAP 
PDX.names <- c("DF20", "DF101", "DF68")
PDXs <- c(PDX_DF20, PDX_DF101, PDX_DF68)
UMAP.plots <- vector("list", length(PDXs))
names(UMAP.plots) <- PDX.names

for (i in 1:length(PDXs)){
  obj <- PDX.names[[i]]
  umap <- UMAPPlot(PDXs[[i]], group.by = "treatment.status") + labs(title = glue("{obj} UMAP by Treatment"))
  p <- FeaturePlot(PDXs[[i]], features = hms.centered, combine = FALSE)
  p[[1]] <- p[[1]] + labs(title = glue("{obj} UMAP by Oxphos Scores"))
  p[[2]] <- p[[2]] + labs(title = glue("{obj} UMAP by UPR Scores"))
  
  UMAP.plots[[obj]] <- umap + p[[1]] + p[[2]]
}

UMAP.plots[["DF20"]]
UMAP.plots[["DF101"]]
UMAP.plots[["DF68"]]
```

* **OBSERVATIONS**
    - Since cells do not separate by treatment status, it is difficult to determine whether there is a correlation between treatment condition and hallmark scores based on our UMAP visualization. 
    - We therefore try to visualize DE hallmark scores across treatment conditions with VlnPlots

```{r DEAnalysis1 VlnPlot, message = FALSE, warning = FALSE, fig.align = 'center', fig.width = 20, fig.height = 10}

#VlnPlot
PDX.names <- c("DF20", "DF101", "DF68")
PDXs <- c(PDX_DF20, PDX_DF101, PDX_DF68)
Vln.plots <- vector("list", length(PDXs))
names(Vln.plots) <- PDX.names

for (i in 1:length(PDXs)){
  obj <- PDX.names[[i]]
  p <- VlnPlot(PDXs[[i]], features = hms.centered, group.by = "treatment.status", combine = F)
  p[[1]] <- p[[1]] + labs(title = glue("OXPHOS scores across treatment in {obj}"), x = obj) + geom_boxplot(show.legend = FALSE)
  p[[2]] <- p[[2]] + labs(title = glue("UPR scores across treatment in {obj}"), x = obj) + geom_boxplot(show.legend = FALSE)
  p <- p[[1]] + p[[2]] + plot_layout(guides= 'collect')
  
  Vln.plots[[obj]] <- p
}

Vln.plots[["DF20"]]
Vln.plots[["DF101"]]
Vln.plots[["DF68"]]
```

* **OBSERVATIONS**
    - We observe differences in hallmark scores across treatment conditions: 
        * Results from **DF20** seems to agree with our hypothesis where cells in **MRD** express enriched levels of **OXPHOS and UPR** genes in comparison to the other two conditions. 
        * Results from **DF101** reveal a clear enrichment of **OXPHOS** genes in the **vehicle** treatment condition, which does not support our hypothesis. It also seems like cells in **relapse** express higher levels of **UPR** genes in comparison to the other two conditions, which also does not support our hypothesis. 
        * Results from **DF68** shows a slight enrichment of **OXPHOS** genes and a slight depltetion of **UPR** genes in **MRD**. 
    - We conduct statistical tests to determine whether the differences in OXPHOS and UPR hallmark scores across treatment conditions are statistically significant. We will be using the **wilcoxon rank sum test**
```{r DEAnalysis1 Statistical test, message = FALSE, warning = FALSE}
#DF20 ---------------------------------
DF20.vehicle <- subset(PDX_DF20, subset = (treatment.status == "vehicle"))
DF20.MRD <- subset(PDX_DF20, subset = (treatment.status == "MRD"))
DF20.relapse <- subset(PDX_DF20, subset = (treatment.status == "relapse"))

DF20.oxphos.df <- data.frame(
  "MRDvsVehicle" = wilcox.test(DF20.MRD$HALLMARK_OXIDATIVE_PHOSPHORYLATION25.centered, DF20.vehicle$HALLMARK_OXIDATIVE_PHOSPHORYLATION25.centered)$p.value, 
  "MRDvsRelapse" = wilcox.test(DF20.MRD$HALLMARK_OXIDATIVE_PHOSPHORYLATION25.centered, DF20.relapse$HALLMARK_OXIDATIVE_PHOSPHORYLATION25.centered)$p.value, 
  "VehiclevsRelapse" = wilcox.test(DF20.vehicle$HALLMARK_OXIDATIVE_PHOSPHORYLATION25.centered, DF20.relapse$HALLMARK_OXIDATIVE_PHOSPHORYLATION25.centered)$p.value
)

DF20.UPR.df <- 
  data.frame(
  "MRDvsVehicle" = wilcox.test(DF20.MRD$HALLMARK_UNFOLDED_PROTEIN_RESPONSE33.centered, DF20.vehicle$HALLMARK_UNFOLDED_PROTEIN_RESPONSE33.centered)$p.value, 
  "MRDvsRelapse" = wilcox.test(DF20.MRD$HALLMARK_UNFOLDED_PROTEIN_RESPONSE33.centered, DF20.relapse$HALLMARK_UNFOLDED_PROTEIN_RESPONSE33.centered)$p.value, 
  "VehiclevsRelapse" = wilcox.test(DF20.vehicle$HALLMARK_UNFOLDED_PROTEIN_RESPONSE33.centered, DF20.relapse$HALLMARK_UNFOLDED_PROTEIN_RESPONSE33.centered)$p.value
)

#DF101 ---------------------------------
DF101.vehicle <- subset(PDX_DF101, subset = (treatment.status == "vehicle"))
DF101.MRD <- subset(PDX_DF101, subset = (treatment.status == "MRD"))
DF101.relapse <- subset(PDX_DF101, subset = (treatment.status == "relapse"))

DF101.oxphos.df <- data.frame(
  "MRDvsVehicle" = wilcox.test(DF101.MRD$HALLMARK_OXIDATIVE_PHOSPHORYLATION25.centered, DF101.vehicle$HALLMARK_OXIDATIVE_PHOSPHORYLATION25.centered)$p.value, 
  "MRDvsRelapse" = wilcox.test(DF101.MRD$HALLMARK_OXIDATIVE_PHOSPHORYLATION25.centered, DF101.relapse$HALLMARK_OXIDATIVE_PHOSPHORYLATION25.centered)$p.value, 
  "VehiclevsRelapse" = wilcox.test(DF101.vehicle$HALLMARK_OXIDATIVE_PHOSPHORYLATION25.centered, DF101.relapse$HALLMARK_OXIDATIVE_PHOSPHORYLATION25.centered)$p.value
)

DF101.UPR.df <- 
  data.frame(
  "MRDvsVehicle" = wilcox.test(DF101.MRD$HALLMARK_UNFOLDED_PROTEIN_RESPONSE33.centered, DF101.vehicle$HALLMARK_UNFOLDED_PROTEIN_RESPONSE33.centered)$p.value, 
  "MRDvsRelapse" = wilcox.test(DF101.MRD$HALLMARK_UNFOLDED_PROTEIN_RESPONSE33.centered, DF101.relapse$HALLMARK_UNFOLDED_PROTEIN_RESPONSE33.centered)$p.value, 
  "VehiclevsRelapse" = wilcox.test(DF101.vehicle$HALLMARK_UNFOLDED_PROTEIN_RESPONSE33.centered, DF101.relapse$HALLMARK_UNFOLDED_PROTEIN_RESPONSE33.centered)$p.value
)

#DF68 ---------------------------------
DF68.vehicle <- subset(PDX_DF68, subset = (treatment.status == "vehicle"))
DF68.MRD <- subset(PDX_DF68, subset = (treatment.status == "MRD"))
DF68.relapse <- subset(PDX_DF68, subset = (treatment.status == "relapse"))

DF68.oxphos.df <- data.frame(
  "MRDvsVehicle" = wilcox.test(DF68.MRD$HALLMARK_OXIDATIVE_PHOSPHORYLATION25.centered, DF68.vehicle$HALLMARK_OXIDATIVE_PHOSPHORYLATION25.centered)$p.value, 
  "MRDvsRelapse" = wilcox.test(DF68.MRD$HALLMARK_OXIDATIVE_PHOSPHORYLATION25.centered, DF68.relapse$HALLMARK_OXIDATIVE_PHOSPHORYLATION25.centered)$p.value, 
  "VehiclevsRelapse" = wilcox.test(DF68.vehicle$HALLMARK_OXIDATIVE_PHOSPHORYLATION25.centered, DF68.relapse$HALLMARK_OXIDATIVE_PHOSPHORYLATION25.centered)$p.value
)

DF68.UPR.df <- 
  data.frame(
  "MRDvsVehicle" = wilcox.test(DF68.MRD$HALLMARK_UNFOLDED_PROTEIN_RESPONSE33.centered, DF68.vehicle$HALLMARK_UNFOLDED_PROTEIN_RESPONSE33.centered)$p.value, 
  "MRDvsRelapse" = wilcox.test(DF68.MRD$HALLMARK_UNFOLDED_PROTEIN_RESPONSE33.centered, DF68.relapse$HALLMARK_UNFOLDED_PROTEIN_RESPONSE33.centered)$p.value, 
  "VehiclevsRelapse" = wilcox.test(DF68.vehicle$HALLMARK_UNFOLDED_PROTEIN_RESPONSE33.centered, DF68.relapse$HALLMARK_UNFOLDED_PROTEIN_RESPONSE33.centered)$p.value
)

#combine ------------------------------
oxphos.DF <- rbind(DF20.oxphos.df, DF101.oxphos.df, DF68.oxphos.df)
rownames(oxphos.DF) <- c("OXPHOS.DF20", "OXPHOS.DF101", "OXPHOS.DF68")

UPR.DF <- rbind(DF20.UPR.df, DF101.UPR.df, DF68.UPR.df)
rownames(UPR.DF) <- c("UPR.DF20", "UPR.DF101", "UPR.DF68")

all.DF <- rbind(oxphos.DF, UPR.DF)
DT::datatable(all.DF) %>% 
  DT::formatRound(names(all.DF), digits = 7) %>% 
  DT::formatStyle(names(all.DF), color = DT::styleInterval(0.05, c('red', 'black')))
```

* **CONCLUSIONS**
    - The only statistically significant difference (p < 0.05) in hallmark scores is between treatment conditions within model **DF101** for **OXPHOS**. 
    - However, the trends found in DF101 do not agree with our hypothesis that cells in **MRD** differentially enrich OXPHOS genes) : 
        * DF101 reveals an enrichment of **OXPHOS** genes in the **vehicle** treatment condition in comparison to the other two conditions. 

Considering how our approach with analyzing and comparing **module score** did not support our hypothesis, we now try to confirm our results with our second appraoch: **GSEA**

* **APPROACH #2** Geneset Enrichment Analysis (GSEA)
      - GSEA enrichment plots for hallmarks of interest between condition.1 vs. condition.2 
      - Rank genes and compute GSEA enrichment scores 
      - Statistical test: how significant are the enrichment scores? 

**have not finalized the statistical test and ranking method to use for GSEA**

```{r GSEA, message=FALSE, warning=FALSE, fig.align='center', fig.width=20, fig.height=10,}

```

### STEP 3 DE ANALYSIS #2. IDENTIFYING INDIVIDUAL DE GENES

**have not finalized the statistical test to use for volcano plot (need to match what we use for GSEA)**

### STEP 4 CELL CYCLE ANALYSIS
* **QUESTION #1** Are more cells at different cell cycle phases between treatments? 
* **HYPOTHESIS** There should be less cycling cells in the MRD treatment conditions relative to the other two treatment conditions. 
    - **APPROACH** Violin Plots and UMAP 
        * Visualize differences in cell cycle score across treatment conditions with: 
            - Stacked Bar Plot
            - UMAP by treatment vs. UMAP by cell cycle phase vs. UMAP by S and G2M Score 
            - Violin Plot 
        * Statistical test: is the difference in cell cycle scores across treatment conditions statistically
        
```{r cellcycle1, message = FALSE, warning = FALSE, fig.align = 'center', fig.width = 20, fig.height = 10}
#Stacked Bar Plot --------------------------
bar.plots <- vector("list", length = 3)
names(bar.plots) <- c("DF20", "DF101", "DF68")

for(i in 1:length(PDXs)){
  obj = PDXs[[i]]
  name = PDX.names[[i]]
  
  obj$Phase <- factor(obj$Phase, levels = c("G1", "S", "G2M"))
  p = table(obj$Phase, obj$treatment.status) %>% 
    as.data.frame() %>% 
    rename(Phase = Var1, Treatment = Var2, numCells = Freq) %>% 
    ggplot(aes(x=Treatment, y=numCells, fill=Phase)) + 
    geom_bar(position="fill", stat="identity") + 
    labs(title = glue("{name} % of Malignant Cells in Each Cell Cycle Phase/ Treatment")) +
    theme(plot.title = element_text(size = 8))
  bar.plots[[i]] <- p
}

bar.plots[["DF20"]] + bar.plots[["DF101"]] + bar.plots[["DF68"]] + plot_layout(guides= 'collect')
```

* **OBSERVATIONS**
    - Different fractions of cells do seem to be in different cell cycle phases depending on the treatment condition. However, the trends we see do not necessary agree with our hypothesis that the **MRD** treatment condition will have the least number of cycling cells: 
        * In DF68, cells in **vehicle** do seem to have the most cells in the G2M phase, although cells in **relapse** has the most noncycling cells in the G1 phase. 
        * In DF20, cells in **MRD** have the highest fraction of cells in the G2M phase, which does not support our hypothesis. 
        * In all cases, cells in **MRD** do not seem to have the highest fraction of noncycling G1 cells. 

To investigate this further, we compare UMAPs by treatment vs. UMAPs by cell cycle phase and scores

```{r cellcyleUMAP, message=FALSE, warning = FALSE, fig.align = 'center', fig.width=20, fig.height=10}
#center cell cycle scores 
cc.scores <- names(PDX_All@meta.data)[43:44]
for(i in cc.scores){
    DF20.score.centered <- scale(PDX_DF20[[i]], center = TRUE, scale = FALSE)
    PDX_DF20 <- AddMetaData(PDX_DF20, DF20.score.centered, col.name = glue("{i}.centered"))
    
    DF101.score.centered <- scale(PDX_DF101[[i]], center = TRUE, scale = FALSE)
    PDX_DF101 <- AddMetaData(PDX_DF101, DF101.score.centered, col.name = glue("{i}.centered"))
    
    DF68.score.centered <- scale(PDX_DF68[[i]], center = TRUE, scale = FALSE)
    PDX_DF68 <- AddMetaData(PDX_DF68, DF68.score.centered, col.name = glue("{i}.centered"))
}

#UMAP -------------------------
ccUMAP.plots <- vector("list", length = 3)
names(ccUMAP.plots) <- c("DF20", "DF101", "DF68")

for(i in 1:length(PDXs)){
  obj = PDXs[[i]]
  name = PDX.names[[i]]
  obj$Phase <- factor(obj$Phase, levels = c("G1", "S", "G2M"))
  
  byt <- UMAPPlot(obj, group.by = "treatment.status") + 
    labs(title = glue("{name} UMAP by treatment")) + 
    theme(plot.title = element_text(hjust = 0.5))
  bycc <- UMAPPlot(obj, group.by = "Phase") + 
    labs(title = glue("{name} UMAP by CellCycle Phase")) + 
    theme(plot.title = element_text(hjust = 0.5))
  byscore <- FeaturePlot(obj, features = c("S.Score.centered", "G2M.Score.centered"), combine = FALSE)
  byscore[[1]] <- byscore[[1]] + labs(title = glue("{name} UMAP by S Score"))
  byscore[[2]] <- byscore[[2]] + labs(title = glue("{name} UMAP by G2M Score"))
  
  ccUMAP.plots[[i]] <- byt + bycc + byscore[[1]] + byscore[[2]]
}

ccUMAP.plots[["DF20"]]
ccUMAP.plots[["DF101"]]
ccUMAP.plots[["DF68"]]
```

* **OBSERVATIONS** 
    - Although it doesn't seem like the cells cluster by treatment condition within each model, it does seem like cells in the same cell cycle phase, especially those with higher S or G2M scores tend to be near each other on UMAP space.
    - However, it is hard to tell whether there is a correlation between treatment condition and cell cycle phase / score based on these plots. We further investigate this by building violin plots of cell cycle scores separated by treatment status. 
    
```{r cellcycleVln, message=FALSE, warning=FALSE, fig.align='center', fig.width=20, fig.height=10}
#VlnPlot -------------------------
cc.Vln.plots <- vector("list", length = 3)
names(Vln.plots) <- PDX.names

for (i in 1:length(PDXs)){
  obj <- PDX.names[[i]]
  p <- VlnPlot(PDXs[[i]], features = c("S.Score.centered", "G2M.Score.centered"), group.by = "treatment.status", combine = F)
  p[[1]] <- p[[1]] + labs(title = glue("{obj} S.Score across treatment"), x = obj) + geom_boxplot(show.legend = FALSE)
  p[[2]] <- p[[2]] + labs(title = glue("{obj} G2M.Score across treatment"), x = obj) + geom_boxplot(show.legend = FALSE)

  cc.Vln.plots[[obj]] <- p[[1]] + p[[2]] + plot_layout(guides= 'collect')
}

cc.Vln.plots[["DF20"]]
cc.Vln.plots[["DF101"]]
cc.Vln.plots[["DF68"]]
```

* **OBSERVATIONS** 
    - There does not seem to be any obvious trends in S or G2M scores across treatment conditions other than in DF20, which shows an increase of cells in the S or G2M phases in **MRD and relapse** relative to **vehicle**. To confirm this, we evaluate the statistical significance of the difference between S and G2M scores across treatment conditions. 

```{r cellcycleStat, message=FALSE, warning=FALSE}
#DF20 ---------------------------------
DF20.sscore.df <- data.frame(
  "MRDvsVehicle" = wilcox.test(DF20.MRD$S.Score.centered, DF20.vehicle$S.Score.centered)$p.value, 
  "MRDvsRelapse" = wilcox.test(DF20.MRD$S.Score.centered, DF20.relapse$S.Score.centered)$p.value, 
  "VehiclevsRelapse" = wilcox.test(DF20.vehicle$S.Score.centered, DF20.relapse$S.Score.centered)$p.value
)

DF20.g2m.df <- 
  data.frame(
  "MRDvsVehicle" = wilcox.test(DF20.MRD$G2M.Score.centered, DF20.vehicle$G2M.Score.centered)$p.value, 
  "MRDvsRelapse" = wilcox.test(DF20.MRD$G2M.Score.centered, DF20.relapse$G2M.Score.centered)$p.value, 
  "VehiclevsRelapse" = wilcox.test(DF20.vehicle$G2M.Score.centered, DF20.relapse$G2M.Score.centered)$p.value
)

#DF101 ---------------------------------
DF101.sscore.df <- data.frame(
  "MRDvsVehicle" = wilcox.test(DF101.MRD$S.Score.centered, DF101.vehicle$S.Score.centered)$p.value, 
  "MRDvsRelapse" = wilcox.test(DF101.MRD$S.Score.centered, DF101.relapse$S.Score.centered)$p.value, 
  "VehiclevsRelapse" = wilcox.test(DF101.vehicle$S.Score.centered, DF101.relapse$S.Score.centered)$p.value
)

DF101.g2m.df <- 
  data.frame(
  "MRDvsVehicle" = wilcox.test(DF101.MRD$G2M.Score.centered, DF101.vehicle$G2M.Score.centered)$p.value, 
  "MRDvsRelapse" = wilcox.test(DF101.MRD$G2M.Score.centered, DF101.relapse$G2M.Score.centered)$p.value, 
  "VehiclevsRelapse" = wilcox.test(DF101.vehicle$G2M.Score.centered, DF101.relapse$G2M.Score.centered)$p.value
)

#DF68 ---------------------------------
DF68.sscore.df <- data.frame(
  "MRDvsVehicle" = wilcox.test(DF68.MRD$S.Score.centered, DF68.vehicle$S.Score.centered)$p.value, 
  "MRDvsRelapse" = wilcox.test(DF68.MRD$S.Score.centered, DF68.relapse$S.Score.centered)$p.value, 
  "VehiclevsRelapse" = wilcox.test(DF68.vehicle$S.Score.centered, DF68.relapse$S.Score.centered)$p.value
)

DF68.g2m.df <- 
  data.frame(
  "MRDvsVehicle" = wilcox.test(DF68.MRD$G2M.Score.centered, DF68.vehicle$G2M.Score.centered)$p.value, 
  "MRDvsRelapse" = wilcox.test(DF68.MRD$G2M.Score.centered, DF68.relapse$G2M.Score.centered)$p.value, 
  "VehiclevsRelapse" = wilcox.test(DF68.vehicle$G2M.Score.centered, DF68.relapse$G2M.Score.centered)$p.value
)

#combine ------------------------------
sscore.DF <- rbind(DF20.sscore.df, DF101.sscore.df, DF68.sscore.df)
rownames(sscore.DF) <- c("S.DF20", "S.DF101", "S.DF68")

g2m.DF <- rbind(DF20.g2m.df, DF101.g2m.df, DF68.g2m.df)
rownames(g2m.DF) <- c("g2m.DF20", "g2m.DF101", "g2m.DF68")

both.DF <- rbind(sscore.DF, g2m.DF)
DT::datatable(both.DF) %>% 
  DT::formatRound(names(both.DF), digits = 7) %>% 
  DT::formatStyle(names(both.DF), color = DT::styleInterval(0.05, c('red', 'black')))
```

* **CONCLUSIONS** 
    - Only comparisons between **MRD vs. Vehicle** and **Vehicle vs. Relapse** in model **DF20** yielded statistically significant results (p < 0.05) for both **S and G2M scores** 
    - Based on the Violin Plot: 
        * **MRD** and **Relapse** have higher S and G2M scores than **Vehicle** 
    - These results do not support our hypothesis that there should be less cycling cells in the **MRD** treatment condition than the other two treatment conditions. 

Our results from above collectively suggest that **OXPHOS and UPR** expression, and cell cycle phase, do not significantly correlate with the treatment condition a cell is in. However, considering the idea that cell cycle might influence the expression of signatures, we are now interested in examining whether there is a correlation between the expression of **OXPHOS and UPR** and **cell cycle phase**. 

* **QUESTION #2** Is there a correlation between cell cycle phase and the expression of signatures? 
* **HYPOTHESIS** We hypothesize that low cycling cells should express higher levels of **OXPHOS and UPR** genes. 
    - **APPROACH** Violin Plots
        * Create Violin Plots for each hallmark of interest like **DE Analysis #1/APPROACH #1. 
        * Stratify the Violin Plots by cell cycle phase
        
```{r cellcycle2, message=FALSE, warning=FALSE, fig.align='center', fig.width=20, fig.height = 15}
cc.exp.plot <- vector("list", length = 3)
names(cc.exp.plot) <- PDX.names

for(i in 1:length(PDXs)){
  obj = PDXs[[i]]
  name = PDX.names[[i]]
  obj$Phase <- factor(obj$Phase, levels = c("G1", "S", "G2M"))
  
  p1 <- VlnPlot(obj, features = hms.centered, group.by = "Phase", combine = FALSE)
  p1[[1]] <- p1[[1]] + labs(title = glue("{name} OXPHOS score by Cell Cycle Phase")) + geom_boxplot(show.legend = FALSE)
  p1[[2]] <- p1[[2]] + labs(title = glue("{name} UPR score by Cell Cycle Phase")) + geom_boxplot(show.legend = FALSE)

  
  p2 <- VlnPlot(obj, features = hms.centered, group.by = "treatment.status", split.by = "Phase", combine = FALSE)
  p2[[1]] <- p2[[1]] + labs(title = glue("{name} OXPHOS score by Cell Cycle Phase / Treatment")) + geom_boxplot(show.legend = FALSE)
  p2[[2]] <- p2[[2]] + labs(title = glue("{name} UPR score by Cell Cycle Phase / Treatment")) + geom_boxplot(show.legend = FALSE)
  
  cc.exp.plot[[i]] <- p1[[1]] + p1[[2]] + p2[[1]]+ p2[[2]] + plot_layout(guides= 'collect')
}

cc.exp.plot[["DF20"]]
cc.exp.plot[["DF101"]]
cc.exp.plot[["DF68"]]
```

* **OBSERVATIONS** 
    - Results from comparing **OXPHOS and UPR** gene scores across cell cycle phase within each model reveal that cells in S phase seems to generally express higher levels of **OXPHOS** genes in comparison to the other cell cycle phases. This trend is especially true within **DF20**. 
    - Results from comparing **OXPHOS and UPR** gene scores across cell cycle phase within each treatment condition per model revel that cells in S phase generally enrich **OXPHOS** genes especially within the **MRD** treatment condition in comparison to the other cell cycle phases. 
    - We hypothesized that low / cycling cells (G1) should express higher levels of **OXPHOS and/or UPR genes** based on findings from previous studies that low cycling cells seem to survive cancer treatment. We try to rationalize our findings that cells in S phase actually express higher levels of our hallmarks: 
        * Even if a cell is found to be in the S phase, it might not mean that it is cycling. It is possible that the cell entered the S phase but does not continue through the cell cycle upon treatment. Furthermore, since DNA is being replicated during S phase and mistakes are most commonly made during DNA replication, it is possible that DNA damage repair programs are also most active during this phase, and may subsequently promote survival upon cancer treatment. 
        * This rationale would also agree with our results from comparing S and G2M scores across treatment conditions, where we found that cells in **MRD and Relapse** have statistically significantly higher S and G2M scores than cells in **Vehicle**. 
        * However, this rationale does not explain why we do not see enriched expression of OXPHOS and UPR genes in the MRD treatment condition. 

        





