---
title: "Izar 2020 PDX (Cohort 3) DE Analysis"
author: "Jesslyn Goh and Mike Cuoco"
date: "7/20/2020"
editor_options: 
  chunk_output_type: console
output:
 html_document:
  code_folding: hide
  toc_float: TRUE 
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = here::here())
knitr::opts_chunk$set(echo = TRUE)
options(width=120)
```

# IZAR 2020 PDX (COHORT 3) DATA DIFFERANTIAL EXPRESISON ANALYSIS 

### OVERVIEW
* The PDX data from Izar2020 consists of **only Malignant cells** from three HGSOC PDX models derived from patients with different treatment histories were selected for implantation: 
    - DF20 (BRCA WT treatment-naive, clinically platinum sensitive) 
    - DF101 (BRCA1 mutant, 2 lines of prior therapy, clinically platinum resistant) 
    - DF68 (BRCA1 mutant, 6 lines of prior therapy, clinically platinum resistant) 
* After tumors were established, animals were divided into two groups per model: 
    - Vehicle (treated with DMSO) 
    - Carboplatin (treated with IP carboplatin) 
        * Carboplatin-treated mice for minimal residual disease (**MRD**) group were harvested for scRNA-seq 
        * The remaining carboplatin-treated mice were harvested at endpoint (**vehicle**)
* In our 5-part analysis of the Izar 2020 PDX data, we are interested in identifying differentially expressed genes and hallmark genesets between treatment statuses **within** each model. 
* We split our PDX analysis into three parts: 
    1. **Load Data and Create PDX Seurat Object**
        a. The code to this part of our analysis is in the **read_Izar_2020.R** file in the **code** folder. During this part of our analysis we: 
            i. Load in PDX count matrix and Create Seurat Object 
            ii. Assign Metadata identities including: 
                - Mouse ID 
                - Model ID 
                - Treatment Status 
            iii. Score cells for **cell cycle** and **hallmark genesets** 
                - **Note**: It does not matter whether we call AddModuleScore before or after subsetting and scaling each model because AddModuleScore uses the **data** slot. 
            iv. Save Seurat Object
          
    2. **Process Data and Exploratory Data Analysis**
        a. The code to this part of our analysis can be found in the **03_Izar2020_PDX_Load** file in the **old/edited** folder. During this part of our analysis we: 
            i. Load in PDX Seurat Object from Part 1 and **subset by model**. Continue analysis separately for each model. 
            ii. Scale and FindVariableFeatures (prepares data for dimensionality reduction)
            iii. Dimensionality Reduction (PCA + UMAP)
            vi. Save Seurat Objects 
      
    3. **Exploratory Data Analysis**
        a. The code to this part of our analysis can be found in the **03.0_Izar2020_PDX_Exploratory Analysis** file in the **analysis** folder. During this part of our analysis we: 
            i. Load in PDX Seurat Object from Part 2. Analyze separately for each model.
            ii. Compute summary metrics for PDX data such as: 
                - Number of cells per model per treatment 
                - Number of cells per treatment per cell cycle phase 
            iii. Visualize how cells separate based on metadata identities via UMAP 
                - Intermodel heterogeneity: How do cells separate by model? 
                - Intramodel heterogeneity: 
                    * How do cells separate by treament status? 
                    * How do cells separate by cell cycle phase?
            
    4. **DE Analysis**
        a. **TYPE #1 DE ANALYSIS**: Visualizing and Quantifying Differentially Expression on **Predefined** GO Genesets
            i. Violin Plots and UMAP 
            ii. Gene Set Enrichment Analysis (GSEA)
        b. **TYPE #2 DE ANALYSIS**: Finding DE Genes from scratch
            i. Volcano Plots 
          
    5. **CELL CYCLE ANALYSIS**
        a. **TYPE #1 CELL CYCLE ANALYSIS**: Examine correlation between treatment condition and cell cycle phase
        b. Evaluate the idea that cell cycle might influence expression of signatures 

This is the fourth part of our 5-part analysis of the Izar 2020 PDX (Cohort 3) data. 

### DIFFERENTIAL EXPRESSION ANALYSIS IN-DEPTH EXPLANATION 
We are interested in answering a few questions for our DE Analysis:

**DE ANALYSIS #1. Visualizing and Quantifying DE Hallmark Genesets** 

* **QUESTION** Are modules (**OXPHOS and UPR**) differentially expressed across treatment conditions within each model? 
    - **APPROACH #1** Violin Plots and UMAP 
        * Visualize differences in hallmark score across treatment conditions with: 
            - Violin Plot 
            - UMAP by treatment vs. UMAP by hallmark score 
        * Statistical test: is the difference in hallmark score across treatment conditions statistically significant? 
    - **APPROACH #2** GSEA
        * GSEA enrichment plots for hallmarks of interest between condition.1 vs. condition.2 
        * Rank genes and compute GSEA enrichment scores 
        * Statistical test: how significant are the enrichment scores? 

**DE ANALYSIS #2. Identifying Individual DE Genes** 

### STEP 1 LOAD IN SEURAT OBJECTS AND GENESETS 

```{r init, message = FALSE, warning = FALSE}
# Load packages
source(here::here('packages.R'))

#Read in PDX RDS object
PDX_All = readRDS("data/Izar_2020/test/jesslyn_PDX_All_processed.RDS")
PDX_DF20 = readRDS("data/Izar_2020/test/jesslyn_PDX_DF20_processed.RDS")
PDX_DF101 = readRDS("data/Izar_2020/test/jesslyn_PDX_DF101_processed.RDS")
PDX_DF68 = readRDS("data/Izar_2020/test/jesslyn_PDX_DF68_processed.RDS")

#Read in hallmarks of interest
hallmark_names = read_lines("data/gene_lists/hallmarks.txt")
hallmark.list <- vector(mode = "list", length = length(hallmark_names) + 4)
names(hallmark.list) <- c(hallmark_names, "GO.OXPHOS", "KEGG.OXPHOS", "UNUPDATED.OXPHOS", "UNUPDATED.UPR")
for(hm in hallmark_names){
  file <- read_lines(glue("data/gene_lists/hallmarks/{hm}_updated.txt"), skip = 1)
  hallmark.list[[hm]] <- file
}

hallmark.list[["GO.OXPHOS"]] <- read_lines("data/gene_lists/extra/GO.OXPHOS.txt", skip = 1)
hallmark.list[["KEGG.OXPHOS"]] <- read_lines("data/gene_lists/extra/KEGG.OXPHOS.txt", skip = 2)
hallmark.list[["UNUPDATED.OXPHOS"]] <- read_lines("data/gene_lists/oxphos.txt", skip =1)
hallmark.list[["UNUPDATED.UPR"]] <- read_lines("data/gene_lists/upr.txt", skip =1)

#center module and cell cycle scores and reassign to the metadata of each Seurat object
hm.names <- names(PDX_All@meta.data)[9:48]

for(i in hm.names){
    DF20.hm.centered <- scale(PDX_DF20[[i]], center = TRUE, scale = FALSE)
    PDX_DF20 <- AddMetaData(PDX_DF20, DF20.hm.centered, col.name = glue("{i}.centered"))
    
    DF101.hm.centered <- scale(PDX_DF101[[i]], center = TRUE, scale = FALSE)
    PDX_DF101 <- AddMetaData(PDX_DF101, DF101.hm.centered, col.name = glue("{i}.centered"))
    
    DF68.hm.centered <- scale(PDX_DF68[[i]], center = TRUE, scale = FALSE)
    PDX_DF68 <- AddMetaData(PDX_DF68, DF68.hm.centered, col.name = glue("{i}.centered"))
}
```

### STEP 2 DETERMINING OXPHOS GENESET 
* Before doing any DE Analysis on the genesets, we investigate which of the three OXPHOS genesets we found is best for our data. The **OXPHOS** genesets that we test are listed as follows:
    - Unupdated Version of HALLMARK_OXIDATIVE_PHOSPHORYLATION (200 genes) https://www.gsea-msigdb.org/gsea/msigdb/cards/HALLMARK_OXIDATIVE_PHOSPHORYLATION
    - Updated Version of HALLMARK_OXIDATIVE_PHOSPHORYLATION (200 genes) https://www.gsea-msigdb.org/gsea/msigdb/cards/HALLMARK_OXIDATIVE_PHOSPHORYLATION
    - GO_OXIDATIVE_PHOSPHORYLATION (144 genes) https://www.gsea-msigdb.org/gsea/msigdb/cards/GO_OXIDATIVE_PHOSPHORYLATION 
    - KEGG_OXIDATIVE_PHOSPHORYLATION (131 genes) https://www.gsea-msigdb.org/gsea/msigdb/cards/KEGG_OXIDATIVE_PHOSPHORYLATION

* We determine which geneset to use by asking these questions: 
    1. How many and which genes are not found in the PDX Seurat Object for each geneset?
    2. What makes the genesets different from each other? Which genes in each geneset are the most DE? Are they the same?
        a. Volcano Plot: Call FindMarkers on each model, plot onto Volcano Plot, label cells by each geneset 
    3. Which geneset gives us the most statistically significant results? 
        a. VlnPlot: Plot hallmark scores for each geneset group by treatment status for each model 
        b. Label the VlnPlot with the number of genes found in PDX Seurat Object for each OXPHOS geneset 
  
**ANSWERING QUESTION #1: How many and which genes are not found in the PDX Seurat Object for each geneset?**

```{r oxphosMetrics, message=FALSE, warning=FALSE, fig.align = 'center'}

hm.length.df <- data.frame(
  "UNUPDATED.HM.OXPHOS" = length(hallmark.list[["UNUPDATED.OXPHOS"]]),
  "HALLMARK.OXPHOS" = length(hallmark.list[["HALLMARK_OXIDATIVE_PHOSPHORYLATION"]]), 
  "GO.OXPHOS" = length(hallmark.list[["GO.OXPHOS"]]), 
  "KEGG.OXPHOS" = length(hallmark.list[["KEGG.OXPHOS"]])
)

Found.df <- data.frame(
  "UNUPDATED.HM.OXPHOS" = sum((hallmark.list[["UNUPDATED.OXPHOS"]] %in% rownames(PDX_All))),
  "HALLMARK.OXPHOS" = sum((hallmark.list[["HALLMARK_OXIDATIVE_PHOSPHORYLATION"]] %in% rownames(PDX_All))), 
  "GO.OXPHOS" = sum((hallmark.list[["GO.OXPHOS"]] %in% rownames(PDX_All))), 
  "KEGG.OXPHOS" = sum((hallmark.list[["KEGG.OXPHOS"]] %in% rownames(PDX_All)))
)

PFound.df <- data.frame(
  "UNUPDATED.HM.OXPHOS" = (sum((hallmark.list[["UNUPDATED.OXPHOS"]] %in% rownames(PDX_All)))/length(hallmark.list[["UNUPDATED.OXPHOS"]]))*100,
  "HALLMARK.OXPHOS" = (sum((hallmark.list[["HALLMARK_OXIDATIVE_PHOSPHORYLATION"]] %in% rownames(PDX_All)))/length(hallmark.list[["HALLMARK_OXIDATIVE_PHOSPHORYLATION"]])) * 100, 
  "GO.OXPHOS" = (sum((hallmark.list[["GO.OXPHOS"]] %in% rownames(PDX_All)))/length(hallmark.list[["GO.OXPHOS"]]))*100, 
  "KEGG.OXPHOS" = (sum((hallmark.list[["KEGG.OXPHOS"]] %in% rownames(PDX_All)))/length(hallmark.list[["KEGG.OXPHOS"]]))*100
)

NA.df <- data.frame(
  "UNUPDATED.HM.OXPHOS" = sum(!(hallmark.list[["UNUPDATED.OXPHOS"]] %in% rownames(PDX_All))),
  "HALLMARK.OXPHOS" = sum(!(hallmark.list[["HALLMARK_OXIDATIVE_PHOSPHORYLATION"]] %in% rownames(PDX_All))), 
  "GO.OXPHOS" = sum(!(hallmark.list[["GO.OXPHOS"]] %in% rownames(PDX_All))), 
  "KEGG.OXPHOS" = sum(!(hallmark.list[["KEGG.OXPHOS"]] %in% rownames(PDX_All)))
)

all.df <- rbind(hm.length.df, Found.df, PFound.df, NA.df)
rownames(all.df) <- c("NumGenes", "Found", "%Found", "Not Found")
all.df[,"GO.OXPHOS"] <- round(all.df[,"GO.OXPHOS"])
all.df[,"KEGG.OXPHOS"] <- round(all.df[,"KEGG.OXPHOS"])
all.df

# IDENTIFY GENES THAT ARE NOT FOUND -----------
NA.genes.df <- vector("list", length = 4)
names <- c("UNUPDATED.OXPHOS", "HALLMARK_OXIDATIVE_PHOSPHORYLATION", "GO.OXPHOS", "KEGG.OXPHOS")
names(NA.genes.df) <- names
for(i in names){
  NA.genes.df[[i]] <- (hallmark.list[[i]])[which(!(hallmark.list[[i]] %in% rownames(PDX_All)))]
}
```

* **OBSERVATIONS** 
    - The **UNUPDATED.OXPHOS** genelist actually has the most genes that are found in the PDX object. 
    - Tried to delete the dash (-) in genes with ("MT-"), and also tried to delete the MT part of the gene name, but those genes remain not to be found in the PDX object. 
    - Since the **UNUPDATED.OXPHOS** genelist works better than the updated version (HALLMARK_OXIDATIVE_PHOSPHORYLATION), we will continue our comparison of genesets using the UNUPDATED version. 
* We now wonder: 
    - Which OXPHOS genes are the most DE if we call FindMarkers? 
    - Which geneset(s) do the most DE OXPHOS genes belong to? 

**ANSWERING QUESTION 2: Which genes in each geneset are the most DE? Are they the same?**
* Used the **wilcoxon rank sum test** for FindMarkers 

```{r oxphosVolcano, message=FALSE, warning=FALSE, fig.align='center', fig.width=15, fig.height=8}
PDXs <- c(PDX_DF20, PDX_DF101, PDX_DF68)
PDX.names <- c("DF20", "DF101", "DF68")
oxphos.hm <- c("UNUPDATED.OXPHOS", "GO.OXPHOS", "KEGG.OXPHOS")
PDX.hm.plots <- vector("list", length = 3)
names(PDX.hm.plots) <- PDX.names

markers <- vector("list", length = 3)
names(markers) <- PDX.names

markers[["DF20"]] <- FindMarkers(PDX_DF20, group.by = "treatment.status", ident.1 = "MRD", ident.2 = "vehicle", test.use = "wilcox", logfc.threshold = 0) 
markers[["DF101"]]  <- FindMarkers(PDX_DF101, group.by = "treatment.status", ident.1 = "MRD", ident.2 = "vehicle", test.use = "wilcox", logfc.threshold = 0)
markers[["DF68"]] <- FindMarkers(PDX_DF68, group.by = "treatment.status", ident.1 = "MRD", ident.2 = "vehicle", test.use = "wilcox", logfc.threshold = 0)

for(i in 1:length(PDXs)){
    PDX <- PDX.names[[i]]
    DF.hm.plot <- vector("list", length = 3)
    names(DF.hm.plot) <- oxphos.hm
    marker <- markers[[PDX]]
    
  for(oxphos in oxphos.hm){
    
    avgLFC <- rownames(marker[which(abs(marker$avg_logFC) > 0.5),])
    keyvals <- ifelse(!(rownames(marker) %in% hallmark.list[[oxphos]]), 'black', 
                      ifelse(abs(marker$avg_logFC) > 0.5, 'red', 'grey'))
    names(keyvals)[keyvals == 'red'] <- 'OXPHOS & logFC'
    names(keyvals)[keyvals == 'grey'] <- 'OXPHOS x logFC'
    names(keyvals)[keyvals == 'black'] <- 'Not OXPHOS'
    found = length(avgLFC[which(avgLFC %in% hallmark.list[[oxphos]])])
    
    p <- EnhancedVolcano(marker, 
           lab = rownames(marker),
           selectLab = avgLFC[which(avgLFC %in% hallmark.list[[oxphos]])],
           labCol = "red",
           x='avg_logFC', y='p_val_adj', pCutoff = 0.05, 
           FCcutoff = 0.5, 
           colCustom = keyvals,
           pointSize = c(ifelse(rownames(marker) %in% hallmark.list[[oxphos]], 2.5,1)), 
           drawConnectors = TRUE,
           boxedLabels = TRUE,
           labvjust = 1,
           title= glue("{PDX.names[[i]]} MRD vs. vehicle DE Genes"), subtitle= "LogFC cutoff: 0.5, p cutoff: 0.05",
           caption = glue("{oxphos}: {found} DE oxphos genes found")
           )
    DF.hm.plot[[oxphos]] <- p
  }
  
  PDX.hm.plots[[PDX]] <- DF.hm.plot[["UNUPDATED.OXPHOS"]] + DF.hm.plot[["GO.OXPHOS"]] + DF.hm.plot[["KEGG.OXPHOS"]]
}

PDX.hm.plots[["DF20"]]
PDX.hm.plots[["DF101"]]
PDX.hm.plots[["DF68"]]

# number of DE oxphos genes found in each geneset within each model -----------------
DF20.de.df <- data.frame(
  "UNUPDATED.OXPHOS" = sum(rownames(markers[["DF20"]][which(abs(markers[["DF20"]]$avg_logFC) > 0.5),]) %in% hallmark.list[["UNUPDATED.OXPHOS"]]),
  "GO.OXPHOS" = sum(rownames(markers[["DF20"]][which(abs(markers[["DF20"]]$avg_logFC) > 0.5),]) %in% hallmark.list[["GO.OXPHOS"]]),
  "KEGG.OXPHOS" = sum(rownames(markers[["DF20"]][which(abs(markers[["DF20"]]$avg_logFC) > 0.5),]) %in% hallmark.list[["KEGG.OXPHOS"]])
)

DF101.de.df <- data.frame(
  "UNUPDATED.OXPHOS" = sum(rownames(markers[["DF101"]][which(abs(markers[["DF101"]]$avg_logFC) > 0.5),]) %in% hallmark.list[["UNUPDATED.OXPHOS"]]),
  "GO.OXPHOS" = sum(rownames(markers[["DF101"]][which(abs(markers[["DF101"]]$avg_logFC) > 0.5),]) %in% hallmark.list[["GO.OXPHOS"]]),
  "KEGG.OXPHOS" = sum(rownames(markers[["DF101"]][which(abs(markers[["DF101"]]$avg_logFC) > 0.5),]) %in% hallmark.list[["KEGG.OXPHOS"]])
)

DF68.de.df <- data.frame(
  "UNUPDATED.OXPHOS" = sum(rownames(markers[["DF68"]][which(abs(markers[["DF68"]]$avg_logFC) > 0.5),]) %in% hallmark.list[["UNUPDATED.OXPHOS"]]),
  "GO.OXPHOS" = sum(rownames(markers[["DF68"]][which(abs(markers[["DF68"]]$avg_logFC) > 0.5),]) %in% hallmark.list[["GO.OXPHOS"]]),
  "KEGG.OXPHOS" = sum(rownames(markers[["DF68"]][which(abs(markers[["DF68"]]$avg_logFC) > 0.5),]) %in% hallmark.list[["KEGG.OXPHOS"]])
)

all.de.df <- rbind(DF20.de.df, DF101.de.df, DF68.de.df)
rownames(all.de.df) <- c("DF20.MRDvVehicle", "DF101.MRDvVehicle", "DF68.MRDvVehicle")
all.de.df
```

* **OBSERVATIONS** 
    - **UNUPDATED.OXPHOS** geneset consistently has the most DE OXPHOS genes (logFC > 0.5, regardless of padj) relative to the other two genesets for each model comparison between MRD and vehicle. 
    - Although it seems that there are a few overlaps of OXPHOS genes found, it is hard to visualize which ones overlap, and whether the ones that overlap are among the top DE genes (regardless of padj). We therefore create tables for each model, extract the top 5 DE genes from each geneset and compare if they're the same. 

```{r oxphosGSDE, message=FALSE, warning=FALSE}

DF20.top5 <- markers[["DF20"]] %>% arrange(-abs(avg_logFC))
DF101.top5 <- markers[["DF101"]] %>% arrange(-abs(avg_logFC))
DF68.top5 <- markers[["DF68"]] %>% arrange(-abs(avg_logFC))

DF20.gs.df <- data.frame(
  "UNUPDATED.OXPHOS.DE" = head(rownames(DF20.top5)[which(rownames(DF20.top5) %in% hallmark.list[["UNUPDATED.OXPHOS"]])], 5), 
  "UNUPDATED.OXPHOS" = select(DF20.top5[head(rownames(DF20.top5)[which(rownames(DF20.top5) %in% hallmark.list[["UNUPDATED.OXPHOS"]])], 5),], avg_logFC),
  "GO.OXPHOS.DE" = head(rownames(DF20.top5)[which(rownames(DF20.top5) %in% hallmark.list[["GO.OXPHOS"]])], 5),
  "GO.OXPHOS" = select(DF20.top5[head(rownames(DF20.top5)[which(rownames(DF20.top5) %in% hallmark.list[["GO.OXPHOS"]])], 5),], avg_logFC),
  "KEGG.OXPHOS.DE" = head(rownames(DF20.top5)[which(rownames(DF20.top5) %in% hallmark.list[["KEGG.OXPHOS"]])], 5), 
  "KEGG.OXPHOS" = select(DF20.top5[head(rownames(DF20.top5)[which(rownames(DF20.top5) %in% hallmark.list[["KEGG.OXPHOS"]])], 5),], avg_logFC)
)
rownames(DF20.gs.df) <- seq(from = 1, length = nrow(DF20.gs.df))

DF101.gs.df <- data.frame(
  "UNUPDATED.OXPHOS.DE" = head(rownames(DF101.top5)[which(rownames(DF101.top5) %in% hallmark.list[["UNUPDATED.OXPHOS"]])], 5), 
  "UNUPDATED.OXPHOS" = select(DF101.top5[head(rownames(DF101.top5)[which(rownames(DF101.top5) %in% hallmark.list[["UNUPDATED.OXPHOS"]])], 5),], avg_logFC),
  "GO.OXPHOS.DE" = head(rownames(DF101.top5)[which(rownames(DF101.top5) %in% hallmark.list[["GO.OXPHOS"]])], 5),
  "GO.OXPHOS" = select(DF101.top5[head(rownames(DF101.top5)[which(rownames(DF101.top5) %in% hallmark.list[["GO.OXPHOS"]])], 5),], avg_logFC),
  "KEGG.OXPHOS.DE" = head(rownames(DF101.top5)[which(rownames(DF101.top5) %in% hallmark.list[["KEGG.OXPHOS"]])], 5), 
  "KEGG.OXPHOS" = select(DF101.top5[head(rownames(DF101.top5)[which(rownames(DF101.top5) %in% hallmark.list[["KEGG.OXPHOS"]])], 5),], avg_logFC)
)
rownames(DF101.gs.df) <- seq(from = 1, length = nrow(DF20.gs.df))


DF68.gs.df <- data.frame(
  "UNUPDATED.OXPHOS.DE" = head(rownames(DF68.top5)[which(rownames(DF68.top5) %in% hallmark.list[["UNUPDATED.OXPHOS"]])], 5), 
  "UNUPDATED.OXPHOS" = select(DF68.top5[head(rownames(DF68.top5)[which(rownames(DF68.top5) %in% hallmark.list[["UNUPDATED.OXPHOS"]])], 5),], avg_logFC),
  "GO.OXPHOS.DE" = head(rownames(DF68.top5)[which(rownames(DF68.top5) %in% hallmark.list[["GO.OXPHOS"]])], 5),
  "GO.OXPHOS" = select(DF68.top5[head(rownames(DF68.top5)[which(rownames(DF68.top5) %in% hallmark.list[["GO.OXPHOS"]])], 5),], avg_logFC),
  "KEGG.OXPHOS.DE" = head(rownames(DF68.top5)[which(rownames(DF68.top5) %in% hallmark.list[["KEGG.OXPHOS"]])], 5), 
  "KEGG.OXPHOS" = select(DF68.top5[head(rownames(DF68.top5)[which(rownames(DF68.top5) %in% hallmark.list[["KEGG.OXPHOS"]])], 5),], avg_logFC)
)
rownames(DF68.gs.df) <- seq(from = 1, length = nrow(DF20.gs.df))

DF20.gs.df
DF101.gs.df
DF68.gs.df
```

* **OBSERVATIONS** 
    - Note: All of these top 5 DE OXPHOS genes (logFC > 0.5) have a padj of 1.00
    - There are not a lot of overlaps of DE OXPHOS genes across all three genesets for all three models 
    - Seems like a trend where the top DE OXPHOS genes present within the **UNUPDATED OXPHOS** geneset have higher logFC values in comparison to the other two genesets.
    
* Our answers to **QUESTION #1 and #2** seem to suggest that **UNUPDATED.OXPHOS** hallmark geneset is the best our of all the genesets tested because: 
    1. It has the least number of genes that are not found within the PDX Seurat Object, and the highest percentage of genes that are found. 
    2. It has the most DE OXPHOS genes (logFC > 0.5, regardless of padj) relative to the other two genesets for each model comparison between MRD and vehicle
    3. The DE OXPHOS genes present within this geneset have relatively higher logFC values (they are more DE, regardless of padj) in comparison to the DE OXPHOS genes present in the other two genesets. 
* We answer our 3rd question to confirm whether the **UNUPDATED.OXPHOS** geneset is indeed the most appropriate one for our data relative to the other two genesets. 

**ANSWERING QUESTION #3: which geneset gives us the most statistically significant results**

```{r oxphosGS, message=FALSE, warning=FALSE, fig.align='center', fig.width=15, fig.height=8}
oxphos.centered <- c("UNUPDATED.OXPHOS37.centered", "GO.OXPHOS35.centered", "KEGG.OXPHOS36.centered")
Oxphos.Vln.plots <- vector("list", length(PDXs))
names(Oxphos.Vln.plots) <- PDX.names

for (i in 1:length(PDXs)){
  obj <- PDXs[[i]]
  name <- PDX.names[[i]]
  numCells <- nrow(PDXs[[i]]@meta.data)
  
  my_comparisons <- list(
    c("MRD", "vehicle"),
    c("MRD", "relapse"), 
    c("vehicle", "relapse")
  )
  
if(name == "DF68"){
  p <- VlnPlot(obj, features = oxphos.centered, group.by = "treatment.status", pt.size = 0, cols = c("#00AFBB", "#E7B800", "#FC4E07"), combine = F, y.max = 1.5)
}
if(name == "DF101"){
   p <- VlnPlot(obj, features = oxphos.centered, group.by = "treatment.status", pt.size = 0, cols = c("#00AFBB", "#E7B800", "#FC4E07"), combine = F, y.max = 1.8)
}
else{
  p <- VlnPlot(obj, features = oxphos.centered, group.by = "treatment.status", pt.size = 0, cols = c("#00AFBB", "#E7B800", "#FC4E07"), combine = F, y.max = 2.0)
}
  p[[1]] <- p[[1]] + labs(title = glue("{name} UNUPDATED_OXPHOS scores across treatment"), x = name, subtitle = glue("Number of cells in {name}: {numCells}"), caption = paste("# OXPHOS genes in geneset found in PDX:", sum(hallmark.list[["UNUPDATED.OXPHOS"]] %in% rownames(obj)), "out of", length(hallmark.list[["UNUPDATED.OXPHOS"]]))) + 
    theme(plot.title = element_text(size = 12), plot.caption = element_text(size = 10)) + 
    geom_boxplot(width = 0.15, position = position_dodge(0.9), alpha = 0.3, show.legend = F) + 
    geom_text(label = paste(sum(obj$treatment.status == "vehicle"), "cells"), x = "vehicle", y = min(obj$UNUPDATED.OXPHOS37.centered) -0.03) + 
    geom_text(label = paste(sum(obj$treatment.status == "MRD"), "cells"), x = "MRD", y = min(obj$UNUPDATED.OXPHOS37.centered) - 0.03) + 
    geom_text(label = paste(sum(obj$treatment.status == "relapse"), "cells"), x = "relapse", y = min(obj$UNUPDATED.OXPHOS37.centered) - 0.03) + 
    stat_compare_means(comparisons = my_comparisons, method = "wilcox.test", label = "p.format", step.increase = 0.06) + 
    stat_compare_means(comparisons = my_comparisons, method = "wilcox.test", label = "p.signif", step.increase = 0.06, bracket.size = 0, vjust = 1.8)
  
  p[[2]] <- p[[2]] + labs(title = glue("{name} GO_OXPHOS scores across treatment"), x = name, subtitle = glue("Number of cells in {name}: {numCells}"), caption = paste("# OXPHOS genes in geneset found in PDX:", sum(hallmark.list[["GO.OXPHOS"]] %in% rownames(obj)),"out of", length(hallmark.list[["GO.OXPHOS"]]))) +
    theme(plot.title = element_text(size = 12), plot.caption = element_text(size = 10)) + 
    geom_boxplot(width = 0.15, position = position_dodge(0.9), alpha = 0.3, show.legend = F) + 
    stat_compare_means(comparisons = my_comparisons, method = "wilcox.test", label = "p.format", step.increase = 0.06) + 
    stat_compare_means(comparisons = my_comparisons, method = "wilcox.test", label = "p.signif", step.increase = 0.06, bracket.size = 0, vjust = 1.8)
  
  p[[3]] <- p[[3]] + labs(title = glue("{name} KEGG_OXPHOS scores across treatment"), x = name, subtitle = glue("Number of cells in {name}: {numCells}"), caption = paste("# OXPHOS genes in geneset found in PDX:", sum(hallmark.list[["KEGG.OXPHOS"]] %in% rownames(obj)), "out of", length(hallmark.list[["KEGG.OXPHOS"]]))) +
    theme(plot.title = element_text(size = 12), plot.caption = element_text(size = 10)) + 
    geom_boxplot(width = 0.15, position = position_dodge(0.9), alpha = 0.3, show.legend = F) + 
    stat_compare_means(comparisons = my_comparisons, method = "wilcox.test", label = "p.format", step.increase = 0.06) + 
    stat_compare_means(comparisons = my_comparisons, method = "wilcox.test", label = "p.signif", step.increase = 0.06, bracket.size = 0, vjust = 1.8)
  
  p <- p[[1]] + p[[2]] + p[[3]] + plot_layout(guides= 'collect')
  
  Oxphos.Vln.plots[[name]] <- p
}

Oxphos.Vln.plots[["DF20"]]
Oxphos.Vln.plots[["DF101"]]
Oxphos.Vln.plots[["DF68"]]
```

* **OBSERVATIONS**
    - The overall trend we see between treatment groups within each model is the same across all three oxphos genesets. However, it seems like the difference in scores is most obvious/drastic when using the **KEGG** geneset. 
    - More comparisons are statistically significant when using the **GO** or **KEGG** genesets in comparison to using the **HALLMARK** geneset. 
        - All comparisons within DF101 are significant 
            * **Vehicle** > **MRD** > **Relapse** (does not support our hypothesis)
        - DF68 **MRD** > **Relapse** is significant (supports our hypothesis)

We also tested how the **UNUPDATED** version of **HALLMARK_UNFOLDED_PROTEIN_RESPONSE** differs from the updated version.
```{r uprGS, message = FALSE, warning = FALSE}
upr.length.df <- data.frame(
  "UNUPDATED.HM.UPR" = length(hallmark.list[["UNUPDATED.UPR"]]),
  "HALLMARK.UPR" = length(hallmark.list[["HALLMARK_UNFOLDED_PROTEIN_RESPONSE"]])
)

Found.upr.df <- data.frame(
  "UNUPDATED.HM.UPR" = sum((hallmark.list[["UNUPDATED.UPR"]] %in% rownames(PDX_All))),
  "HALLMARK.UPR" = sum((hallmark.list[["HALLMARK_UNFOLDED_PROTEIN_RESPONSE"]] %in% rownames(PDX_All)))
)

PFound.upr.df <- data.frame(
  "UNUPDATED.HM.UPR" = (sum((hallmark.list[["UNUPDATED.UPR"]] %in% rownames(PDX_All)))/length(hallmark.list[["UNUPDATED.UPR"]]))*100,
  "HALLMARK.UPR" = (sum((hallmark.list[["HALLMARK_UNFOLDED_PROTEIN_RESPONSE"]] %in% rownames(PDX_All)))/length(hallmark.list[["HALLMARK_UNFOLDED_PROTEIN_RESPONSE"]])) * 100
)

upr.df <- rbind(upr.length.df, Found.upr.df, PFound.upr.df)
upr.df[,"UNUPDATED.HM.UPR"] <- round(upr.df[,"UNUPDATED.HM.UPR"])
upr.df[,"HALLMARK.UPR"] <- round(upr.df[,"HALLMARK.UPR"])
rownames(upr.df) <- c("Num UPR Genes", "Num Found", "% Found")
upr.df
```

* Since the **UNUPDATED** version identifies two more genes than the updated version, we decide to proceed with the **UNUPDATED.UPR** geneset. 

After deciding the geneset to use for our data, we can now analyze the differential expression of OXPHOS and UPR genes across treatment condition within each model. 

### STEP 3 DE ANALYSIS #1. VISUALIZING AND QUANTIFYING DE HALLMARK GENESETS
* **QUESTION** Are modules (**OXPHOS and UPR**) differentially expressed across treatment conditions within each model? 
* **HYPOTHESIS** We hypothesize that the **MRD** treatment condition will express enriched levels of OXPHOS and UPR hallmarks relative to cells in the **vehicle** and **relapse** treatment conditions (MRD > vehicle, MRD > relapse). 
    - **APPROACH #1** Violin Plots and UMAP
        * Center **OXPHOS and UPR** module scores at 0 and reassign to metadata
        * Visualize differences in hallmark score across treatment conditions with Violin Plot (center module scores at 0)
        * Statistical test (**wilcoxon rank sum test**): is the difference in hallmark score across treatment conditions statistically significant? 

```{r DEAnalysis1 VlnPlot, message = FALSE, warning = FALSE, fig.align = 'center', fig.width = 12, fig.height = 8}
hms.centered <- c("UNUPDATED.OXPHOS37.centered", "UNUPDATED.UPR38.centered")
#VlnPlot
PDX.names <- c("DF20", "DF101", "DF68")
PDXs <- c(PDX_DF20, PDX_DF101, PDX_DF68)
Vln.plots <- vector("list", length(PDXs))
names(Vln.plots) <- PDX.names

for (i in 1:length(PDXs)){
  obj <- PDX.names[[i]]
  numCells <- nrow(PDXs[[i]]@meta.data)
  
  my_comparisons <- list(
    c("MRD", "vehicle"),
    c("MRD", "relapse"), 
    c("vehicle", "relapse")
  )
  
  if(obj == "DF101"){
    p <- VlnPlot(PDXs[[i]], features = hms.centered, group.by = "treatment.status", pt.size = 0, c("#00AFBB", "#E7B800", "#FC4E07"), combine = F, y.max = 1.3)
  }
  else if(obj == "DF68"){
    p <- VlnPlot(PDXs[[i]], features = hms.centered, group.by = "treatment.status", pt.size = 0, c("#00AFBB", "#E7B800", "#FC4E07"), combine = F, y.max = 1.4)
  }
  else{
  p <- VlnPlot(PDXs[[i]], features = hms.centered, group.by = "treatment.status", pt.size = 0, c("#00AFBB", "#E7B800", "#FC4E07"), combine = F, y.max = 1.7)
  }
  
  p[[1]] <- p[[1]] + labs(title = glue("OXPHOS scores across treatment in {obj}"), x = obj, subtitle = glue("Number of cells in {obj}: {numCells}"), caption = "www.gsea-msigdb.org: HALLMARK_OXIDATIVE_PHOSPHORYLATION") +
    theme(plot.caption = element_text(size = 10)) + 
    geom_boxplot(width = 0.15, position = position_dodge(0.9), alpha = 0.3, show.legend = F) + 
    geom_text(label = paste(sum(PDXs[[i]]$treatment.status == "vehicle"), "cells"), x = "vehicle", y = min(PDXs[[i]]$UNUPDATED.OXPHOS37.centered) -0.03) + 
    geom_text(label = paste(sum(PDXs[[i]]$treatment.status == "MRD"), "cells"), x = "MRD", y = min(PDXs[[i]]$UNUPDATED.OXPHOS37.centered) - 0.03) + 
    geom_text(label = paste(sum(PDXs[[i]]$treatment.status == "relapse"), "cells"), x = "relapse", y = min(PDXs[[i]]$UNUPDATED.OXPHOS37.centered) - 0.03) + 
    stat_compare_means(comparisons = my_comparisons, method = "wilcox.test", label = "p.format", step.increase = 0.05) + 
    stat_compare_means(comparisons = my_comparisons, method = "wilcox.test", label = "p.signif", step.increase = 0.05, bracket.size = 0, vjust = 1.8)
  
  p[[2]] <- p[[2]] + labs(title = glue("UPR scores across treatment in {obj}"), x = obj, caption = "www.gsea-msigdb.org: HALLMARK_UNFOLDED_PROTEIN_RESPONSE") +
    theme(plot.caption = element_text(size = 10)) + 
    geom_boxplot(width = 0.15, position = position_dodge(0.9), alpha = 0.3, show.legend = F) + 
    stat_compare_means(comparisons = my_comparisons, method = "wilcox.test", label = "p.format", step.increase = 0.1) + 
    stat_compare_means(comparisons = my_comparisons, method = "wilcox.test", label = "p.signif", step.increase = 0.1, bracket.size = 0, vjust = 1.8)
  
  p <- p[[1]] + p[[2]] + plot_layout(guides= 'collect')
  
  Vln.plots[[obj]] <- p
}

Vln.plots[["DF20"]]
Vln.plots[["DF101"]]
Vln.plots[["DF68"]]
```

* **CONCLUSIONS**
    - The only statistically significant difference (p < 0.05) in hallmark scores is between treatment conditions within model **DF101** for **OXPHOS** 
    - DF101 OXPHOS: **Vehicle > MRD > Relapse**
    - However, the trends found in DF101 do not agree with our hypothesis that cells in **MRD** differentially enrich OXPHOS genes)

Considering how our approach with analyzing and comparing **module score** did not support our hypothesis, we try to confirm our results again with our second appraoch: **GSEA**

* **APPROACH #2** Geneset Enrichment Analysis (GSEA)
      - GSEA enrichment plots for hallmarks of interest between condition.1 vs. condition.2 
      - Rank genes and compute GSEA enrichment scores 
      - Statistical test: how significant are the enrichment scores? 

**have not finalized the statistical test and ranking method to use for GSEA**

```{r GSEA, message=FALSE, warning=FALSE, fig.align='center', fig.width=20, fig.height=10}

```

### STEP 4 DE ANALYSIS #2. IDENTIFYING INDIVIDUAL DE GENES

**have not finalized the statistical test to use for volcano plot (need to match what we use for GSEA)**

